---
title: "01_integrate_BCRF"
format: gfm
---

# 01_integrate_BCRF

Combine v1 and v2 using seurat v5 rpca integration, did not do any other testing- Magda
combine v1 and v2, need batch correction, check w pca/ clustering - Siri
S3S v2 has more reads

KEEP/ From TBCRCn, RAHBTn, COMETn

### Set-up 

```{r}
# Load libraries
library(future)
library(Seurat)
library(Matrix)
library(ggplot2)
```

```{r}
# Set global options 
options(future.globals.maxSize = Inf)
print(getOption("future.globals.maxSize"))
```

### Elaine's work on integrating BCRF dataset

This is how Elaine integrated the final BCRF dataset as far as I can tell 

```{r}
#| eval: FALSE 
options(future.globals.maxSize = Inf)

print(getOption("future.globals.maxSize"))
options(Seurat.object.assay.version = "v5")

# Load Seurat objects
BCRF_set1_3 = readRDS('/Users/eguo/Downloads/BCRF RNA set 1-3 & EN.rds') 
BCRF_set4_10 = readRDS('/Users/eguo/Downloads/BCRF RNA set 4-10.rds') 

BCRF1_3n <- BCRF_set1_3
BCRF4_10n <- BCRF_set4_10

# Merge Seurat objects
obj <- merge(x = BCRF1_3n,
             y = list(BCRF4_10n), 
             add.cell.ids = c('BCRF1_3&EN','BCRF4_10'))
```


```{r}
#| eval: FALSE
# Assign dataset labels
# obj$dataset = sapply(Cells(obj), function(x) strsplit(x, '_')[[1]][1])

# Normalize, scale from BCRF set xxx.R 
obj <- NormalizeData(obj) # Adjust for seq depth, lib size to compare bw samples, includes log transform
obj <- FindVariableFeatures(obj)
# obj$log2counts = log2(obj$nCount_RNA + 1)
obj <- ScaleData(obj) # Scales normalized, log-transformed counts for PCA or clustering
obj <- RunPCA(obj, features = VariableFeatures(object = obj))
obj <- FindNeighbors(obj, dims = 1:30, reduction = "pca")


# Integrate layers using RPCA
obj <- IntegrateLayers(
  object = obj, method = RPCAIntegration, k.weight = 90,
  orig.reduction = "pca", new.reduction = "integrated.rpca",
  verbose = FALSE
)

head(obj@meta.data)

# Find clusters after integration
obj <- FindClusters(obj, resolution = 2.5, cluster.name = "rpca_clusters")
obj <- RunUMAP(obj, reduction = "integrated.rpca", dims = 1:30, reduction.name = "umap.rpca")

# obj <- JoinLayers(obj) # Join layers if needed
# KEEP 

saveRDS(obj, file = "/Users/eguo/Downloads/integrated_BCRF_set_1-10_&_EN.rds")
```

Elaine's plot work.

```{r}
#| eval: FALSE

setwd("/Users/eguo/Downloads/integrated BCRF set 1-10 & EN")

DimPlot(obj, reduction = "umap.rpca", label = TRUE)
ggsave("umap.rpca.png", width = 8, height = 8)

obj$tissue[is.na(obj$tissue)] <- 'NA'
DimPlot(obj, reduction = "umap.rpca", label = TRUE, group.by = "tissue")
ggsave("tissue.png", width = 8, height = 8)

DimPlot(obj, reduction = "umap.rpca", group.by = "protocol")
ggsave("protocol.png", width = 8, height = 8)

DimPlot(obj, reduction = "umap.rpca", group.by = "dxs")
ggsave("dxs.png", width = 8, height = 8)  

DimPlot(obj, reduction = "umap.rpca", group.by = "dx")
ggsave("dx.png", width = 12, height = 8)

n_patients <- length(unique(obj$patient))
DimPlot(obj, reduction = "umap.rpca", group.by = "patient") +
  ggtitle(paste("n patients = ", n_patients))
ggsave("patient.png", width = 12, height = 8)  

DimPlot(obj, reduction = "umap.rpca", group.by = "sequence batch number")
ggsave("sequence batch number.png", width = 12, height = 8)

# log2 of total counts (library size) per sample
obj$log2counts <- log2(obj$nCount_RNA + 1)
log2counts <- FeaturePlot(obj,
            features = "log2counts",
            combine = TRUE, 
            label.size = 1) 
print(log2counts)
ggsave("log2counts.png", width = 8, height = 8)  


FeaturePlot(obj, features = c("ESR1", "ERBB2", "MYC", "TP53", "PIK3CA"), cols = c("red", "blue"))
ggsave("genes.png", width = 8, height = 8)
```

### UMAPs

```{r}
DimPlot(joined_BCRF, reduction = "umap.rpca", label = TRUE)

joined_BCRF$tissue[is.na(joined_BCRF$tissue)] <- 'NA'
DimPlot(joined_BCRF, reduction = "umap.rpca", label = TRUE, group.by = "tissue")

DimPlot(joined_BCRF, reduction = "umap.rpca", group.by = "protocol")

DimPlot(joined_BCRF, reduction = "umap.rpca", group.by = "dxs")

DimPlot(joined_BCRF, reduction = "umap.rpca", group.by = "dx")

n_patients <- length(unique(joined_BCRF$patient))
DimPlot(joined_BCRF, reduction = "umap.rpca", group.by = "patient") +
  ggtitle(paste("n patients = ", n_patients))

DimPlot(joined_BCRF, reduction = "umap.rpca", group.by = "sequence batch number")
```

```{r}
# log2 of total counts (library size) per sample
joined_BCRF$log2counts <- log2(joined_BCRF$nCount_RNA + 1)
log2counts <- FeaturePlot(joined_BCRF,
            features = "log2counts",
            combine = TRUE, 
            label.size = 1) 
print(log2counts)

FeaturePlot(joined_BCRF, features = c("ESR1", "ERBB2", "MYC", "TP53", "PIK3CA"), cols = c("red", "blue"))
```

### Finding cluster markers


```{r}
BCRF <- readRDS("../data/integrated_BCRF_set_1-10_&_EN.rds")
```

```{r}
cluster_markers_all <- FindAllMarkers(joined_BCRF, 
                       only.pos = FALSE,  # Includes both pos and neg markers
                       min.pct = 0.1,     # Genes must be expressed in min 10% of samples in a cluster
                       logfc.threshold = 0.25,  # LFC threshold
                       test.use = "DESeq2")

head(cluster_markers_all)
```

```{r}
# Filter by adj pval and Log2FC
cluster_markers_filtered <- cluster_markers_all %>%
  filter(p_val_adj < 0.05)

top_genes_per_cluster <- cluster_markers_filtered %>%
  filter(avg_log2FC > 1 | avg_log2FC < -1) %>%   # Filter for log2FC > 1 or < -1
  group_by(cluster) %>%                          # Group by cluster
  top_n(100, wt = abs(avg_log2FC)) %>%           # Select top DE genes by avg_log2FC for each cluster
  ungroup()

```
