---
title: "02_BCRF_DESeq2"
format: gfm
---

# 02_BCRF_DESeq2

The goal of this script will be to run DESeq2 for several comparisons in the BCRF dataset including:

- Epithelial v. Stromal Samples (Ep v. Str)

- Epithelial normal v. Epithelial Invasive (Ep NL v. Ep Invasive)

- Stromal Normal v. Stromal Invasive (Str NL v. Str Invasive)

## Set-up to run DESeq2

```{r}
# Load libraries
library(Seurat)
library(DESeq2)
library(BiocParallel)
register(MulticoreParam(7))
library(tidyverse)
library(EnhancedVolcano)

options(Seurat.object.assay.version = "v5")
```

```{r}
# Import data
BCRF <- readRDS("data/integrated_BCRF_set_1-10_&_EN.rds")
```

```{r}
# Join counts layers from BCRF RNA set 1-3 and BCRF RNA set 4-10
joined_BCRF <- JoinLayers(BCRF)
```

## Ep v. Str (pseudobulk)

```{r}
# Subset BCRF data to omit "NA", "Mix", and "EMT" tissue types 
Ep_Str_subsetted_BCRF <- subset(
  joined_BCRF,
  subset = tissue %in% c("Ep", "Str")
  )
```

```{r}
# Extract counts matrix from BCRF data 
# Pseudobulk to sum gene counts for all cells from same sample (patient) for each cell type (tissue)
Ep_Str_pb_counts <- AggregateExpression(
  subsetted_BCRF, 
  assays = "RNA", 
  group.by = c("patient", "tissue") # From meta.data 
)

Ep_Str_pb_counts_mat <- Ep_Str_pb_counts$RNA
```

```{r}
# Build colData 
# Select samples data from pseudobulk, will be identified by patient, tissue type i.e. g683_Str
Ep_Str_selected_samples <- colnames(Ep_Str_pb_counts_mat)

# Parse patient + tissue into separate columns
Ep_Str_pb_counts_parsed <- do.call(rbind, strsplit(Ep_Str_selected_samples, "_"))
colnames(Ep_Str_pb_counts_parsed) <- c("patient", "tissue")

# Remove g prefix from patient 
Ep_Str_pb_counts_parsed[, "patient"] <- sub("^g", "", Ep_Str_pb_counts_parsed[, "patient"])
```

```{r}
# Build colData
Ep_Str_pb_coldata <- data.frame(
  patient = factor(Ep_Str_pb_counts_parsed[, "patient"]), # Treat patient as categorical variable
  tissue = factor(Ep_Str_pb_counts_parsed[, "tissue"], levels = c("Ep", "Str")),
  row.names = Ep_Str_selected_samples
)
```

```{r}
# Create DESeqDataSet from subset counts and group information 
Ep_Str_dds <- DESeqDataSetFromMatrix(
  countData = Ep_Str_pb_counts_mat,
  colData = Ep_Str_pb_coldata, 
  design = ~ patient + tissue 
)
```

```{r}
# Filter to keep genes with at least 10 counts 
Ep_Str_dds_keep <- rowSums(counts(Ep_Str_dds)) >= 10
Ep_Str_dds <- Ep_Str_dds[Ep_Str_dds_keep,]
```

```{r}
# Run DESeq
Ep_Str_dds <- DESeq(Ep_Str_dds)
```

## Ep v. Str (working)

```{r}
# Subset BCRF data to omit "NA", "Mix", and "EMT" tissue types 
Ep_Str_subsetted_BCRF <- subset(
  joined_BCRF,
  subset = tissue %in% c("Ep", "Str")
  )

# Get counts data 
Ep_Str_counts <- GetAssayData(Ep_Str_subsetted_BCRF, assay = "RNA", layer = "counts")

# Build colData from meta.data
Ep_Str_coldata <- data.frame(
  patient = factor(Ep_Str_subsetted_BCRF@meta.data$patient), 
  tissue = factor(Ep_Str_subsetted_BCRF@meta.data$tissue, levels = c("Ep", "Str")),
  row.names = colnames(Ep_Str_counts)
)
```

```{r}
# Create DESeqDataSet from subset counts and group information 
Ep_Str_dds <- DESeqDataSetFromMatrix(
  countData = Ep_Str_counts,
  colData = Ep_Str_coldata, 
  design = ~ patient + tissue 
)
```

```{r}
# Keep genes with at least 1 count in at least 10 cells
Ep_Str_dds_keep <- rowSums(counts(Ep_Str_dds) >= 1) >= 10
Ep_Str_dds <- Ep_Str_dds[Ep_Str_dds_keep,]
```

```{r}
Ep_Str_dds <- DESeq(Ep_Str_dds, parallel = TRUE)
```

## Ep v. Str (non-pb) OLD BAD 

```{r}
# Define groups 
# Subset metadata 
Ep_Str_metadata <- joined_BCRF@meta.data

# Select samples with Ep or Str tissue labels 
# Omitted "NA", "Mix", "EMT" labels 
Ep_Str_selected_samples <- rownames(Ep_Str_metadata)[Ep_Str_metadata$tissue %in% c("Ep", "Str")]

# Subset counts matrix to include only the selected samples
Ep_Str_subset_counts <- counts[, Ep_Str_selected_samples]

# Create subset group 
Ep_Str_subset_group <- factor(
  Ep_Str_metadata[Ep_Str_selected_samples, "tissue"],
  levels = c("Ep", "Str")
)
```

```{r}
# Create DESeqDataSet from subset counts and group information 
Ep_Str_dds <- DESeqDataSetFromMatrix(
  countData = Ep_Str_subset_counts,
  colData = data.frame(
    group = Ep_Str_subset_group,
    row.names = colnames(Ep_Str_subset_counts)), # Make row names of colData explicit  
  design = ~group 
)
```

```{r}
# Run DESeq2 analysis 
Ep_Str_dds_object <- DESeq(Ep_Str_dds)
```

## Ep NL v. Ep Invasive 

```{r}
# Subset BCRF data to omit dxs that are not "NL" or "Invasive" 
Ep_subsetted_BCRF <- subset(
  joined_BCRF,
  subset = dxs %in% c("NL", "Invasive")
  )

# Subset BCRF data to select only "Ep" tissue samples 
Ep_subsetted_BCRF <- subset(
  Ep_subsetted_BCRF, 
  subset = tissue %in% "Ep"
)
```

```{r}
# Extract counts matrix from BCRF data 
# Pseudobulk to sum gene counts for all cells from same sample (patient) for each cell type (tissue)
Ep_pb_counts <- AggregateExpression(
  Ep_subsetted_BCRF, 
  assays = "RNA", 
  group.by = c("patient", "dxs") # From meta.data 
)

Ep_pb_counts_mat <- Ep_pb_counts$RNA
```

```{r}
# Build colData 
# Select samples data from pseudobulk, will be identified by patient, tissue type i.e. g683_Str
Ep_selected_samples <- colnames(Ep_pb_counts_mat)

# Parse patient + tissue into separate columns
Ep_pb_counts_parsed <- do.call(rbind, strsplit(Ep_selected_samples, "_"))
colnames(Ep_pb_counts_parsed) <- c("patient", "dxs")

# Remove g prefix from patient 
Ep_pb_counts_parsed[, "patient"] <- sub("^g", "", Ep_pb_counts_parsed[, "patient"])
```

```{r}
# Build colData
Ep_pb_coldata <- data.frame(
  patient = factor(Ep_pb_counts_parsed[, "patient"]), # Treat patient as categorical variable
  dxs = factor(Ep_pb_counts_parsed[, "dxs"], levels = c("NL", "Invasive")),
  row.names = Ep_selected_samples
)
```

```{r}
# Create DESeqDataSet from subset counts and group information 
Ep_dds <- DESeqDataSetFromMatrix(
  countData = Ep_pb_counts_mat,
  colData = Ep_pb_coldata, 
  design = ~ patient + dxs
)
```

```{r}
# Filter to keep genes with at least 10 counts 
Ep_dds_keep <- rowSums(counts(Ep_dds)) >= 10
Ep_dds <- Ep_dds[Ep_dds_keep,]
```

```{r}
# Run DESeq
Ep_dds <- DESeq(Ep_dds, parallel = TRUE)
```

```{r}
# Save DESeq dds 
saveRDS(Ep_dds, "data/DESeq2/EpNL_EpInv_dds.rds")
```

```{r}
# Save DESeq results 
Ep_dds_res <- results(Ep_dds, contrast = c("dxs", "NL", "Invasive"))
summary(Ep_dds_res)

# Convert results to dataframe
Ep_dds_res_df <- Ep_dds_res |>
  as.data.frame() |>
  rownames_to_column(var = "Gene") # Make column for gene ids 

# Save results 
write.csv(Ep_dds_res_df, "data/DESeq2/EpNL_EpInv_res.csv")
```

```{r}
# Filter for significant results (adjusted p-value < 0.05)
Ep_sig_res <- Ep_dds_res[which(Ep_dds_res$padj < 0.05), ]

# Convert to dataframe
Ep_sig_res_df <- as.data.frame(Ep_sig_res)
```

### Volcano Plot

```{r}
EpNL_EpInv_VP <- EnhancedVolcano(Ep_dds_res_df,
  x = "log2FoldChange", # fold change statistic to plot
  y = "pvalue", # significance values
  lab = Ep_dds_res_df$Gene, # labels for points
  pCutoff = 1e-05, # The p value cutoff we will use (default)
  FCcutoff = 1, # The fold change cutoff (default)
  title = "Genes differentially expressed in Ep Invasive relative to Ep NL", # add title
  caption = NULL,
  labSize = 1.5 # smaller labels
) +
  # change the overall theme
  theme_classic() +
  # move the legend to the bottom
  theme(legend.position = "bottom")

# Save plots as .png
ggsave("EpNL_EpInv_VP.png", EpNL_EpInv_VP, path = "plots/DESeq2")
```
