---
title: "02_BCRF_DESeq2"
format: gfm
---

# 02_BCRF_DESeq2

The goal of this script will be to run DESeq2 for several comparisons in the BCRF dataset including:

- Epithelial v. Stromal Samples (Ep v. Str)

- Epithelial normal v. Epithelial Invasive (Ep NL v. Ep Invasive)

- Stromal Normal v. Stromal Invasive (Str NL v. Str Invasive)

## Set-up to run DESeq2

```{r}
# Load libraries
library(Seurat)
library(DESeq2)
library(BiocParallel)
register(SnowParam(7))
library(tidyverse)
library(EnhancedVolcano)

options(Seurat.object.assay.version = "v5")
```

```{r}
# Import data
BCRF <- readRDS("data/integrated_BCRF_set_1-10_&_EN.rds")
```

```{r}
# Join counts layers from BCRF RNA set 1-3 and BCRF RNA set 4-10
joined_BCRF <- JoinLayers(BCRF)
```

## Ep v. Str

```{r}
# Subset BCRF data to omit "NA", "Mix", and "EMT" tissue types 
Ep_Str_subsetted_BCRF <- subset(
  joined_BCRF,
  subset = tissue %in% c("Ep", "Str")
  )

# Get counts data 
Ep_Str_counts <- GetAssayData(Ep_Str_subsetted_BCRF, assay = "RNA", layer = "counts")

# Build colData from meta.data
Ep_Str_coldata <- data.frame(
  patient = factor(Ep_Str_subsetted_BCRF@meta.data$patient), 
  tissue = factor(Ep_Str_subsetted_BCRF@meta.data$tissue, levels = c("Ep", "Str")),
  row.names = colnames(Ep_Str_counts)
)
```

```{r}
# Find all samples that have zero nonzero counts
Ep_Str_nonzero_per_cell <- colSums(Ep_Str_counts > 0)
summary(Ep_Str_nonzero_per_cell) # Min. = 0 

# Any cells with zero or near-zero nonzero genes?
sum(Ep_Str_nonzero_per_cell == 0)  # 1 
which(Ep_Str_nonzero_per_cell < 5) # BCRF1_3&EN_11886_364_NL_S_S2_R1_001, 107
```

```{r}
# Filter out samples that have zero counts for all genes
Ep_Str_coldata <- Ep_Str_coldata[Ep_Str_nonzero_per_cell > 0, ]
Ep_Str_counts <- Ep_Str_counts[, Ep_Str_nonzero_per_cell > 0]

# double check
dim(Ep_Str_counts)
dim(Ep_Str_coldata)
identical(colnames(Ep_Str_counts), rownames(Ep_Str_coldata)) # TRUE
```

```{r}
# Create DESeqDataSet from subset counts and group information 
Ep_Str_dds <- DESeqDataSetFromMatrix(
  countData = Ep_Str_counts,
  colData = Ep_Str_coldata, 
  design = ~ patient + tissue 
)
```

```{r}
# Keep genes with at least 1 count in at least 10 cells
Ep_Str_dds_keep <- rowSums(counts(Ep_Str_dds) >= 1) >= 10
Ep_Str_dds <- Ep_Str_dds[Ep_Str_dds_keep,]
```

```{r}
Ep_Str_dds <- DESeq(Ep_Str_dds, sfType = "poscounts", parallel = TRUE)
```

## Ep NL v. Ep Invasive 

```{r}
# Subset BCRF data to omit dxs that are not "NL" or "Invasive"
NL_Inv_subsetted_BCRF <- subset(
  joined_BCRF,
  subset = dxs %in% c("NL", "Invasive")
  )

# Subset BCRF data to only include "Ep" tissue samples
EpNL_EpInv_subsetted_BCRF <- subset(
  NL_Inv_subsetted_BCRF,
  subset = tissue %in% "Ep"
)

# Get counts data 
EpNL_EpInv_counts <- GetAssayData(EpNL_EpInv_subsetted_BCRF, assay = "RNA", layer = "counts")

# Build colData from meta.data
EpNL_EpInv_coldata <- data.frame(
  patient = factor(EpNL_EpInv_subsetted_BCRF@meta.data$patient), 
  dxs = factor(EpNL_EpInv_subsetted_BCRF@meta.data$dxs, levels = c("NL", "Invasive")),
  row.names = colnames(EpNL_EpInv_counts)
)
```

```{r}
# Create DESeqDataSet from subset counts and group information 
EpNL_EpInv_dds <- DESeqDataSetFromMatrix(
  countData = EpNL_EpInv_counts,
  colData = EpNL_EpInv_coldata, 
  design = ~ patient + dxs 
)
```

```{r}
# Keep genes with at least 1 count in at least 10 cells
EpNL_EpInv_dds_keep <- rowSums(counts(EpNL_EpInv_dds) >= 1) >= 10
EpNL_EpInv_dds <- EpNL_EpInv_dds[EpNL_EpInv_dds_keep,]
```

```{r}
saveRDS(EpNL_EpInv_dds, "data/DESeq2/EpNL_EpInv_dds_torun.rds")
EpNL_EpInv_dds <- readRDS("data/DESeq2/EpNL_EpInv_dds_torun.rds")
```

```{r}
# Run DEseq
EpNL_EpInv_dds <- DESeq(EpNL_EpInv_dds, parallel = TRUE)
```

```{r}
# Save DESeq results 
EpNL_EpInv_dds_res <- results(EpNL_EpInv_dds, contrast = c("dxs", "Invasive", "NL"))
summary(EpNL_EpInv_dds_res)
saveRDS(EpNL_EpInv_dds_res, "data/DESeq2/EpNL_EpInv_dds_res.rds")

# Convert results to dataframe
EpNL_EpInv_dds_df <- EpNL_EpInv_dds_res |>
  as.data.frame() |>
  rownames_to_column(var = "Gene") # Make column for gene ids 

# Save results 
write.csv(EpNL_EpInv_dds_df, "data/DESeq2/EpNL_EpInv_res.csv")
```

### Volcano Plot 

```{r}
EpNL_EpInv_res <- read.csv("data/DESeq2/EpNL_EpInv_res.csv")
```

```{r}
keyGenes <- c("ESR1", "ERBB2", "MYC", "TP53", "PIK3CA", "CCND1")

EpNL_EpInv_VP <- EnhancedVolcano(EpNL_EpInv_res,
  x = "log2FoldChange", # fold change statistic to plot
  y = "pvalue", # significance values
  lab = EpNL_EpInv_res$Gene, # labels for points
  pCutoff = 1e-05, # The p value cutoff we will use (default)
  FCcutoff = 1, # The fold change cutoff (default)
  title = "Genes differentially expressed in Ep Invasive relative to Ep NL",
  selectLab = keyGenes,         # Label only key genes
  labSize = 3.0, 
  labCol = "black",
  labFace = "bold",
  boxedLabels = TRUE,           # boxes make labels stand out from the cloud
  drawConnectors = TRUE,        # connector lines from point to label
  widthConnectors = 0.5,
  colConnectors = 'black',
  arrowheads = FALSE 
) +
  # change the overall theme
  theme_classic() +
  # move the legend to the bottom
  theme(legend.position = "bottom")

# Save plots as .png
ggsave("EpNL_EpInv_VP.png", EpNL_EpInv_VP, path = "plots/DESeq2")
```

## Str NL v. Str Invasive 

```{r}
# Subset BCRF data to omit dxs that are not "NL" or "Invasive"
NL_Inv_subsetted_BCRF <- subset(
  joined_BCRF,
  subset = dxs %in% c("NL", "Invasive")
  )

# Subset BCRF data to only include "Str" tissue samples
StrNL_StrInv_subsetted_BCRF <- subset(
  NL_Inv_subsetted_BCRF,
  subset = tissue %in% "Str"
)

# Get counts data 
StrNL_StrInv_counts <- GetAssayData(StrNL_StrInv_subsetted_BCRF, assay = "RNA", layer = "counts")

# Build colData from meta.data
StrNL_StrInv_coldata <- data.frame(
  patient = factor(StrNL_StrInv_subsetted_BCRF@meta.data$patient), 
  dxs = factor(StrNL_StrInv_subsetted_BCRF@meta.data$dxs, levels = c("NL", "Invasive")),
  row.names = colnames(StrNL_StrInv_counts)
)
```

```{r}
# Find all samples that have zero nonzero counts
# How many nonzero genes does each cell have, after filtering?
StrNL_StrInv_nonzero_per_cell <- colSums(StrNL_StrInv_counts > 0)
summary(StrNL_StrInv_nonzero_per_cell) # Min. = 1 

# Any cells with zero or near-zero nonzero genes?
sum(StrNL_StrInv_nonzero_per_cell == 0)  # 1 
which(StrNL_StrInv_nonzero_per_cell < 5) # BCRF1_3&EN_11886_364_NL_S_S2_R1_001, 9
```

```{r}
# Filter out samples that have zero counts for all genes
# Defined in troubleshooting section 
StrNL_StrInv_coldata <- StrNL_StrInv_coldata[StrNL_StrInv_nonzero_per_cell > 0, ]
StrNL_StrInv_counts <- StrNL_StrInv_counts[, StrNL_StrInv_nonzero_per_cell > 0]

# double check
dim(StrNL_StrInv_counts)
dim(StrNL_StrInv_coldata)
identical(colnames(StrNL_StrInv_counts), rownames(StrNL_StrInv_coldata)) # TRUE
```

```{r}
# Create DESeqDataSet from subset counts and group information 
StrNL_StrInv_dds <- DESeqDataSetFromMatrix(
  countData = StrNL_StrInv_counts,
  colData = StrNL_StrInv_coldata, 
  design = ~ patient + dxs 
)
```

```{r}
# Keep genes with at least 1 count in at least 10 cells
StrNL_StrInv_dds_keep <- rowSums(counts(StrNL_StrInv_dds) >= 1) >= 10
StrNL_StrInv_dds <- StrNL_StrInv_dds[StrNL_StrInv_dds_keep,]
```

```{r}
# Save DESeqDataSet object so environment can be cleared 
saveRDS(StrNL_StrInv_dds, "data/DESeq2/StrNL_StrInv_dds_torun.rds")
StrNL_StrInv_dds <- readRDS("data/DESeq2/StrNL_StrInv_dds_torun.rds")
```

```{r}
# Run DESeq
StrNL_StrInv_dds <- DESeq(StrNL_StrInv_dds, parallel = TRUE)
```

```{r}
# Save DESeq results 
StrNL_StrInv_dds_res <- results(StrNL_StrInv_dds, contrast = c("dxs", "Invasive", "NL"))
summary(StrNL_StrInv_dds_res)
saveRDS(StrNL_StrInv_dds_res, "data/DESeq2/StrNL_StrInv_res.rds")

# Convert results to dataframe
StrNL_StrInv_dds_df <- StrNL_StrInv_dds_res |>
  as.data.frame() |>
  rownames_to_column(var = "Gene") # Make column for gene ids 

# Save results 
write.csv(StrNL_StrInv_dds_df, "data/DESeq2/StrNL_StrInv_res.csv")
```

### Volcano Plot 
```{r}
# Load in results data 
StrNL_StrInv_res <- read.csv("data/DESeq2/StrNL_StrInv_res.csv")
```

```{r}
# Generate volcano plot 
keyGenes <- c("ESR1", "ERBB2", "MYC", "TP53", "PIK3CA", "CCND1")

# Any expression may be dissection artifacts? 

StrNL_StrInv_VP <- EnhancedVolcano(StrNL_StrInv_res,
  x = "log2FoldChange",                 
  y = "padj",                
  lab = StrNL_StrInv_res$Gene, 
  pCutoff = 1e-05, 
  FCcutoff = 1,
  title = "Genes differentially expressed in Str Invasive relative to Str NL",
  selectLab = keyGenes,         # Label only key genes
  labSize = 3.0, 
  labCol = "black",
  labFace = "bold",
  boxedLabels = TRUE,           # boxes make labels stand out from the cloud
  drawConnectors = TRUE,        # connector lines from point to label
  widthConnectors = 0.5,
  colConnectors = 'black',
  arrowheads = FALSE 
) +
  # change the overall theme
  theme_classic() +
  # move the legend to the bottom
  theme(legend.position = "bottom")

# Save plots as .png
ggsave("StrNL_StrInv_VP.png", StrNL_StrInv_VP, path = "plots/DESeq2",
       width = 7, height = 7, dpi = 300)
```

## Troubleshooting

Trying to run DESeq on Ep v. Str and Str NL v. StrI Inv returns: 
Error in estimateSizeFactorsForMatrix(counts(object), locfunc = locfunc,  : 
  every gene contains at least one zero, cannot compute log geometric means

Trying to run with argument: sfType = "poscounts" returns: 
Error in .local(object, ..., value) : all(!is.na(value)) is not TRUE

```{r}
# Check if there are any NAs in counts data 
StrNL_StrInv_counts_df <- as.data.frame(StrNL_StrInv_counts)
anyNA(StrNL_StrInv_counts) # FALSE 
```

```{r}
# Check for NAs in colData
sum(is.na(StrNL_StrInv_coldata$patient)) # 0 
sum(is.na(StrNL_StrInv_coldata$dxs))     # 0 
table(StrNL_StrInv_coldata$patient, StrNL_StrInv_coldata$dxs, useNA = "ifany")
# No NAs 
```

There were no immediate red flags, so I wanted to create a histogram of raw counts to see the distribution of counts per gene before deciding on filtering parameters.  

```{r}
StrNL_StrInv_counts_long <- StrNL_StrInv_counts_df |>
  rownames_to_column("gene") |>
  pivot_longer(-gene, names_to = "sample", values_to = "count")
```

```{r}
StrNL_StrInv_counts_histogram <- StrNL_StrInv_counts_long |>
  ggplot(aes(x = count, fill = sample)) +
  geom_histogram(bins = 50) +
  scale_x_log10() +
  theme(legend.position = "none") +
  labs(x = "Raw count (log10 scale)", y = "Number of gene-sample entries",
       title = "Distribution of raw counts")

StrNL_StrInv_counts_histogram
```

```{r}
StrNL_StrInv_counts_toplot <- StrNL_StrInv_counts_df |>
  mutate(
    total_counts = rowSums(across(everything())),
    .before = 1
  ) |>
  filter(total_counts == 0)
```

```{r}
StrNL_StrInv_dds_counts <- counts(StrNL_StrInv_dds)

# How many nonzero genes does each cell have, after filtering?
nonzero_per_cell <- colSums(StrNL_StrInv_dds_counts > 0)
summary(nonzero_per_cell) # Min. = 1 

# Any cells with zero or near-zero nonzero genes?
sum(nonzero_per_cell == 0)  # 1 
which(nonzero_per_cell < 5) # BCRF1_3&EN_11886_364_NL_S_S2_R1_001, 9
```

I have diagnosed the sample that has zero counts for all genes. 

# Pseudobulking lol TT.TT

## Ep v. Str

```{r}
# Subset BCRF data to omit "NA", "Mix", and "EMT" tissue types 
Ep_Str_subsetted_BCRF <- subset(
  joined_BCRF,
  subset = tissue %in% c("Ep", "Str")
  )
```

```{r}
# Extract counts matrix from BCRF data 
# Pseudobulk to sum gene counts for all cells from same sample (patient) for each cell type (tissue)
Ep_Str_pb_counts <- AggregateExpression(
  subsetted_BCRF, 
  assays = "RNA", 
  group.by = c("patient", "tissue") # From meta.data 
)

Ep_Str_pb_counts_mat <- Ep_Str_pb_counts$RNA
```

```{r}
# Build colData 
# Select samples data from pseudobulk, will be identified by patient, tissue type i.e. g683_Str
Ep_Str_selected_samples <- colnames(Ep_Str_pb_counts_mat)

# Parse patient + tissue into separate columns
Ep_Str_pb_counts_parsed <- do.call(rbind, strsplit(Ep_Str_selected_samples, "_"))
colnames(Ep_Str_pb_counts_parsed) <- c("patient", "tissue")

# Remove g prefix from patient 
Ep_Str_pb_counts_parsed[, "patient"] <- sub("^g", "", Ep_Str_pb_counts_parsed[, "patient"])
```

```{r}
# Build colData
Ep_Str_pb_coldata <- data.frame(
  patient = factor(Ep_Str_pb_counts_parsed[, "patient"]), # Treat patient as categorical variable
  tissue = factor(Ep_Str_pb_counts_parsed[, "tissue"], levels = c("Ep", "Str")),
  row.names = Ep_Str_selected_samples
)
```

```{r}
# Create DESeqDataSet from subset counts and group information 
Ep_Str_dds <- DESeqDataSetFromMatrix(
  countData = Ep_Str_pb_counts_mat,
  colData = Ep_Str_pb_coldata, 
  design = ~ patient + tissue 
)
```

```{r}
# Filter to keep genes with at least 10 counts 
Ep_Str_dds_keep <- rowSums(counts(Ep_Str_dds)) >= 10
Ep_Str_dds <- Ep_Str_dds[Ep_Str_dds_keep,]
```

```{r}
# Run DESeq
Ep_Str_dds <- DESeq(Ep_Str_dds)
```

## Ep NL v. Ep Invasive

```{r}
# Subset BCRF data to omit dxs that are not "NL" or "Invasive" 
NL_Inv_subsetted_BCRF <- subset(
  joined_BCRF,
  subset = dxs %in% c("NL", "Invasive")
  )

# Subset BCRF data to select only "Ep" tissue samples 
EpNL_EpInv_subsetted_BCRF <- subset(
  NL_Inv_subsetted_BCRF, 
  subset = tissue %in% "Ep"
)
```

```{r}
# Extract counts matrix from BCRF data 
# Pseudobulk to sum gene counts for all cells from same sample (patient) for each cell type (tissue)
EpNL_EpInv_pb_counts <- AggregateExpression(
  EpNL_EpInv_subsetted_BCRF, 
  assays = "RNA", 
  group.by = c("patient", "dxs") # From meta.data 
)

EpNL_EpInv_pb_counts_mat <- EpNL_EpInv_pb_counts$RNA
```

```{r}
# Build colData 
# Select samples data from pseudobulk, will be identified by patient, tissue type i.e. g683_Str
EpNL_EpInv_selected_samples <- colnames(EpNL_EpInv_pb_counts_mat)

# Parse patient + tissue into separate columns
EpNL_EpInv_pb_counts_parsed <- do.call(rbind, strsplit(EpNL_EpInv_selected_samples, "_"))
colnames(EpNL_EpInv_pb_counts_parsed) <- c("patient", "dxs")

# Remove g prefix from patient 
EpNL_EpInv_pb_counts_parsed[, "patient"] <- sub("^g", "", EpNL_EpInv_pb_counts_parsed[, "patient"])
```

```{r}
# Build colData
EpNL_EpInv_pb_coldata <- data.frame(
  patient = factor(EpNL_EpInv_pb_counts_parsed[, "patient"]), # Treat patient as categorical variable
  dxs = factor(EpNL_EpInv_pb_counts_parsed[, "dxs"], levels = c("NL", "Invasive")),
  row.names = EpNL_EpInv_selected_samples
)
```

```{r}
# Create DESeqDataSet from subset counts and group information 
EpNL_EpInv_pb_dds <- DESeqDataSetFromMatrix(
  countData = EpNL_EpInv_pb_counts_mat,
  colData = EpNL_EpInv_pb_coldata, 
  design = ~ patient + dxs
)
```

```{r}
# Filter to keep genes with at least 10 counts 
EpNL_EpInv_pb_dds_keep <- rowSums(counts(EpNL_EpInv_dds)) >= 10
EpNL_EpInv_dds <- EpNL_EpInv_dds[EpNL_EpInv_pb_dds_keep,]
```

```{r}
# Run DESeq
EpNL_EpInv_pb_dds <- DESeq(EpNL_EpInv_pb_dds, parallel = TRUE)
```

```{r}
# Save DESeq results 
EpNL_EpInv_pb_dds_res <- results(EpNL_EpInv_pb_dds, contrast = c("dxs", "Invasive", "NL"))
summary(EpNL_EpInv_pb_dds_res)
saveRDS(EpNL_EpInv_pb_dds_res, "data/DESeq2/EpNL_EpInv_pb_dds_res.rds")

# Convert results to dataframe
EpNL_EpInv_pb_dds_res_df <- EpNL_EpInv_pb_dds_res |>
  as.data.frame() |>
  rownames_to_column(var = "Gene") # Make column for gene ids 

# Save results 
write.csv(EpNL_EpInv_pb_dds_res_df, "data/DESeq2/EpNL_EpInv_pb_res.csv")
```

```{r}
# Filter for significant results (adjusted p-value < 0.05)
Ep_sig_res <- Ep_dds_res[which(Ep_dds_res$padj < 0.05), ]

# Convert to dataframe
Ep_sig_res_df <- as.data.frame(Ep_sig_res)
```

### Volcano Plot

```{r}
EpNL_EpInv_pb_VP <- EnhancedVolcano(EpNL_EpInv_pb_res,
  x = "log2FoldChange", # fold change statistic to plot
  y = "pvalue", # significance values
  lab = EpNL_EpInv_pb_res$Gene, # labels for points
  pCutoff = 1e-05, # The p value cutoff we will use (default)
  FCcutoff = 1, # The fold change cutoff (default)
  title = "Genes differentially expressed in Ep Invasive relative to Ep NL", # add title
  caption = NULL,
  labSize = 1.5 # smaller labels
) +
  # change the overall theme
  theme_classic() +
  # move the legend to the bottom
  theme(legend.position = "bottom")

# Save plots as .png
ggsave("EpNL_EpInv_pb_VP.png", EpNL_EpInv_pb_VP, path = "plots/DESeq2")
```

## Str NL v. Str Invasive 

```{r}
# Subset BCRF data to omit dxs that are not "NL" or "Invasive" 
NL_Inv_subsetted_BCRF <- subset(
  joined_BCRF,
  subset = dxs %in% c("NL", "Invasive")
  )

# Subset BCRF data to select only "Str" tissue samples 
StrNL_StrInv_subsetted_BCRF <- subset(
  NL_Inv_subsetted_BCRF, 
  subset = tissue %in% "Str"
)
```

```{r}
# Extract counts matrix from BCRF data 
# Pseudobulk to sum gene counts for all cells from same sample (patient) for each condiition (dxs)
StrNL_StrInv_pb_counts <- AggregateExpression(
  StrNL_StrInv_subsetted_BCRF, 
  assays = "RNA", 
  group.by = c("patient", "dxs") # From meta.data 
)

StrNL_StrInv_pb_counts <- StrNL_StrInv_pb_counts$RNA
```

```{r}
# Build colData 
# Select samples data from pseudobulk, will be identified by patient, tissue type i.e. g683_Str
StrNL_StrInv_selected_samples <- colnames(StrNL_StrInv_pb_counts)

# Parse patient + tissue into separate columns
StrNL_StrInv_pb_counts_parsed <- do.call(rbind, strsplit(StrNL_StrInv_selected_samples, "_"))
colnames(StrNL_StrInv_pb_counts_parsed) <- c("patient", "dxs")

# Remove g prefix from patient 
StrNL_StrInv_pb_counts_parsed[, "patient"] <- sub("^g", "", StrNL_StrInv_pb_counts_parsed[, "patient"])
```

```{r}
# Build colData
StrNL_StrInv_pb_coldata <- data.frame(
  patient = factor(StrNL_StrInv_pb_counts_parsed[, "patient"]), # Treat patient as categorical variable
  dxs = factor(StrNL_StrInv_pb_counts_parsed[, "dxs"], levels = c("NL", "Invasive")),
  row.names = StrNL_StrInv_selected_samples
)
```

```{r}
# Create DESeqDataSet from subset counts and group information 
StrNL_StrInv_pb_dds <- DESeqDataSetFromMatrix(
  countData = StrNL_StrInv_pb_counts,
  colData = StrNL_StrInv_pb_coldata, 
  design = ~ patient + dxs
)
```

```{r}
# Keep genes with at least 1 count in at least 10 cells
StrNL_StrInv_pb_dds_keep <- rowSums(counts(StrNL_StrInv_pb_dds) >= 1) >= 10
StrNL_StrInv_pb_dds <- StrNL_StrInv_pb_dds[StrNL_StrInv_pb_dds_keep,]
```

```{r}
# Run DESeq
StrNL_StrInv_pb_dds <- DESeq(StrNL_StrInv_pb_dds, sfType = "poscounts", parallel = TRUE)
```

```{r}
# Save DESeq dds 
saveRDS(StrNL_StrInv_pb_dds, "data/DESeq2/StrNL_StrInv_pb_dds.rds")
```

```{r}
# Save DESeq results 
StrNL_StrInv_pb_dds_res <- results(StrNL_StrInv_pb_dds, contrast = c("dxs", "Invasive", "NL"))
summary(StrNL_StrInv_pb_dds_res)

# Convert results to dataframe
StrNL_StrInv_pb_dds_res_df <- StrNL_StrInv_pb_dds_res |>
  as.data.frame() |>
  rownames_to_column(var = "Gene") # Make column for gene ids 

# Save results 
write.csv(StrNL_StrInv_pb_dds_res_df, "data/DESeq2/StrNL_StrInv_pb_res.csv")
```

### Troubleshooting

