---
title: "03_BCRF_GSEA"
format: gfm
---

# 03_BCRF_GSEA

The goal of this script is to perform gene set enrichment analysis on DESeq2 results from several comparisons in the BCRF dataset, including:

- Epithelial v. Stromal Samples (Ep v. Str)

- Epithelial normal v. Epithelial Invasive (Ep NL v. Ep Invasive)

- Stromal Normal v. Stromal Invasive (Str NL v. Str Invasive)


## Set-up to run GSEA

```{r}
# Load libraries
library(readr)
library(org.Hs.eg.db)
library(enrichplot)
library(ggplot2)
library(tidyverse)
library(clusterProfiler)
library(UpSetR)
library(ggplotify)
```

## Ep NL v. Ep Invasive 

### Running GSEA
```{r}
# Prepare data 
# Load in data 
EpNL_EpInv <- read.csv("data/DESeq2/EpNL_EpInv_res.csv", row.names = 1)
print(head(EpNL_EpInv))
```

```{r}
# Prepare ranked list of genes
# Use signed p value as ranking: sign(df$log2fc)*(-log10(df$pval))) to account for genes w/ large FC but are non-significant
EpNL_EpInv_rankings <- sign(EpNL_EpInv$log2FoldChange)*(-log10(EpNL_EpInv$pvalue))
names(EpNL_EpInv_rankings) <- EpNL_EpInv$Gene # genes as names

# Sort by decreasing signed p values 
EpNL_EpInv_rankings <- sort(EpNL_EpInv_rankings, decreasing = TRUE)

# Convert from SYMBOL to Entrez nomenclature 
EpNL_EpInv_genes_mapped <- bitr(
  names(EpNL_EpInv_rankings),
  fromType = "SYMBOL", 
  toType = c("ENTREZID"),
  OrgDb = org.Hs.eg.db)

nrow(EpNL_EpInv_genes_mapped) # Check how many genes were mapped successfully
length(EpNL_EpInv_rankings)   # 22662 v. 31950 
```

```{r}
# Create named vector w/ Entrez IDs as names and corresponding signed p values 
EpNL_EpInv_rankings_mapped <- EpNL_EpInv_rankings[EpNL_EpInv_genes_mapped$SYMBOL]
names(EpNL_EpInv_rankings_mapped) <- EpNL_EpInv_genes_mapped$ENTREZID

# Check all genes mapped correctly
length(EpNL_EpInv_rankings_mapped) == nrow(EpNL_EpInv_genes_mapped)
head(EpNL_EpInv_rankings_mapped)
```

```{r}
# Run GSEA
EpNL_EpInv_gsea_res <- clusterProfiler::gseGO(
  geneList = EpNL_EpInv_rankings_mapped, 
  OrgDb = org.Hs.eg.db, 
  ont = "BP", 
  minGSSize = 5, 
  maxGSSize = 500, 
  pvalueCutoff = 0.05,
  pAdjustMethod = "BH",
  nPermSimple = 10000, # Increase this value as need
  scoreType = "std"
)

# Convert Entrez IDs to gene symbols for readability
EpNL_EpInv_gsea_readable <- clusterProfiler::setReadable(
  EpNL_EpInv_gsea_res,
  OrgDb = org.Hs.eg.db,
  keyType = "ENTREZID"
  )

saveRDS(EpNL_EpInv_gsea_readable, "data/GSEA/EpNL_EpInv_gsea_res.rds")

# Convert gseaResult object into dataframe
EpNL_EpInv_gsea_res_df <- as.data.frame(EpNL_EpInv_gsea_readable)
write_csv(EpNL_EpInv_gsea_res_df, "data/GSEA/EpNL_EpInv_gsea_res.csv")
```

### Visualization

```{r}
# Dotplot
EpNL_EpInv_dotplot <- dotplot(EpNL_EpInv_gsea_res, showCategory = 15) +
  ggtitle("Top 15 GO Terms from GSEA\nin Ep NL v. Ep Inv") +
  theme(axis.text.y = element_text(size = 9, family = "Arial"))

ggsave("EpNL_EpInv_dotplot.png", EpNL_EpInv_dotplot, path = "plots/GSEA",
       width = 6, height = 7, dpi = 300)
```

```{r}
# Barplot
# Select the top 10 upregulated pathways according to NES 
EpNL_EpInv_top10_up <- EpNL_EpInv_gsea_res_df |> 
  arrange(desc(NES)) |>
  slice_head(n = 10)

# Select the top 10 downregulated pathways according to NES 
EpNL_EpInv_top10_down <- EpNL_EpInv_gsea_res_df |>
  arrange(NES) |>
  slice_head(n = 10)

# Combine dfs 
EpNL_EpInv_top10_updown <- bind_rows(EpNL_EpInv_top10_up, EpNL_EpInv_top10_down)

# Generate barplot
EpNL_EpInv_barplot <- EpNL_EpInv_top10_updown |>
  ggplot(aes(x = NES, y = reorder(Description, NES), fill = p.adjust)) + 
  geom_col() + 
  scale_fill_gradient(low = "red", high = "blue") +
  labs(x = "Normalized Enrichment Score (NES)", y = "Pathway", title = "Top 10 Upregulated and Downregulated \nPathways in Ep Inv Over Ep NL") +
  theme_bw() 
  
ggsave("EpNL_EpInv_barplot.png", EpNL_EpInv_barplot, path = "plots/GSEA",
       width = 9, height = 6, dpi = 300)
```

```{r}
# Cnetplot
EpNL_EpInv_cnetplot <- enrichplot::cnetplot(
  EpNL_EpInv_gsea_readable,
  categorySizeBy = ~NES, # or "pvalue", "p.adjust"
  showCategory = 5
  ) 

ggsave("EpNL_EpInv_cnetplot.png", EpNL_EpInv_cnetplot, path = "plots/GSEA",
       width = 12, height = 8, dpi = 300)
```

```{r}
# Upsetplot
EpNL_EpInv_gsea_top10 <- EpNL_EpInv_gsea_res

# Top 5 positively enriched
top_up <- EpNL_EpInv_gsea_res@result |>
  arrange(desc(NES)) |>
  slice_head(n = 5)

# Top 5 negatively enriched
top_down <- EpNL_EpInv_gsea_res@result |>
  arrange(NES) |>
  slice_head(n = 5)

# Combine
EpNL_EpInv_gsea_top10@result <- bind_rows(top_up, top_down)

# Create gene sets for UpSetR input 
EpNL_EpInv_gene_sets <- geneInCategory(EpNL_EpInv_gsea_top10)

# Add arrows to pathway names
names(EpNL_EpInv_gene_sets) <- c(
  paste0("↑ ", top_up$Description),
  paste0("↓ ", top_down$Description)
)

# Wrap names for cleanness
names(EpNL_EpInv_gene_sets) <- stringr::str_wrap(names(EpNL_EpInv_gene_sets), width = 40)

# Generate upsetplot
EpNL_EpInv_upset <- upset(
  fromList(EpNL_EpInv_gene_sets),
  nsets = 10,                              # Plot top 5 up and down pathways
  nintersects = 15,                        # Number of intersections to plot
  mainbar.y.label = "Gene intersections",
  sets.x.label = "Leading-edge genes",
  text.scale = 1.3,
  shade.color = "#D981AA",
  matrix.color = "#C43B7B",
  order.by = "freq", 
  )

ggsave("EpNL_EpInv_upset.png",
       plot = as.ggplot(EpNL_EpInv_upset),
       path = "plots/GSEA",
       width = 8,
       height = 10,
       dpi = 300)
```

```{r}
# Emapplot
EpNL_EpInv_upset <- emapplot(EpNL_EpInv_gsea_readable, max.overlaps = 15) 
```


