---
title: "04_Reclustering_BCRF"
format: gfm
---

# 04_Reclustering_BCRF

The goal of this script is to separate the integrated BCRF dataset into Epithelial and Stromal subsets. Then, rerun clustering and generate UMAPs for the new subsets. 

Combine v1 and v2 using seurat v5 rpca integration, did not do any other testing- Magda
combine v1 and v2, need batch correction, check w pca/ clustering - Siri
S3S v2 has more reads

From TBCRCn, RAHBTn, COMETn

## Set-up 

```{r}
# Load libraries
library(future)
library(Seurat)
library(Matrix)
library(ggplot2)
```

```{r}
# Set global options 
options(future.globals.maxSize = Inf)
print(getOption("future.globals.maxSize"))
options(Seurat.object.assay.version = "v5")
```

```{r}
# Load in integrated data set (dataset used for initial EpNLv.EpStr & StrNLv.StrInv DESeq2/GSEA)
BCRF <- readRDS("data/integrated_BCRF_set_1-10_&_EN.rds")

# Join counts layers from BCRF RNA set 1-3 and BCRF RNA set 4-10
BCRF <- JoinLayers(BCRF)
```

## Subsetting 

```{r}
# Subset BCRF data to only include "Ep" tissue type samples 
subsetEp <- subset(
  BCRF,
  subset = tissue %in% "Ep"
  )

saveRDS(subsetEp, "data/subsetted_BCRF_Ep.rds")

# Subset BCRF data to only include "Str" tissue type samples
subsetStr <- subset(
  BCRF,
  subset = tissue %in% "Str"
)

saveRDS(subsetStr, "data/subsetted_BCRF_Str.rds")
```

## Pre-processing

Pre-processing is defined by functions for normalization, highly variable genes, and scaling of the data. Normalization with NormalizeData() of integrated_BCRF_set_1-10_&_EN.rds used normalization method of LogNormalize wherein: cell counts independent of other cells, so no need to rerun in subsettted data; it would be the same. 

The first step of pre-processing is to resplit the merged layers of the subsetted dataset into batch layers, so that IntegrateLayers() can correct for batch effects. The splitting in the original integrated dataset occured when merge() implicitly created layers. 
```{r}
# Reconstruct the original batch grouping from cell name prefixes
BCRF$batch <- ifelse(grepl("^BCRF1_3&EN_", colnames(BCRF)), "BCRF1_3&EN", "BCRF4_10")

# Sanity check — should roughly mirror the counts.1/counts.2 split sizes from the original object
table(BCRF$batch)
```

batch effects 

## Elaine's code 
```{r}

# Assign dataset labels
# obj$dataset = sapply(Cells(obj), function(x) strsplit(x, '_')[[1]][1])

# Normalize, scale from BCRF sets.R 
obj <- NormalizeData(obj) # Adjust for seq depth, lib size to compare bw samples, includes log transform
obj <- FindVariableFeatures(obj)
# obj$log2counts = log2(obj$nCount_RNA + 1)
obj <- ScaleData(obj) # Scales normalized, log-transformed counts for PCA or clustering
obj <- RunPCA(obj, features = VariableFeatures(object = obj))
obj <- FindNeighbors(obj, dims = 1:30, reduction = "pca")


# Integrate layers using RPCA
obj <- IntegrateLayers(
  object = obj, method = RPCAIntegration, k.weight = 90,
  orig.reduction = "pca", new.reduction = "integrated.rpca",
  verbose = FALSE
)

head(obj@meta.data)

# Find clusters after integration
obj <- FindClusters(obj, resolution = 2.5, cluster.name = "rpca_clusters")
obj <- RunUMAP(obj, reduction = "integrated.rpca", dims = 1:30, reduction.name = "umap.rpca")

# obj <- JoinLayers(obj) 


##########

setwd("/Users/eguo/Downloads/integrated BCRF set 1-10 & EN")

DimPlot(obj, reduction = "umap.rpca", label = TRUE)
ggsave("umap.rpca.png", width = 8, height = 8)

obj$tissue[is.na(obj$tissue)] <- 'NA'
DimPlot(obj, reduction = "umap.rpca", label = TRUE, group.by = "tissue")
ggsave("tissue.png", width = 8, height = 8)

DimPlot(obj, reduction = "umap.rpca", group.by = "protocol")
ggsave("protocol.png", width = 8, height = 8)

DimPlot(obj, reduction = "umap.rpca", group.by = "dxs")
ggsave("dxs.png", width = 8, height = 8)  

DimPlot(obj, reduction = "umap.rpca", group.by = "dx")
ggsave("dx.png", width = 12, height = 8)

n_patients <- length(unique(obj$patient))
DimPlot(obj, reduction = "umap.rpca", group.by = "patient") +
  ggtitle(paste("n patients = ", n_patients))
ggsave("patient.png", width = 12, height = 8)  

DimPlot(obj, reduction = "umap.rpca", group.by = "sequence batch number")
ggsave("sequence batch number.png", width = 12, height = 8)

# log2 of total counts (library size) per sample
obj$log2counts <- log2(obj$nCount_RNA + 1)
log2counts <- FeaturePlot(obj,
            features = "log2counts",
            combine = TRUE, 
            label.size = 1) 
print(log2counts)
ggsave("log2counts.png", width = 8, height = 8)  


FeaturePlot(obj, features = c("ESR1", "ERBB2", "MYC", "TP53", "PIK3CA"), cols = c("red", "blue"))
ggsave("genes.png", width = 8, height = 8)

saveRDS(obj, file = "/Users/eguo/Downloads/integrated_BCRF_set_1-10_&_EN.rds")




#########

cluster_markers_all <- FindAllMarkers(obj, 
                       only.pos = FALSE,  # Includes both pos and neg markers
                       min.pct = 0.1,     # Genes must be expressed in min 10% of samples in a cluster
                       logfc.threshold = 0.25,  # LFC threshold
                       test.use = DESeq2)

head(cluster_markers_all)

# Filter by adj pval and Log2FC
cluster_markers_filtered <- cluster_markers_all %>%
  filter(p_val_adj < 0.05)

top_genes_per_cluster <- cluster_markers_filtered %>%
  filter(avg_log2FC > 1 | avg_log2FC < -1) %>%   # Filter for log2FC > 1 or < -1
  group_by(cluster) %>%                          # Group by cluster
  top_n(100, wt = abs(avg_log2FC)) %>%           # Select top DE genes by avg_log2FC for each cluster
  ungroup()
```

