---
title: "05_NMF"
format: html
---

# 05_NMF

The goal of this script will be to run non-negative matrix factorization (NMF) on the compartamentally subsetted (epithelial and stromal) integrated BCRF datasets. NMF is a linear dimensionality reduction technique. In the context of RNA-seq analysis, NMF calculates embeddings that separate data into interpretable groups. The embeddings can be stored into the reductions slot of a Seurat object and used in downstream analyses. 

General workflow: DESeq2 normalized counts > VST > NMF

## Set-up

```{r}
library(Seurat)
library(DESeq2)
library(BiocParallel)
BiocParallel::register(SnowParam(7))
library(NMF)
set.seed(67)

# Load in data
BCRF_Ep <- readRDS("data/integrated_BCRF_set_1-10_&_EN_v2_Ep.rds")
BCRF_Str <- readRDS("data/integrated_BCRF_set_1-10_&_EN_v2_Str.rds")
```

## NMF of Ep-Subsetted BCRF

### Normalization via DESeq2 

```{r}
# Get counts data
Ep_counts <- GetAssayData(BCRF_Ep, assay = "RNA", layer = "counts")

# Get metadata
Ep_metadata <- BCRF_Ep@meta.data

# Check dimensions are the same b/t counts & metadata
dim(Ep_counts)
dim(Ep_metadata)
identical(colnames(Ep_counts), rownames(Ep_metadata)) # TRUE

# Create DESeqDataSet from subset counts and group information 
Ep_dds <- DESeqDataSetFromMatrix(
  countData = Ep_counts,
  colData = Ep_metadata,
  design = ~ 1 # This design parameter means "no design", for normalization only
  )

# Keep genes with at least 1 counts in at least 10 samples 
Ep_keep <- rowSums(counts(Ep_dds) >= 1) >= 10
Ep_dds <- Ep_dds[Ep_keep,]

# Run DESeq normalization 
Ep_dds <- DESeq(Ep_dds, parallel=TRUE)

# Applies VST to read counts, useful for downstream clustering or PCA
Ep_vsd <- varianceStabilizingTransformation(Ep_dds)

# Saves transformed data
write.csv(round(assay(Ep_vsd), 2), file = "data/NMF/Ep/BCRF_Ep_VST.csv")

# Check VST Results: Visualize w PCA or clustering to ensure that VST results capture the expected variance.
```

### Load-in & Transform VSD Data 
```{r}
# Load in VST results for downstream NMF
Ep_rna_matrix_3 <- read.csv("data/NMF/Ep/BCRF_Ep_VST.csv", row.names = 1)

# RNA matrix is transposed and scaled. Centering is set to FALSE, which means the data are normalized by their respective variances but not mean-centered.
Ep_edata <- t(scale(t(Ep_rna_matrix_3),center = FALSE))
```

### Get Variance for Genes & Filter for Top Variance Genes 

```{r}
# Variance is computed for each gene (row) across samples (columns).
# Genes with variance greater than the filter value (0; non-negative) are retained for analysis.
# This step removes genes that do not vary across samples.
# var.data contains the variance values for all genes

Ep_var.data <- apply(Ep_edata, 1, var)
vfilter <- 0
# Retain genes with variance greater than the filter value
Ep_vdata <- Ep_edata[which(Ep_var.data > vfilter ),]

# Count the number of genes that passed the filter (rows in vdata)
Ep_num_genes_above_filter <- nrow(Ep_vdata) 
# Print dimensions and the count of genes
dim(Ep_vdata) # 37207 522
print(Ep_num_genes_above_filter) # 37k+ is a lot

# Generate histogram to see at what number of genes where variance becomes high  
hist(Ep_var.data)

# Generate ranked variance plot to see where to cut off genes 
plot(sort(Ep_var.data, decreasing = TRUE), type = "l",
     xlab = "Gene rank",
     ylab = "Variance")

# Keep top 3000 genes 
head(sort(Ep_var.data, decreasing = TRUE))
Ep_top.genes <- names(sort(Ep_var.data, decreasing = TRUE))[1:3000]
Ep_vdata <- Ep_edata[Ep_top.genes, ]

saveRDS(Ep_vdata, "data/NMF/Ep/Ep_vdata.rds")
```

### Run NMF 

The range of k (4:10) was chosen based on the fact there are 7 unique conditions as defined by "dxs" in the Ep-subsetted BCRF dataset. 

```{r}
# Load in gene variance matrix 
Ep_vdata <- readRDS("../data/NMF/Ep/Ep_vdata")

# run NMF from 4 to 10 factors, 30 times (takes time and memory) and catch errors
# NMF is applied to filtered data vdata, testing 4-10 factors or clusters
# nrun specifies that each NMF run should be repeated n times for robustness
Ep_res <- nmf(Ep_vdata, 4:10, nrun=30, .opt="vp15", seed = 67)

# Save results and continue with further analysis
save(Ep_res, file = "data/NMF/Ep/Ep_res.RData")

# after NMF results, determine optimal number of factors using silhouette score or cophenetic correlation
```

## NMF of Str-Subsetted BCRF 

### Normalization via DESeq2

```{r}
# Get counts data
Str_counts <- GetAssayData(BCRF_Str, assay = "RNA", layer = "counts")

# Get metadata
Str_metadata <- BCRF_Str@meta.data

# Find all samples that have zero nonzero counts
# How many nonzero genes does each cell have, after filtering?
nonzero_per_cell <- colSums(Str_counts > 0)
summary(nonzero_per_cell) # Min. = 0

# Any cells with zero or near-zero nonzero genes?
sum(nonzero_per_cell == 0)  # 1 
which(nonzero_per_cell < 5) # set1_3_11886_364_NL_S_S2_R1_001 50 

# Filter out samples that have zero counts for all genes
# Defined in troubleshooting section 
Str_metadata <- Str_metadata[nonzero_per_cell > 0, ]
Str_counts <- Str_counts[, nonzero_per_cell > 0]

# Check dimensions are the same b/t counts & metadata
dim(Str_counts)
dim(Str_metadata)
identical(colnames(Str_counts), rownames(Str_metadata)) # TRUE

# Create DESeqDataSet from subset counts and group information 
Str_dds <- DESeqDataSetFromMatrix(
  countData = Str_counts,
  colData = Str_metadata,
  design = ~ 1 # This design parameter means "no design", for normalization only
  )

# Keep genes with at least 1 counts in at least 10 samples 
Str_keep <- rowSums(counts(Str_dds) >= 1) >= 10
Str_dds <- Str_dds[Str_keep,]

# Run DESeq normalization 
Str_dds <- DESeq(Str_dds, parallel=TRUE)

# Applies VST to read counts, useful for downstream clustering or PCA
Str_vsd <- varianceStabilizingTransformation(Str_dds)

# Saves transformed data
write.csv(round(assay(Str_vsd), 2), file = "data/NMF/Str/BCRF_Str_VST.csv")

# Check VST Results: Visualize w PCA or clustering to ensure that VST results capture the expected variance.
```

### Load-in & Transform VSD Data 
```{r}
# Load in VST results for downstream NMF
Str_rna_matrix_3 <- read.csv("data/NMF/Ep/BCRF_Str_VST.csv", row.names = 1)

# RNA matrix is transposed and scaled. Centering is set to FALSE, which means the data are normalized by their respective variances but not mean-centered.
Str_edata <- t(scale(t(Str_rna_matrix_3),center = FALSE))
```

### Get Variance for Genes & Filter for Top Variance Genes 

```{r}
# Variance is computed for each gene (row) across samples (columns).
# Genes with variance greater than the filter value (0; non-negative) are retained for analysis.
# This step removes genes that do not vary across samples.
# var.data contains the variance values for all genes

Str_var.data <- apply(Str_edata, 1, var)
vfilter <- 0
# Retain genes with variance greater than the filter value
Str_vdata <- Str_edata[which(Str_var.data > vfilter ),]

# Count the number of genes that passed the filter (rows in vdata)
Str_num_genes_above_filter <- nrow(Str_vdata) 
# Print dimensions and the count of genes
dim(Str_vdata) # 37207 522
print(Str_num_genes_above_filter) # 37k+ is a lot

# Generate histogram to see at what number of genes where variance becomes high  
hist(Str_var.data)

# Generate ranked variance plot to see where to cut off genes 
plot(sort(Str_var.data, decreasing = TRUE), type = "l",
     xlab = "Gene rank",
     ylab = "Variance")

# Keep top 3000 genes 
head(sort(Str_var.data, decreasing = TRUE))
Str_top.genes <- names(sort(Str_var.data, decreasing = TRUE))[1:3000]
Str_vdata <- Str_edata[Str_top.genes, ]

saveRDS(Str_vdata, "data/NMF/Str/Str_vdata.rds")
```

### Run NMF 

The range of k (4:10) was chosen based on the fact there are 7 unique conditions as defined by "dxs" in the Ep-subsetted BCRF dataset. 

```{r}
# Load in gene variance matrix 
Str_vdata <- readRDS("../data/NMF/Str/Str_vdata")

# Run NMF 
Str_res <- nmf(Str_vdata, 4:10, nrun=30, .opt="vp15", seed = 67)

# Save results and continue with further analysis
save(Str_res, file = "data/NMF/Str/Str_res.RData")

# after NMF results, determine optimal number of factors using silhouette score or cophenetic correlation
```