---
title: "06_NMF_Analysis"
format: gfm
---

# 06_NMF_Analysis

The goal of this script is to generate visualizations i.e. consensus maps, silhouette plots, etc to determine the optimal number of clusters for the newly Ep- and Str-subsetted BCRF datasets.

Also, functions for 

## Set-up 

Perform silhouette analysis, saves gene, and sample associations into CSV.

```{r}
library(NMF)
library(argparse)
library(viridis)
library(fgsea)
library(tidyr)
library(rms)

# Load-in data 
load("data/NMF/Ep/Ep_res.RData")
load("data/NMF/Str/Str_res.RData")
```

## Visualizations

After NMF, assess clustering quality through visualizations: rank survey, consensus maps, and silhouette plots. 
The comprehensive consensus map with all k ranks on one .jpeg results in a graphics error, but individually generating is fine. 

### Ep 

```{r}
# Rank survey plot
# This visually represents how well the NMF model has performed across different ranks
png(file = "plots/NMF/Ep/BCRF_Ep_NMF_rank_survey_res200.png", width = 1724, height = 1052, res = 200)
plot(Ep_res) 
dev.off()

# same here, we evaluate how robust are the clusters
# Via consensusmap, which are produced for various ranks to assess robustness of the clusters 
# jpeg(file = "data/NMF_practice/BCRF_Ep_NMF_consensus_3-15.jpg",width = 6800, height = 4202)
# consensusmap(res)
# dev.off()

# k=4
png(file = "plots/NMF/Ep/BCRF_Ep_NMF_consensus_k4.png",width = 3000, height = 2000)
consensusmap(Ep_res$consensus$`4`)
dev.off()
# k=5
png(file = "plots/NMF/Ep/BCRF_Ep_NMF_consensus_k5.png",width = 3000, height = 2000)
consensusmap(Ep_res$consensus$`5`)
dev.off()
# k=6
png(file = "plots/NMF/Ep/BCRF_Ep_NMF_consensus_k6.png",width = 3000, height = 2000)
consensusmap(Ep_res$consensus$`6`)
dev.off()
# k=7
png(file = "plots/NMF/Ep/BCRF_Ep_NMF_consensus_k7.png",width = 3000, height = 2000)
consensusmap(Ep_res$consensus$`7`)
dev.off()
# k=8
png(file = "plots/NMF/Ep/BCRF_Ep_NMF_consensus_k8.png",width = 3000, height = 2000)
consensusmap(Ep_res$consensus$`8`)
dev.off()
# k=9
png(file = "plots/NMF/Ep/BCRF_Ep_NMF_consensus_k9.png",width = 3000, height = 2000)
consensusmap(Ep_res$consensus$`9`)
dev.off()
# k=10
png(file = "plots/NMF/Ep/BCRF_Ep_NMF_consensus_k10.png",width = 3000, height = 2000)
consensusmap(Ep_res$consensus$`10`)
dev.off()

# silhouette plot (avoid solutions with negative values), to visualize clustering quality across ranks. 
# This evaluates separation distances between clusters, with negative values indicating poor clustering.
png(file="plots/NMF/Ep/BCRF_Ep_NMF_silhouette_hires.png", width = 10000, height = 7500)
par(mfrow=c(3,4))
plot(silhouette(Ep_res$fit$`4`))
plot(silhouette(Ep_res$fit$`5`))
plot(silhouette(Ep_res$fit$`6`))
plot(silhouette(Ep_res$fit$`7`))
plot(silhouette(Ep_res$fit$`8`))
plot(silhouette(Ep_res$fit$`9`))
plot(silhouette(Ep_res$fit$`10`))
dev.off()

# Add res = 300 argument to silhouette plot if text difficult to read
png(file="plots/NMF/Ep/BCRF_Ep_NMF_silhouette_hires_res300.png", width = 10000, height = 7500, res = 300)
par(mfrow=c(3,4))
plot(silhouette(Ep_res$fit$`4`))
plot(silhouette(Ep_res$fit$`5`))
plot(silhouette(Ep_res$fit$`6`))
plot(silhouette(Ep_res$fit$`7`))
plot(silhouette(Ep_res$fit$`8`))
plot(silhouette(Ep_res$fit$`9`))
plot(silhouette(Ep_res$fit$`10`))
dev.off()

# graphics.off()
```

### Str

```{r}
# A plot of the NMF results (plot(res)) is generated and saved as JPEG.
# This visually represents how well the NMF model has performed across different ranks

png(file = "plots/NMF/Str/BCRF_Str_NMF_rank_survey_res200.png", width = 1724, height = 1052, res = 200)
plot(Str_res) 
dev.off()

# same here, we evaluate how robust are the clusters
# Via consensusmap, which are produced for various ranks to assess robustness of the clusters 
# jpeg(file = "data/NMF_practice/BCRF_Ep_NMF_consensus_3-15.jpg",width = 6800, height = 4202)
# consensusmap(res)
# dev.off()

# k=4
png(file = "plots/NMF/Str/BCRF_Str_NMF_consensus_k4.png",width = 3000, height = 2000)
consensusmap(Str_res$consensus$`4`)
dev.off()
# k=5
png(file = "plots/NMF/Str/BCRF_Str_NMF_consensus_k5.png",width = 3000, height = 2000)
consensusmap(Str_res$consensus$`5`)
dev.off()
# k=6
png(file = "plots/NMF/Str/BCRF_Str_NMF_consensus_k6.png",width = 3000, height = 2000)
consensusmap(Str_res$consensus$`6`)
dev.off()
# k=7
png(file = "plots/NMF/Str/BCRF_Str_NMF_consensus_k7.png",width = 3000, height = 2000)
consensusmap(Str_res$consensus$`7`)
dev.off()
# k=8
png(file = "plots/NMF/Str/BCRF_Str_NMF_consensus_k8.png",width = 3000, height = 2000)
consensusmap(Str_res$consensus$`8`)
dev.off()
# k=9
png(file = "plots/NMF/Str/BCRF_Str_NMF_consensus_k9.png",width = 3000, height = 2000)
consensusmap(Str_res$consensus$`9`)
dev.off()
# k=10
png(file = "plots/NMF/Str/BCRF_Str_NMF_consensus_k10.png",width = 3000, height = 2000)
consensusmap(Str_res$consensus$`10`)
dev.off()


# silohuette plot (avoid solutions with negative values), to visualize clustering quality across ranks. 
# This evaluates separation distances between clusters, with negative values indicating poor clustering.
png(file="plots/NMF/Str/BCRF_Str_NMF_silhouette_hires.png", width = 10000, height = 7500)
par(mfrow=c(3,4))
plot(silhouette(Str_res$fit$`4`))
plot(silhouette(Str_res$fit$`5`))
plot(silhouette(Str_res$fit$`6`))
plot(silhouette(Str_res$fit$`7`))
plot(silhouette(Str_res$fit$`8`))
plot(silhouette(Str_res$fit$`9`))
plot(silhouette(Str_res$fit$`10`))
dev.off()

png(file="plots/NMF/Str/BCRF_Str_NMF_silhouette_hires_res300.png", width = 10000, height = 7500, res = 300)
par(mfrow=c(3,4))
plot(silhouette(Str_res$fit$`4`))
plot(silhouette(Str_res$fit$`5`))
plot(silhouette(Str_res$fit$`6`))
plot(silhouette(Str_res$fit$`7`))
plot(silhouette(Str_res$fit$`8`))
plot(silhouette(Str_res$fit$`9`))
plot(silhouette(Str_res$fit$`10`))
dev.off()

# graphics.off()
```

## Extract genes list(s) 

The purpose of these functions is to extract the genes associated w/ each cluster for the different rank k values. 
The first function extracts genes for a specified rank while the second function compiles gene cluster information for every rank into a single CSV.

```{r}
# Takes NMF fit object for a specified rank and extracts genes associated w/ each cluster
# Extracts the basis matrix (gene-cluster associations), identifies the cluster with the highest weight for each gene, saves results to CSV, returns df for inspection.
extract_genes_per_cluster <- function(nmf_fit, output_file) {

  # Extract the basis matrix
  # This is the matrix obtained from the NMF fit, where each row = gene, and each column = cluster.

  basis_matrix <- basis(nmf_fit)
  
  # Find the cluster with the maximum weight for each gene
  # gene_clusters will be a vector where each entry = cluster w the highest association for each gene
  gene_clusters <- apply(basis_matrix, 1, which.max)
  
  # Create a data frame with gene names and their corresponding clusters
  gene_cluster_df <- data.frame(
    Gene = rownames(basis_matrix),
    Cluster = gene_clusters
  )
  
  # Save the df to CSV
  write.csv(gene_cluster_df, file = output_file, row.names = FALSE, quote = FALSE)
  
  # Return the df for inspection
  return(gene_cluster_df)
}

# Example for rank 8
gene_cluster_mapping <- extract_genes_per_cluster(Ep_res$fit$`8`, "data/NMF/Ep/NMF_genes_per_cluster_rank8.csv")
print(head(gene_cluster_mapping))
```

```{r}
# Extracts gene weights and their associated clusters for ranks (ie 4-10), saves all ranks’ data into single CSV
extract_genes_all_ranks <- function(nmf_result, ranks) {
  
  # Loop through each rank and extract information
  all_ranks_data <- data.frame()
  for (rank in ranks) {
    cat("Processing rank:", rank, "\n")
    
    # Extract the basis matrix for the current rank
    nmf_fit <- nmf_result$fit[[as.character(rank)]]
    basis_matrix <- basis(nmf_fit)
    
    # Find the cluster with the maximum weight for each gene
    max_weights <- apply(basis_matrix, 1, max)
    gene_clusters <- apply(basis_matrix, 1, which.max)
    
    # Create a data frame for the current rank
    rank_data <- data.frame(
      Gene = rownames(basis_matrix),
      Cluster = gene_clusters,
      Max_Weight = max_weights,
      Rank = rank
    )
    
    # Combine with all ranks data
    all_ranks_data <- rbind(all_ranks_data, rank_data)
  }
  
  return(all_ranks_data)
}
```

### Ep
```{r}
# Specify ranks and output directory
ranks <- 4:10

# Extract and save data for all ranks
Ep_all_ranks_data <- extract_genes_all_ranks(Ep_res, 4:10)
write.csv(Ep_all_ranks_data, "data/NMF/Ep/Ep_all_ranks_data.csv")
print(head(all_ranks_data))
```

### Str
```{r}
Str_all_ranks_data <- extract_genes_all_ranks(Str_res, 4:10)
write.csv(Str_all_ranks_data, "data/NMF/Str/Str_all_ranks_data.csv")
print(head(Str_all_ranks_data))
```

## Extract sample list 

The purpose of this function is to list which clusters best define each sample.  

```{r}
# Function to extract samples and clusters for all ranks
# Extracts the sample-to-cluster associations for the specified ranks
extract_samples_all_ranks <- function(nmf_result, ranks) {
  
  # Loop through each rank and extract sample-cluster associations
  all_ranks_sample_data <- data.frame()
  for (rank in ranks) {
    cat("Processing rank:", rank, "\n")
    
    # Extract the coefficient matrix for the current rank
    nmf_fit <- nmf_result$fit[[as.character(rank)]]
    coef_matrix <- coef(nmf_fit)
    
    # Find the cluster with the maximum coefficient for each sample
    sample_clusters <- apply(coef_matrix, 2, which.max)
    
    # Create a data frame for the current rank
    rank_sample_data <- data.frame(
      Sample = colnames(coef_matrix),
      Cluster = sample_clusters,
      Max_Coefficient = apply(coef_matrix, 2, max),
      Rank = rank
    )
    
    # Combine with all ranks data
    all_ranks_sample_data <- rbind(all_ranks_sample_data, rank_sample_data)
  }
  
  return(all_ranks_sample_data)
}
```

### Ep
```{r}
# Specify ranks and output directory
ranks <- 4:10

# Extract and save data for all ranks
Ep_all_ranks_sample_data <- extract_samples_all_ranks(Ep_res, ranks)
print(head(Ep_all_ranks_sample_data))
write.csv(Ep_all_ranks_sample_data, "data/NMF/Ep/Ep_all_ranks_sample_data.csv")
```

### Str
```{r}
Str_all_ranks_sample_data <- extract_samples_all_ranks(Str_res, ranks)
print(head(Str_all_ranks_sample_data))
write.csv(Str_all_ranks_sample_data, "data/NMF/Str/Str_all_ranks_sample_data.csv")
```

