---
title: "06_Reclustering_BCRF"
format: gfm
---

# 06_Reclustering_BCRF

The goal of this script is to take the NMF results, embed them into the Seurat objects for the newly subsetted Ep and Str compartments, and re-generate compartment-specific UMAPs.

We will be working with two sets of NMF data: uncorrected for batch effect and corrected for batch effect. The workflow for each type of data will be as follows:

- Uncorrected: Map sample IDs and clusters info from NMF data onto the unintegrated and rpca integrated UMAP reductions in the existing Seurat object.
- Corrected: Create a new UMAP reduction based on individual rank k values from NMF.

## Set-up 

```{r}
# Load libraries
library(NMF)
library(Seurat)
library(ggplot2)
library(stringr)
library(tidyverse)

# Set global options 
options(future.globals.maxSize = Inf)
print(getOption("future.globals.maxSize"))
options(Seurat.object.assay.version = "v5")

# Load-in data 
# Original Seurat objects 
BCRF_Ep <- readRDS("data/integrated_BCRF_set_1-10_&_EN_v2_Ep.rds")
BCRF_Str <- readRDS("data/integrated_BCRF_set_1-10_&_EN_v2_Str.rds")

# NMF Results 
load("data/NMF/Ep/Ep_res.RData")                        
load("data/NMF/Str/Str_res.RData")
load("data/NMF/Ep_corrected/Ep_res_corrected.RData")
load("data/NMF/Str_corrected/Str_res_corrected.RData")

# Filtered VST matrices (before NMF)
Ep_vdata <- readRDS("data/NMF/Ep/Ep_vdata.rds")    
Str_vdata <- readRDS("data/NMF/Str/Str_vdata.rds")
Ep_vdata_corrected <- readRDS("data/NMF/Ep_corrected/Ep_vdata_corrected.rds")
Str_vdata_corrected <- readRDS("data/NMF/Str_corrected/Str_vdata_corrected.rds")

# Sample and cluster data 
Ep_sample_clusters <- read.csv("data/NMF/Ep/Ep_all_ranks_sample_data.csv")
Ep_sample_clusters_corrected <- read.csv("data/NMF/Ep_corrected/Ep_all_ranks_sample_data_corrected.csv")
Str_sample_clusters <- read.csv("data/NMF/Str/Str_all_ranks_sample_data.csv")
Str_sample_clusters_corrected <- read.csv("data/NMF/Str_corrected/Str_all_ranks_sample_data_corrected.csv")
```

# k=4

## Str (uncorrected)

### Map NMF result sampleIDs onto BCRF_Str Seurat object 

```{r}
# Extract sample and cluster info from NMF results where k=4
Str_sample_clusters_k4 <- Str_sample_clusters |> 
  select(Sample, Cluster, Rank) |>
  filter(Rank == 4)

# Fix mismatch between sample names 
Str_sample_clusters_k4$Sample_clean <- gsub("\\.", "-", Str_sample_clusters_k4$Sample)

# Subset BCRF_Str by NMF samples only (remove sample w/ 0 counts)
BCRF_Str <- subset(BCRF_Str, cells = Str_sample_clusters_k4$Sample_clean)

extracted <- str_match(Str_sample_clusters_k4$Sample_clean, "set\\d+_\\d+_(\\d+)")[,2]
Str_sample_clusters_k4$sample <- extracted

# force both sides to the same type before matching -- this is the #1 source of silent NAs
Str_sample_clusters_k4$sample <- as.character(Str_sample_clusters_k4$sample)
BCRF_Str$sample <- as.character(BCRF_Str$sample)

# map onto the Seurat object
BCRF_Str$clusters_k4 <- Str_sample_clusters_k4$Cluster[
  match(BCRF_Str$sample, Str_sample_clusters_k4$sample)
]
BCRF_Str$clusters_k4 <- factor(BCRF_Str$clusters_k4)

# sanity check -- how many cells got NA (i.e. sample wasn't found in your NMF table)?
table(BCRF_Str$clusters_k4, useNA = "ifany")

# 1   2   3   4 
# 206  88  82  72 
# Matches silhouette plot cluster numbers 
```

### Plot on existing reductions 

```{r}
# Unintegrated reduction
p1 <- DimPlot(BCRF_Str, reduction = "umap.unintegrated") +
  ggtitle("Original Clusters (unintegrated)")
p2 <- DimPlot(BCRF_Str, reduction = "umap.unintegrated", group.by = "clusters_k4") +
  ggtitle("NMF Clusters (k=4)")

ggsave("plots/NMF_reclustering/Str/k4/Str_unintegrated_UMAP_comparison_k4.png", p1 + p2, width = 12, height = 6, dpi = 300)

# RPCA reduction
p3 <- DimPlot(BCRF_Str, reduction = "umap.rpca") +
  ggtitle("Original Clusters (RPCA)")
p4 <- DimPlot(BCRF_Str, reduction = "umap.rpca", group.by = "clusters_k4") +
  ggtitle("NMF Clusters (k=4)")

ggsave("plots/NMF_reclustering/Str/k4/Str_RPCA_UMAP_comparison_k4.png", p3 + p4, width = 12, height = 6, dpi = 300)

# RPCA dxs comparison 
p5 <- DimPlot(BCRF_Str, reduction = "umap.rpca", group.by = "dxs") +
  ggtitle("dxs (RPCA)") 

ggsave("plots/NMF_reclustering/Str/k4/Str_RPCA_dxs_UMAP_comparison_k4.png", p5 + p4, width = 12, height = 6, dpi = 300)
```

## Str_corrected

### Reorient matrices correctly

```{r}
# Extract k=4 specific NMF result
Str_res_corrected_k4 <- Str_res_corrected$fit[["4"]]

# Extract basis and coefficient matrices
w_StrC_k4 <- basis(Str_res_corrected_k4)
h_StrC_k4 <- coef(Str_res_corrected_k4)

# Transpose coefficient matrix
StrC_NMF_embeddings_k4 <- t(h_StrC_k4)
```

### Check that cell names line up

```{r}
# Edit column names 
colnames(StrC_NMF_embeddings_k4) <- paste0("NMF_", seq_len(ncol(StrC_NMF_embeddings_k4)))

# Put sample names as rownames of embeddings object 
rownames(StrC_NMF_embeddings_k4) <- colnames(Str_vdata_corrected)  # whatever you fed into nmf()

# Change "." to "-" in NMF embedding sample names 
rownames(StrC_NMF_embeddings_k4) <- gsub("\\.", "-", rownames(StrC_NMF_embeddings_k4))

# Subset BCRF_Str by NMF samples only 
BCRF_Str <- subset(BCRF_Str, cells = rownames(StrC_NMF_embeddings_k4))

# then re-check
identical(rownames(StrC_NMF_embeddings_k4), colnames(BCRF_Str)) # TRUE 
```

### Create DimReducObject

```{r}
BCRF_Str[["nmf_k4"]] <- CreateDimReducObject(
  embeddings = StrC_NMF_embeddings_k4,
  loadings   = w_StrC_k4,                # optional but nice to keep (genes x k)
  key        = "NMF_",                   # must end in underscore, must be unique
  assay      = DefaultAssay(BCRF_Str)
)
```

### Run/Generate UMAP on NMF reduction instead of PCA

```{r}
BCRF_Str <- RunUMAP(
  BCRF_Str,
  reduction = "nmf_k4",
  dims = 1:ncol(StrC_NMF_embeddings_k4), # or a subset of factors if some are junk
  reduction.name = "umap_nmf_k4_",
  reduction.key = "umapnmf_k4_"
)
```

```{r}
# Visualizations 
setwd("plots/NMF_reclustering/Str_corrected/k4")

DimPlot(BCRF_Str, reduction = "umap_nmf_k4_", label = TRUE)
ggsave("umap.nmf.k5.png", width = 8, height = 8)

DimPlot(BCRF_Str, reduction = "umap_nmf_k4_", group.by = "protocol")
ggsave("protocol.png", width = 8, height = 8)

DimPlot(BCRF_Str, reduction = "umap_nmf_k4_", group.by = "dxs")
ggsave("dxs.png", width = 8, height = 8)  

DimPlot(BCRF_Str, reduction = "umap_nmf_k4_", group.by = "dx")
ggsave("dx.png", width = 12, height = 8)

n_patients <- length(unique(BCRF_Str$patient))
DimPlot(BCRF_Str, reduction = "umap_nmf_k4_", group.by = "patient") +
  ggtitle(paste("n patients = ", n_patients))
ggsave("patient.png", width = 12, height = 8)  

DimPlot(BCRF_Str, reduction = "umap_nmf_k4_", group.by = "sequence.batch.number")
ggsave("sequence batch number.png", width = 12, height = 8)
```

### Map NMF result sampleIDs onto BCRF_Str Seurat object 

```{r}
# Extract sample and cluster info from NMF results where k=5 
Str_sample_clusters_corrected_k4 <- Str_sample_clusters_corrected |> 
  select(Sample, Cluster, Rank) |>
  filter(Rank == 4)

# Fix mismatch between sample names 
Str_sample_clusters_corrected_k4$Sample_clean <- gsub("\\.", "-", Str_sample_clusters_corrected_k4$Sample)

# Subset BCRF_Str by NMF samples only (remove sample w/ 0 counts)
BCRF_Str <- subset(BCRF_Str, cells = Str_sample_clusters_corrected_k4$Sample_clean)

extracted <- str_match(Str_sample_clusters_corrected_k4$Sample_clean, "set\\d+_\\d+_(\\d+)")[,2]
Str_sample_clusters_corrected_k4$sample <- extracted

# force both sides to the same type before matching -- this is the #1 source of silent NAs
Str_sample_clusters_corrected_k4$sample <- as.character(Str_sample_clusters_corrected_k4$sample)
BCRF_Str$sample <- as.character(BCRF_Str$sample)

# map onto the Seurat object
BCRF_Str$clusters_k4_corrected <- Str_sample_clusters_corrected_k4$Cluster[
  match(BCRF_Str$sample, Str_sample_clusters_corrected_k4$sample)
]
BCRF_Str$clusters_k4_corrected <- factor(BCRF_Str$clusters_k4_corrected)

# sanity check -- how many cells got NA (i.e. sample wasn't found in your NMF table)?
table(BCRF_Str$clusters_k4_corrected, useNA = "ifany")

# 1   2   3   4 
# 105  98  55 190 
# Matches silhouette plot cluster numbers 
```

### Plot on existing reductions 

```{r}
# Integrated reduction
p1 <- DimPlot(BCRF_Str, reduction = "umap_nmf_k4_") +
  ggtitle("Original Clusters (Integrated)")
p2 <- DimPlot(BCRF_Str, reduction = "umap_nmf_k4_", group.by = "clusters_k4_corrected") +
  ggtitle("NMF Clusters (k=4)")

ggsave("plots/NMF_reclustering/Str_corrected/k4/Str_Integrated_UMAP_comparison_k4.png", p1 + p2, width = 12, height = 6, dpi = 300)

# Integrated dxs comparison 
p3 <- DimPlot(BCRF_Str, reduction = "umap_nmf_k4_", group.by = "dxs") +
  ggtitle("dxs (Integrated)") 

ggsave("plots/NMF_reclustering/Str_corrected/k4/Str_Integrated_dxs_UMAP_comparison_k4.png", p3 + p2, width = 12, height = 6, dpi = 300)
```

# k=5

## Ep (uncorrected)

### Map NMF result sampleIDs onto BCRF_Ep Seurat object 

```{r}
# Extract sample and cluster info from NMF results where k=5 
Ep_sample_clusters_k5 <- Ep_sample_clusters |> 
  select(Sample, Cluster, Rank) |>
  filter(Rank == 5)
```

We want to map sample IDs from NMF results onto the existing unintegrated and rpca integrated UMAP reductions of the Seurat object.

Sample-level IDs in the Seurat object are under the "sample" column of its meta.data as a string of numbers. Since sample names in the NMF results data contains additional info, use regex to extract just the number string needed to match the Seurat object sample IDS.  

```{r}
extracted <- str_match(Ep_sample_clusters_k5$Sample, "set\\d+_\\d+_(\\d+)")[,2]
Ep_sample_clusters_k5$sample <- extracted
```

```{r}
# force both sides to the same type before matching -- this is the #1 source of silent NAs
Ep_sample_clusters_k5$sample <- as.character(Ep_sample_clusters_k5$sample)
BCRF_Ep$sample <- as.character(BCRF_Ep$sample)

# map onto the Seurat object
BCRF_Ep$clusters_k5 <- Ep_sample_clusters_k5$Cluster[
  match(BCRF_Ep$sample, Ep_sample_clusters_k5$sample)
]
BCRF_Ep$clusters_k5 <- factor(BCRF_Ep$clusters_k5)

# sanity check -- how many cells got NA (i.e. sample wasn't found in your NMF table)?
table(BCRF_Ep$clusters_k5, useNA = "ifany")

#  1   2   3   4   5 
# 145 101  65 136  75 
# Matches silhouette plot cluster numbers 
```

### Plot on existing reductions 

```{r}
# Unintegrated reduction
p1 <- DimPlot(BCRF_Ep, reduction = "umap.unintegrated") +
  ggtitle("Original Clusters (unintegrated)")
p2 <- DimPlot(BCRF_Ep, reduction = "umap.unintegrated", group.by = "clusters_k5") +
  ggtitle("NMF Clusters (k=5)")

ggsave("plots/NMF_reclustering/Ep/k5/Ep_unintegrated_UMAP_comparison_k5.png", p1 + p2, width = 12, height = 6, dpi = 300)

# RPCA reduction
p3 <- DimPlot(BCRF_Ep, reduction = "umap.rpca") +
  ggtitle("Original Clusters (RPCA)")
p4 <- DimPlot(BCRF_Ep, reduction = "umap.rpca", group.by = "clusters_k5") +
  ggtitle("NMF Clusters (k=5)")

ggsave("plots/NMF_reclustering/Ep/k5/Ep_RPCA_UMAP_comparison_k5.png", p3 + p4, width = 12, height = 6, dpi = 300)

# RPCA dxs comparison 
p5 <- DimPlot(BCRF_Ep, reduction = "umap.rpca", group.by = "dxs") +
  ggtitle("dxs (RPCA)") 

ggsave("plots/NMF_reclustering/Ep/k5/Ep_RPCA_dxs_UMAP_comparison_k5.png", p5 + p4, width = 12, height = 6, dpi = 300)
```

## Ep_corrected

### Reorient matrices correctly

```{r}
# Extract k=5 specific NMF result
Ep_res_corrected_k5 <- Ep_res_corrected$fit[["5"]]

# Extract basis and coefficient matrices
w_EpC_k5 <- basis(Ep_res_corrected_k5)
h_EpC_k5 <- coef(Ep_res_corrected_k5)

# Transpose coefficient matrix
EpC_NMF_embeddings_k5 <- t(h_EpC_k5)
```

### Check that cell names line up

rownames(nmf_embeddings) needs to exactly match colnames(seurat_obj), in the same order or matchable.

```{r}
# Edit column names 
colnames(EpC_NMF_embeddings_k5) <- paste0("NMF_", seq_len(ncol(EpC_NMF_embeddings_k5)))

# Put sample names as rownames of embeddings object 
rownames(EpC_NMF_embeddings_k5) <- colnames(Ep_vdata_corrected)  # whatever you fed into nmf()

# Change "." to "-" in NMF embedding sample names 
rownames(EpC_NMF_embeddings_k5) <- gsub("\\.", "-", rownames(EpC_NMF_embeddings_k5))

# sanity check before proceeding
all(rownames(EpC_NMF_embeddings_k5) %in% colnames(BCRF_Ep)) # TRUE
identical(rownames(EpC_NMF_embeddings_k5), colnames(BCRF_Ep)) # TRUE
identical(sort(rownames(EpC_NMF_embeddings_k5)), sort(colnames(BCRF_Ep))) # TRUE
```

### Create DimReducObject

```{r}
BCRF_Ep[["nmf_k5"]] <- CreateDimReducObject(
  embeddings = EpC_NMF_embeddings_k5,
  loadings   = w_EpC_k5,                 # optional but nice to keep (genes x k)
  key        = "NMF_",                   # must end in underscore, must be unique
  assay      = DefaultAssay(BCRF_Ep)
)
```

### Run/Generate UMAP on NMF reduction instead of PCA

```{r}
BCRF_Ep <- RunUMAP(
  BCRF_Ep,
  reduction = "nmf_k5",
  dims = 1:ncol(EpC_NMF_embeddings_k5), # or a subset of factors if some are junk
  reduction.name = "umap_nmf_k5_",
  reduction.key = "umapnmf_k5_"
)
```

```{r}
# Visualizations 
setwd("BCRFplots/NMF_reclustering/Ep_corrected/k5")

DimPlot(BCRF_Ep, reduction = "umap_nmf_k5_", label = TRUE)
ggsave("umap.nmf.k5.png", width = 8, height = 8)

DimPlot(BCRF_Ep, reduction = "umap_nmf_k5_", group.by = "protocol")
ggsave("protocol.png", width = 8, height = 8)

DimPlot(BCRF_Ep, reduction = "umap_nmf_k5_", group.by = "dxs")
ggsave("dxs.png", width = 8, height = 8)  

DimPlot(BCRF_Ep, reduction = "umap_nmf_k5_", group.by = "dx")
ggsave("dx.png", width = 12, height = 8)

n_patients <- length(unique(BCRF_Ep$patient))
DimPlot(BCRF_Ep, reduction = "umap_nmf_k5_", group.by = "patient") +
  ggtitle(paste("n patients = ", n_patients))
ggsave("patient.png", width = 12, height = 8)  

DimPlot(BCRF_Ep, reduction = "umap_nmf_k5_", group.by = "sequence.batch.number")
ggsave("sequence batch number.png", width = 12, height = 8)
```

### Map NMF result sampleIDs onto BCRF_Ep Seurat object 

```{r}
# Extract sample and cluster info from NMF results where k=5 
Ep_sample_clusters_corrected_k5 <- Ep_sample_clusters_corrected |> 
  select(Sample, Cluster, Rank) |>
  filter(Rank == 5)

extracted <- str_match(Ep_sample_clusters_corrected_k5$Sample, "set\\d+_\\d+_(\\d+)")[,2]
Ep_sample_clusters_corrected_k5$sample <- extracted

# force both sides to the same type before matching -- this is the #1 source of silent NAs
Ep_sample_clusters_corrected_k5$sample <- as.character(Ep_sample_clusters_corrected_k5$sample)
BCRF_Ep$sample <- as.character(BCRF_Ep$sample)

# map onto the Seurat object
BCRF_Ep$clusters_k5_corrected <- Ep_sample_clusters_corrected_k5$Cluster[
  match(BCRF_Ep$sample, Ep_sample_clusters_corrected_k5$sample)
]
BCRF_Ep$clusters_k5_corrected <- factor(BCRF_Ep$clusters_k5_corrected)

# sanity check -- how many cells got NA (i.e. sample wasn't found in your NMF table)?
table(BCRF_Ep$clusters_k5_corrected, useNA = "ifany")

# 1   2   3   4   5 
# 120  49 123 164  66 
# Matches silhouette plot cluster numbers 
```

### Plot on existing reductions 

```{r}
# NMF-Integrated reduction 
p1 <- DimPlot(BCRF_Ep, reduction = "umap_nmf_k5_") +
  ggtitle("Original Clusters (Integrated)")
p2 <- DimPlot(BCRF_Ep, reduction = "umap_nmf_k5_", group.by = "clusters_k5_corrected") +
  ggtitle("NMF Clusters (k=5)")

ggsave("plots/NMF_reclustering/Ep_corrected/k5/Ep_integrated_UMAP_comparison_k5.png", p1 + p2, width = 12, height = 6, dpi = 300)

# NMF-Integrated dxs comparison 
p3 <- DimPlot(BCRF_Ep, reduction = "umap_nmf_k5_", group.by = "dxs") +
  ggtitle("dxs (Integrated)") 

ggsave("plots/NMF_reclustering/Ep_corrected/k5/Ep_integrated_dxs_UMAP_comparison_k5.png", p3 + p2, width = 12, height = 6, dpi = 300)

# RPCA reduction
p4 <- DimPlot(BCRF_Ep, reduction = "umap.rpca") +
  ggtitle("Original Clusters (RPCA)")
p5 <- DimPlot(BCRF_Ep, reduction = "umap.rpca", group.by = "clusters_k5_corrected") +
  ggtitle("NMF Clusters (k=5, corrected)")

ggsave("plots/NMF_reclustering/Ep_corrected/k5/Ep_corrected_RPCA_UMAP_comparison_k5.png", p4 + p5, width = 12, height = 6, dpi = 300)

# RPCA dxs comparison 
p6 <- DimPlot(BCRF_Ep, reduction = "umap.rpca", group.by = "dxs") +
  ggtitle("dxs (RPCA)") 

ggsave("plots/NMF_reclustering/Ep_corrected/k5/Ep_corrected_RPCA_dxs_UMAP_comparison_k5.png", p6 + p5, width = 12, height = 6, dpi = 300)

```

## Str (uncorrected), k=5

### Map NMF result sampleIDs onto BCRF_Str Seurat object 

```{r}
# Extract sample and cluster info from NMF results where k=5 
Str_sample_clusters_k5 <- Str_sample_clusters |> 
  select(Sample, Cluster, Rank) |>
  filter(Rank == 5)

# Fix mismatch between sample names 
Str_sample_clusters_k5$Sample_clean <- gsub("\\.", "-", Str_sample_clusters_k5$Sample)

# Subset BCRF_Str by NMF samples only (remove sample w/ 0 counts)
BCRF_Str <- subset(BCRF_Str, cells = Str_sample_clusters_k5$Sample_clean)

extracted <- str_match(Str_sample_clusters_k5$Sample_clean, "set\\d+_\\d+_(\\d+)")[,2]
Str_sample_clusters_k5$sample <- extracted

# force both sides to the same type before matching -- this is the #1 source of silent NAs
Str_sample_clusters_k5$sample <- as.character(Str_sample_clusters_k5$sample)
BCRF_Str$sample <- as.character(BCRF_Str$sample)

# map onto the Seurat object
BCRF_Str$clusters_k5 <- Str_sample_clusters_k5$Cluster[
  match(BCRF_Str$sample, Str_sample_clusters_k5$sample)
]
BCRF_Str$clusters_k5 <- factor(BCRF_Str$clusters_k5)

# sanity check -- how many cells got NA (i.e. sample wasn't found in your NMF table)?
table(BCRF_Str$clusters_k5, useNA = "ifany")

# 1   2   3   4   5 
# 49  98  95  66 140 
# Matches silhouette plot cluster numbers 
```

### Plot on existing reductions 

```{r}
# Unintegrated reduction
p1 <- DimPlot(BCRF_Str, reduction = "umap.unintegrated") +
  ggtitle("Original Clusters (unintegrated)")
p2 <- DimPlot(BCRF_Str, reduction = "umap.unintegrated", group.by = "clusters_k5") +
  ggtitle("NMF Clusters (k=5)")

ggsave("plots/NMF_reclustering/Str/k5/Str_unintegrated_UMAP_comparison_k5.png", p1 + p2, width = 12, height = 6, dpi = 300)

# RPCA reduction
p3 <- DimPlot(BCRF_Str, reduction = "umap.rpca") +
  ggtitle("Original Clusters (RPCA)")
p4 <- DimPlot(BCRF_Str, reduction = "umap.rpca", group.by = "clusters_k5") +
  ggtitle("NMF Clusters (k=5)")

ggsave("plots/NMF_reclustering/Str/k5/Str_RPCA_UMAP_comparison_k5.png", p3 + p4, width = 12, height = 6, dpi = 300)

# RPCA dxs comparison 
p5 <- DimPlot(BCRF_Str, reduction = "umap.rpca", group.by = "dxs") +
  ggtitle("dxs (RPCA)") 

ggsave("plots/NMF_reclustering/Str/k5/Str_RPCA_dxs_UMAP_comparison_k5.png", p5 + p4, width = 12, height = 6, dpi = 300)
```
## Str_corrected

### Reorient matrices correctly

```{r}
# Extract k=5 specific NMF result
Str_res_corrected_k5 <- Str_res_corrected$fit[["5"]]

# Extract basis and coefficient matrices
w_StrC_k5 <- basis(Str_res_corrected_k5)
h_StrC_k5 <- coef(Str_res_corrected_k5)

# Transpose coefficient matrix
StrC_NMF_embeddings_k5 <- t(h_StrC_k5)
```

### Check that cell names line up

```{r}
# Edit column names 
colnames(StrC_NMF_embeddings_k5) <- paste0("NMF_", seq_len(ncol(StrC_NMF_embeddings_k5)))

# Put sample names as rownames of embeddings object 
rownames(StrC_NMF_embeddings_k5) <- colnames(Str_vdata_corrected)  # whatever you fed into nmf()

# Change "." to "-" in NMF embedding sample names 
rownames(StrC_NMF_embeddings_k5) <- gsub("\\.", "-", rownames(StrC_NMF_embeddings_k5))

# Subset BCRF_Str by NMF samples only 
BCRF_Str <- subset(BCRF_Str, cells = rownames(StrC_NMF_embeddings_k5))

# then re-check
identical(rownames(StrC_NMF_embeddings_k5), colnames(BCRF_Str)) # TRUE 
```

### Create DimReducObject

```{r}
BCRF_Str[["nmf_k5"]] <- CreateDimReducObject(
  embeddings = StrC_NMF_embeddings_k5,
  loadings   = w_StrC_k5,                # optional but nice to keep (genes x k)
  key        = "NMF_",                   # must end in underscore, must be unique
  assay      = DefaultAssay(BCRF_Str)
)
```

### Run/Generate UMAP on NMF reduction instead of PCA

```{r}
BCRF_Str <- RunUMAP(
  BCRF_Str,
  reduction = "nmf_k5",
  dims = 1:ncol(StrC_NMF_embeddings_k5), # or a subset of factors if some are junk
  reduction.name = "umap_nmf_k5_",
  reduction.key = "umapnmf_k5_"
)
```

```{r}
# Visualizations 
setwd("plots/NMF_reclustering/Str_corrected/k5")

DimPlot(BCRF_StrC_matched, reduction = "umap_nmf_k5_", label = TRUE)
ggsave("umap.nmf.k5.png", width = 8, height = 8)

DimPlot(BCRF_StrC_matched, reduction = "umap_nmf_k5_", group.by = "protocol")
ggsave("protocol.png", width = 8, height = 8)

DimPlot(BCRF_StrC_matched, reduction = "umap_nmf_k5_", group.by = "dxs")
ggsave("dxs.png", width = 8, height = 8)  

DimPlot(BCRF_StrC_matched, reduction = "umap_nmf_k5_", group.by = "dx")
ggsave("dx.png", width = 12, height = 8)

n_patients <- length(unique(BCRF_StrC_matched$patient))
DimPlot(BCRF_StrC_matched, reduction = "umap_nmf_k5_", group.by = "patient") +
  ggtitle(paste("n patients = ", n_patients))
ggsave("patient.png", width = 12, height = 8)  

DimPlot(BCRF_StrC_matched, reduction = "umap_nmf_k5_", group.by = "sequence.batch.number")
ggsave("sequence batch number.png", width = 12, height = 8)
```

### Map NMF result sampleIDs onto BCRF_Str Seurat object 

```{r}
# Extract sample and cluster info from NMF results where k=5 
Str_sample_clusters_corrected_k5 <- Str_sample_clusters_corrected |> 
  select(Sample, Cluster, Rank) |>
  filter(Rank == 5)

# Fix mismatch between sample names 
Str_sample_clusters_corrected_k5$Sample_clean <- gsub("\\.", "-", Str_sample_clusters_corrected_k5$Sample)

# Subset BCRF_Str by NMF samples only (remove sample w/ 0 counts)
BCRF_Str <- subset(BCRF_Str, cells = Str_sample_clusters_corrected_k5$Sample_clean)

extracted <- str_match(Str_sample_clusters_corrected_k5$Sample_clean, "set\\d+_\\d+_(\\d+)")[,2]
Str_sample_clusters_corrected_k5$sample <- extracted

# force both sides to the same type before matching -- this is the #1 source of silent NAs
Str_sample_clusters_corrected_k5$sample <- as.character(Str_sample_clusters_corrected_k5$sample)
BCRF_Str$sample <- as.character(BCRF_Str$sample)

# map onto the Seurat object
BCRF_Str$clusters_k5_corrected <- Str_sample_clusters_corrected_k5$Cluster[
  match(BCRF_Str$sample, Str_sample_clusters_corrected_k5$sample)
]
BCRF_Str$clusters_k5_corrected <- factor(BCRF_Str$clusters_k5_corrected)

# sanity check -- how many cells got NA (i.e. sample wasn't found in your NMF table)?
table(BCRF_Str$clusters_k5_corrected, useNA = "ifany")

# 1   2   3   4   5 
# 110  62  97  39 140 
# Matches silhouette plot cluster numbers 
```

### Plot on existing reductions 

```{r}
# Integrated reduction
p1 <- DimPlot(BCRF_Str, reduction = "umap_nmf_k5_") +
  ggtitle("Original Clusters (Integrated)")
p2 <- DimPlot(BCRF_Str, reduction = "umap_nmf_k5_", group.by = "clusters_k5_corrected") +
  ggtitle("NMF Clusters (k=5)")

ggsave("plots/NMF_reclustering/Str_corrected/k5/Str_Integrated_UMAP_comparison_k5.png", p1 + p2, width = 12, height = 6, dpi = 300)

# Integrated dxs comparison 
p3 <- DimPlot(BCRF_Str, reduction = "umap_nmf_k5_", group.by = "dxs") +
  ggtitle("dxs (Integrated)") 

ggsave("plots/NMF_reclustering/Str_corrected/k5/Str_Integrated_dxs_UMAP_comparison_k5.png", p3 + p2, width = 12, height = 6, dpi = 300)
```

# k=6

## Ep (uncorrected)

### Map NMF result sampleIDs onto BCRF_Ep Seurat object 

```{r}
# Extract sample and cluster info from NMF results where k=6 
Ep_sample_clusters_k6 <- Ep_sample_clusters |> 
  select(Sample, Cluster, Rank) |>
  filter(Rank == 6)

extracted <- str_match(Ep_sample_clusters_k6$Sample, "set\\d+_\\d+_(\\d+)")[,2]
Ep_sample_clusters_k6$sample <- extracted

# force both sides to the same type before matching -- this is the #1 source of silent NAs
Ep_sample_clusters_k6$sample <- as.character(Ep_sample_clusters_k6$sample)
BCRF_Ep$sample <- as.character(BCRF_Ep$sample)

# map onto the Seurat object
BCRF_Ep$clusters_k6 <- Ep_sample_clusters_k6$Cluster[
  match(BCRF_Ep$sample, Ep_sample_clusters_k6$sample)
]
BCRF_Ep$clusters_k6 <- factor(BCRF_Ep$clusters_k6)

# sanity check -- how many cells got NA (i.e. sample wasn't found in your NMF table)?
table(BCRF_Ep$clusters_k6, useNA = "ifany")

# 1   2   3   4   5   6 
# 74 106  63 118 108  53 
# Matches silhouette plot cluster numbers 
```

### Plot on existing reductions 

```{r}
# Unintegrated reduction
p1 <- DimPlot(BCRF_Ep, reduction = "umap.unintegrated") +
  ggtitle("Original Clusters (unintegrated)")
p2 <- DimPlot(BCRF_Ep, reduction = "umap.unintegrated", group.by = "clusters_k6") +
  ggtitle("NMF Clusters (k=6)")

ggsave("plots/NMF_reclustering/Ep/k6/Ep_unintegrated_UMAP_comparison_k6.png", p1 + p2, width = 12, height = 6, dpi = 300)

# RPCA reduction
p3 <- DimPlot(BCRF_Ep, reduction = "umap.rpca") +
  ggtitle("Original Clusters (RPCA)")
p4 <- DimPlot(BCRF_Ep, reduction = "umap.rpca", group.by = "clusters_k6") +
  ggtitle("NMF Clusters (k=6)")

ggsave("plots/NMF_reclustering/Ep/k6/Ep_RPCA_UMAP_comparison_k6.png", p3 + p4, width = 12, height = 6, dpi = 300)

# RPCA dxs comparison 
p5 <- DimPlot(BCRF_Ep, reduction = "umap.rpca", group.by = "dxs") +
  ggtitle("dxs (RPCA)") 

ggsave("plots/NMF_reclustering/Ep/k6/Ep_RPCA_dxs_UMAP_comparison_k6.png", p5 + p4, width = 12, height = 6, dpi = 300)
```

## Ep_corrected

### Reorient matrices correctly

```{r}
# Extract k=6 specific NMF result
Ep_res_corrected_k6 <- Ep_res_corrected$fit[["6"]]

# Extract basis and coefficient matrices
w_EpC_k6 <- basis(Ep_res_corrected_k6)
h_EpC_k6 <- coef(Ep_res_corrected_k6)

# Transpose coefficient matrix
EpC_NMF_embeddings_k6 <- t(h_EpC_k6)
```

### Check that cell names line up

```{r}
# Edit column names 
colnames(EpC_NMF_embeddings_k6) <- paste0("NMF_", seq_len(ncol(EpC_NMF_embeddings_k6)))

# Put sample names as rownames of embeddings object 
rownames(EpC_NMF_embeddings_k6) <- colnames(Ep_vdata_corrected)  # whatever you fed into nmf()

# Change "." to "-" in NMF embedding sample names 
rownames(EpC_NMF_embeddings_k6) <- gsub("\\.", "-", rownames(EpC_NMF_embeddings_k6))

# sanity check before proceeding
all(rownames(EpC_NMF_embeddings_k6) %in% colnames(BCRF_Ep)) # TRUE
identical(rownames(EpC_NMF_embeddings_k6), colnames(BCRF_Ep)) # TRUE
identical(sort(rownames(EpC_NMF_embeddings_k6)), sort(colnames(BCRF_Ep))) # TRUE
```

### Create DimReducObject

```{r}
BCRF_Ep[["nmf_k6"]] <- CreateDimReducObject(
  embeddings = EpC_NMF_embeddings_k6,
  loadings   = w_EpC_k6,                 # optional but nice to keep (genes x k)
  key        = "NMF_",                   # must end in underscore, must be unique
  assay      = DefaultAssay(BCRF_Ep)
)
```

### Run/Generate UMAP on NMF reduction instead of PCA

```{r}
BCRF_Ep <- RunUMAP(
  BCRF_Ep,
  reduction = "nmf_k6",
  dims = 1:ncol(EpC_NMF_embeddings_k6), # or a subset of factors if some are junk
  reduction.name = "umap_nmf_k6_",
  reduction.key = "umapnmf_k6_"
)
```

```{r}
# Visualizations 
setwd("plots/NMF_reclustering/Ep_corrected/k6")

DimPlot(BCRF_Ep, reduction = "umap_nmf_k6_", label = TRUE)
ggsave("umap.nmf.k6.png", width = 8, height = 8)

DimPlot(BCRF_Ep, reduction = "umap_nmf_k6_", group.by = "protocol")
ggsave("protocol.png", width = 8, height = 8)

DimPlot(BCRF_Ep, reduction = "umap_nmf_k6_", group.by = "dxs")
ggsave("dxs.png", width = 8, height = 8)  

DimPlot(BCRF_Ep, reduction = "umap_nmf_k6_", group.by = "dx")
ggsave("dx.png", width = 12, height = 8)

n_patients <- length(unique(BCRF_Ep$patient))
DimPlot(BCRF_Ep, reduction = "umap_nmf_k6_", group.by = "patient") +
  ggtitle(paste("n patients = ", n_patients))
ggsave("patient.png", width = 12, height = 8)  

DimPlot(BCRF_Ep, reduction = "umap_nmf_k6_", group.by = "sequence.batch.number")
ggsave("sequence batch number.png", width = 12, height = 8)
```

### Map NMF result sampleIDs onto BCRF_Ep Seurat object 

```{r}
# Extract sample and cluster info from NMF results where k=6 
Ep_sample_clusters_corrected_k6 <- Ep_sample_clusters_corrected |> 
  select(Sample, Cluster, Rank) |>
  filter(Rank == 6)

extracted <- str_match(Ep_sample_clusters_corrected_k6$Sample, "set\\d+_\\d+_(\\d+)")[,2]
Ep_sample_clusters_corrected_k6$sample <- extracted

# force both sides to the same type before matching -- this is the #1 source of silent NAs
Ep_sample_clusters_corrected_k6$sample <- as.character(Ep_sample_clusters_corrected_k6$sample)
BCRF_Ep$sample <- as.character(BCRF_Ep$sample)

# map onto the Seurat object
BCRF_Ep$clusters_k6_corrected <- Ep_sample_clusters_corrected_k6$Cluster[
  match(BCRF_Ep$sample, Ep_sample_clusters_corrected_k6$sample)
]
BCRF_Ep$clusters_k6_corrected <- factor(BCRF_Ep$clusters_k6_corrected)

# sanity check -- how many cells got NA (i.e. sample wasn't found in your NMF table)?
table(BCRF_Ep$clusters_k6_corrected, useNA = "ifany")

# 1   2   3   4   5   6 
# 33  63  50 116 116 144 
# Matches silhouette plot cluster numbers 
```

### Plot on existing reductions 

```{r}
# Integrated reduction 
p1 <- DimPlot(BCRF_Ep, reduction = "umap_nmf_k6_") +
  ggtitle("Original Clusters (Integrated)")
p2 <- DimPlot(BCRF_Ep, reduction = "umap_nmf_k6_", group.by = "clusters_k6_corrected") +
  ggtitle("NMF Clusters (k=6)")

ggsave("plots/NMF_reclustering/Ep_corrected/k6/Ep_integrated_UMAP_comparison_k6.png", p1 + p2, width = 12, height = 6, dpi = 300)

# RPCA dxs comparison 
p3 <- DimPlot(BCRF_Ep, reduction = "umap_nmf_k6_", group.by = "dxs") +
  ggtitle("dxs (Integrated)") 

ggsave("plots/NMF_reclustering/Ep_corrected/k6/Ep_integrated_dxs_UMAP_comparison_k6.png", p3 + p2, width = 12, height = 6, dpi = 300)
```

## Str (uncorrected)

### Map NMF result sampleIDs onto BCRF_Str Seurat object 

```{r}
# Extract sample and cluster info from NMF results where k=6
Str_sample_clusters_k6 <- Str_sample_clusters |> 
  select(Sample, Cluster, Rank) |>
  filter(Rank == 6)

# Fix mismatch between sample names 
Str_sample_clusters_k6$Sample_clean <- gsub("\\.", "-", Str_sample_clusters_k6$Sample)

# Subset BCRF_Str by NMF samples only (remove sample w/ 0 counts)
BCRF_Str <- subset(BCRF_Str, cells = Str_sample_clusters_k6$Sample_clean)

extracted <- str_match(Str_sample_clusters_k6$Sample_clean, "set\\d+_\\d+_(\\d+)")[,2]
Str_sample_clusters_k6$sample <- extracted

# force both sides to the same type before matching -- this is the #1 source of silent NAs
Str_sample_clusters_k6$sample <- as.character(Str_sample_clusters_k6$sample)
BCRF_Str$sample <- as.character(BCRF_Str$sample)

# map onto the Seurat object
BCRF_Str$clusters_k6 <- Str_sample_clusters_k6$Cluster[
  match(BCRF_Str$sample, Str_sample_clusters_k6$sample)
]
BCRF_Str$clusters_k6 <- factor(BCRF_Str$clusters_k6)

# sanity check -- how many cells got NA (i.e. sample wasn't found in your NMF table)?
table(BCRF_Str$clusters_k6, useNA = "ifany")

# 1   2   3   4   5   6 
# 75  95  86  33  39 120 
# Matches silhouette plot cluster numbers 
```

### Plot on existing reductions 

```{r}
# Unintegrated reduction
p1 <- DimPlot(BCRF_Str, reduction = "umap.unintegrated") +
  ggtitle("Original Clusters (unintegrated)")
p2 <- DimPlot(BCRF_Str, reduction = "umap.unintegrated", group.by = "clusters_k6") +
  ggtitle("NMF Clusters (k=6)")

ggsave("plots/NMF_reclustering/Str/k6/Str_unintegrated_UMAP_comparison_k6.png", p1 + p2, width = 12, height = 6, dpi = 300)

# RPCA reduction
p3 <- DimPlot(BCRF_Str, reduction = "umap.rpca") +
  ggtitle("Original Clusters (RPCA)")
p4 <- DimPlot(BCRF_Str, reduction = "umap.rpca", group.by = "clusters_k6") +
  ggtitle("NMF Clusters (k=6)")

ggsave("plots/NMF_reclustering/Str/k6/Str_RPCA_UMAP_comparison_k6.png", p3 + p4, width = 12, height = 6, dpi = 300)

# RPCA dxs comparison 
p5 <- DimPlot(BCRF_Str, reduction = "umap.rpca", group.by = "dxs") +
  ggtitle("dxs (RPCA)") 

ggsave("plots/NMF_reclustering/Str/k6/Str_RPCA_dxs_UMAP_comparison_k6.png", p5 + p4, width = 12, height = 6, dpi = 300)
```

## Str_corrected

### Reorient matrices correctly

```{r}
# Extract k=6 specific NMF result
Str_res_corrected_k6 <- Str_res_corrected$fit[["6"]]

# Extract basis and coefficient matrices
w_StrC_k6 <- basis(Str_res_corrected_k6)
h_StrC_k6 <- coef(Str_res_corrected_k6)

# Transpose coefficient matrix
StrC_NMF_embeddings_k6 <- t(h_StrC_k6)
```

### Check that cell names line up

```{r}
# Edit column names 
colnames(StrC_NMF_embeddings_k6) <- paste0("NMF_", seq_len(ncol(StrC_NMF_embeddings_k6)))

# Put sample names as rownames of embeddings object 
rownames(StrC_NMF_embeddings_k6) <- colnames(Str_vdata_corrected)  # whatever you fed into nmf()

# Change "." to "-" in NMF embedding sample names 
rownames(StrC_NMF_embeddings_k6) <- gsub("\\.", "-", rownames(StrC_NMF_embeddings_k6))

# Subset BCRF_Str by NMF samples only 
BCRF_Str <- subset(BCRF_Str, cells = rownames(StrC_NMF_embeddings_k6))

# then re-check
identical(rownames(StrC_NMF_embeddings_k6), colnames(BCRF_Str)) # TRUE 
```

### Create DimReducObject

```{r}
BCRF_Str[["nmf_k6"]] <- CreateDimReducObject(
  embeddings = StrC_NMF_embeddings_k6,
  loadings   = w_StrC_k6,                # optional but nice to keep (genes x k)
  key        = "NMF_",                   # must end in underscore, must be unique
  assay      = DefaultAssay(BCRF_Str)
)
```

### Run/Generate UMAP on NMF reduction instead of PCA

```{r}
BCRF_Str <- RunUMAP(
  BCRF_Str,
  reduction = "nmf_k6",
  dims = 1:ncol(StrC_NMF_embeddings_k6), # or a subset of factors if some are junk
  reduction.name = "umap_nmf_k6_",
  reduction.key = "umapnmf_k6_"
)
```

```{r}
# Visualizations 
setwd("plots/NMF_reclustering/Str_corrected/k6")

DimPlot(BCRF_Str, reduction = "umap_nmf_k6_", label = TRUE)
ggsave("umap.nmf.k5.png", width = 8, height = 8)

DimPlot(BCRF_Str, reduction = "umap_nmf_k6_", group.by = "protocol")
ggsave("protocol.png", width = 8, height = 8)

DimPlot(BCRF_Str, reduction = "umap_nmf_k6_", group.by = "dxs")
ggsave("dxs.png", width = 8, height = 8)  

DimPlot(BCRF_Str, reduction = "umap_nmf_k6_", group.by = "dx")
ggsave("dx.png", width = 12, height = 8)

n_patients <- length(unique(BCRF_Str$patient))
DimPlot(BCRF_Str, reduction = "umap_nmf_k6_", group.by = "patient") +
  ggtitle(paste("n patients = ", n_patients))
ggsave("patient.png", width = 12, height = 8)  

DimPlot(BCRF_Str, reduction = "umap_nmf_k6_", group.by = "sequence.batch.number")
ggsave("sequence batch number.png", width = 12, height = 8)
```

### Map NMF result sampleIDs onto BCRF_Str Seurat object 

```{r}
# Extract sample and cluster info from NMF results where k=6 
Str_sample_clusters_corrected_k6 <- Str_sample_clusters_corrected |> 
  select(Sample, Cluster, Rank) |>
  filter(Rank == 6)

# Fix mismatch between sample names 
Str_sample_clusters_corrected_k6$Sample_clean <- gsub("\\.", "-", Str_sample_clusters_corrected_k6$Sample)

# Subset BCRF_Str by NMF samples only (remove sample w/ 0 counts)
BCRF_Str <- subset(BCRF_Str, cells = Str_sample_clusters_corrected_k6$Sample_clean)

extracted <- str_match(Str_sample_clusters_corrected_k6$Sample_clean, "set\\d+_\\d+_(\\d+)")[,2]
Str_sample_clusters_corrected_k6$sample <- extracted

# force both sides to the same type before matching -- this is the #1 source of silent NAs
Str_sample_clusters_corrected_k6$sample <- as.character(Str_sample_clusters_corrected_k6$sample)
BCRF_Str$sample <- as.character(BCRF_Str$sample)

# map onto the Seurat object
BCRF_Str$clusters_k6_corrected <- Str_sample_clusters_corrected_k6$Cluster[
  match(BCRF_Str$sample, Str_sample_clusters_corrected_k6$sample)
]
BCRF_Str$clusters_k6_corrected <- factor(BCRF_Str$clusters_k6_corrected)

# sanity check -- how many cells got NA (i.e. sample wasn't found in your NMF table)?
table(BCRF_Str$clusters_k6_corrected, useNA = "ifany")

#  1   2   3   4   5   6 
# 90 109 105  55  64  25 
# Matches silhouette plot cluster numbers 
```

### Plot on existing reductions 

```{r}
# Integrated reduction
p1 <- DimPlot(BCRF_Str, reduction = "umap_nmf_k6_") +
  ggtitle("Original Clusters (Integrated)")
p2 <- DimPlot(BCRF_Str, reduction = "umap_nmf_k6_", group.by = "clusters_k6_corrected") +
  ggtitle("NMF Clusters (k=6)")

ggsave("plots/NMF_reclustering/Str_corrected/k6/Str_Integrated_UMAP_comparison_k6.png", p1 + p2, width = 12, height = 6, dpi = 300)

# Integrated dxs comparison 
p3 <- DimPlot(BCRF_Str, reduction = "umap_nmf_k6_", group.by = "dxs") +
  ggtitle("dxs (Integrated)") 

ggsave("plots/NMF_reclustering/Str_corrected/k6/Str_Integrated_dxs_UMAP_comparison_k6.png", p3 + p2, width = 12, height = 6, dpi = 300)
```



# Save RDS of NMF cluster metadata and reductions 

```{r}
# Save Ep object w/ k=5,6 metadata (and also reductions, for corrected)
saveRDS(BCRF_Ep, "data/integrated_BCRF_Ep_inc_NMF.rds")

# Save Str object w/ k=4,5,6 metadata (and also reductions, for corrected)
saveRDS(BCRF_Str, "data/integrated_BCRF_Str_inc_NMF.rds")
```

