---
title: "07_CIBERSORTx"
format: gfm
---

# 07_CIBERSORTx

The goal of this script is to perform bulk RNA-Seq deconvolution via CIBERSORTx to analyze what fraction of each bulk sample is composed of what cell types. 

## Set-up

```{r}
# Load libraries 
library(Seurat)  # For constructing raw reference and input (mixture) files 
library(SeuratDisk)
library(ggplot2) # For analyzing results files 
library(tidyr)
library(dplyr)

# To install SeuratDisk to read Kumar data 
# install.packages("remotes")  # if not already installed
# remotes::install_github("mojaveazure/seurat-disk")

# Set global options 
options(future.globals.maxSize = Inf)
print(getOption("future.globals.maxSize"))
options(Seurat.object.assay.version = "v5")

# Load in Seurat objects to create mixture data
BCRF_Ep <- readRDS("data/raw_data/RE-DONE integrate BCRF v1_Ep 1.rds")
BCRF_Str <- readRDS("data/raw_data/RE-DONE integrated BCRF v1_Str 3.rds")
```

## (Bassez) Construct raw reference matrix 

The raw reference matrix generated by this script will be uploaded as "Single Cell Reference" onto CIBERSORTx web portal, and the signature reference matrix will be generated by the custom Step 1: "Create Signature Matrix" step. Since the Bassez data is from scRNAseq, a source GEP file will also be  generated to be used with B mode batch correction in Step 2: "Impute Cell Fractions". 

```{r}
# Load in data 
counts <- readRDS("data/CIBERSORTx/reference/Bassez_counts_cohort1.rds")
meta <- read.csv("data/CIBERSORTx/reference/Bassez_metadata_cohort1.csv", row.names = 1)

# Create Seurat object 
Bassez_seurat <- CreateSeuratObject(counts = counts, meta.data = meta)
saveRDS(Bassez_seurat, "data/CIBERSORTx/reference/Bassez_seurat.rds")

Bassez_seurat <- readRDS("data/CIBERSORTx/reference/Bassez_seurat.rds")
```

The signature matrix should be constructed as a .tsv file where rows are genes and columns are individual cells, labeled by their cell type name (column header).

```{r}
# Pull raw counts and cell type labels 
# raw counts - NOT normalized, CIBERSORTx does its own internal normalization
counts <- GetAssayData(Bassez_seurat, assay = "RNA", layer = "counts")
counts <- as.matrix(counts)

# pull cell type labels - adjust column name to whatever you found in the metadata
celltypes <- Bassez_seurat$cellType
```

### Downsample cells per type 

Account for unbalanced cell type group sizes by adjusting to a consistent number per cell type. 

```{r}
set.seed(67)
max_cells_per_type <- 700  # adjust based on your smallest group size

cells_to_keep <- unlist(lapply(unique(celltypes), function(ct) {
  ct_cells <- names(celltypes)[celltypes == ct]
  if (length(ct_cells) > max_cells_per_type) {
    sample(ct_cells, max_cells_per_type)
  } else {
    ct_cells
  }
}))

counts_sub <- counts[, cells_to_keep]
celltypes_sub <- celltypes[cells_to_keep]
```

### Build the CIBERSORTx-format matrix (cell-type labels as column headers)

```{r}
# replace column names with cell type labels
colnames(counts_sub) <- as.character(celltypes_sub)

# add gene symbols as first column
sig_matrix_df <- data.frame(GeneSymbol = rownames(counts_sub), counts_sub, check.names = FALSE)
```

### Clean up gene symbols 

```{r}
# remove duplicate gene symbols (keep first occurrence, or sum if you prefer)
sig_matrix_df <- sig_matrix_df[!duplicated(sig_matrix_df$GeneSymbol), ]

# drop genes with zero expression across all retained cells
gene_sums <- rowSums(sig_matrix_df[, -1])
sig_matrix_df <- sig_matrix_df[gene_sums > 0, ]

# Sanity check
table(colnames(sig_matrix_df)[-1]) # All celltypes only contain 700 
```

### Construct signature matrix file 

```{r}
write.table(
  sig_matrix_df,
  file = "data/CIBERSORTx/reference/Bassez_cohort1_raw_reference.txt",
  sep = "\t",
  quote = FALSE,
  row.names = FALSE
)
```

## (Kumar) Construct raw reference matrix 

The publically available Kumar scRNAseq dataset is in the form of an .h5ad file. The following script builds a Seurat object by converting and pulling information from this file type. 

```{r}
Convert("../data/CIBERSORTx/reference/Kumar_data.h5ad", dest = "h5seurat", overwrite = TRUE)

h5 <- H5File$new("../data/CIBERSORTx/reference/Kumar_data.h5seurat", mode = "r")
h5$ls(recursive = TRUE)  # inspect structure first — confirm paths below match what you see

# raw counts
counts_data    <- h5[["assays/RNA/counts/data"]]$read()
counts_indices <- h5[["assays/RNA/counts/indices"]]$read()
counts_indptr  <- h5[["assays/RNA/counts/indptr"]]$read()
cell.names <- h5[["cell.names"]]$read()

# Pull gene symbols NOT Ensembl IDs
feature_cats  <- h5[["assays/RNA/meta.features/feature_name/categories"]]$read()
feature_codes <- h5[["assays/RNA/meta.features/feature_name/codes"]]$read()
features <- feature_cats[feature_codes + 1]   # same +1 offset logic as read_meta_col

n_cells <- length(cell.names)
n_genes <- length(features)

counts_mat <- new("dgRMatrix",
                   j = as.integer(counts_indices),
                   p = as.integer(counts_indptr),
                   x = as.numeric(counts_data),
                   Dim = c(n_cells, n_genes))
counts_mat <- t(as(counts_mat, "CsparseMatrix"))
rownames(counts_mat) <- features
colnames(counts_mat) <- cell.names

# Check for duplicate genes 
sum(duplicated(rownames(counts_mat))) # 16 

# drop rows with empty/NA gene symbols first (common in feature_name fields)
keep <- !is.na(rownames(counts_mat)) & rownames(counts_mat) != "" 
counts_mat <- counts_mat[keep, ]

# collapse duplicates by summing counts across rows sharing the same symbol
# (safer than "keep first" here, since you're at the raw counts stage, not post-hoc)
counts_mat <- rowsum(as.matrix(counts_mat), group = rownames(counts_mat))
counts_mat <- as(counts_mat, "CsparseMatrix")  # convert back to sparse if needed for memory

dim(counts_mat) # 32367 714331
sum(duplicated(rownames(counts_mat)))  # now 0

# metadata — categorical columns stored as categories+codes, others plain
read_meta_col <- function(h5, path) {
  obj <- h5[[path]]
  if (inherits(obj, "H5Group")) {
    cats  <- obj[["categories"]]$read()
    codes <- obj[["codes"]]$read()
    factor(cats[codes + 1], levels = cats)
  } else {
    obj$read()
  }
}

meta_names <- names(h5[["meta.data"]])
meta.data <- as.data.frame(
  setNames(lapply(meta_names, function(nm) read_meta_col(h5, paste0("meta.data/", nm))), meta_names)
)
rownames(meta.data) <- cell.names

h5$close_all()

Kumar_seurat <- CreateSeuratObject(counts = counts_mat, meta.data = meta.data)
saveRDS(Kumar_seurat, "../data/CIBERSORTx/reference/Kumar_seurat.rds")
```

### Downsample cells by type 

Use Seurat's built-in downsampling before converting counts w/ as.matrix(). 

```{r}
Kumar_seurat <- readRDS("../data/CIBERSORTx/reference/Kumar_seurat.rds")
set.seed(67)  # for reproducibility

# Choose metadata column to downsample by
Idents(Kumar_seurat) <- "cell_type" 

# check group sizes first to pick sensible target n
table(Idents(Kumar_seurat))

# Downsample to 500 cells per cell type (adjusted based on smallest groups)
cells_to_keep <- WhichCells(Kumar_seurat, downsample = 500)
Kumar_seurat_sub <- subset(Kumar_seurat, cells = cells_to_keep)

table(Idents(Kumar_seurat_sub))  # Confirm downsampling 
```

### Build the CIBERSORTx-format matrix (cell-type labels as column headers)

```{r}
# Pull raw counts and cell type labels 
# raw counts - NOT normalized, CIBERSORTx does its own internal normalization
counts_sub <- GetAssayData(Kumar_seurat_sub, assay = "RNA", layer = "counts")
counts_sub <- as.matrix(counts_sub)

dim(counts_sub) # 32383 genes x 18630 cells 

# pull cell type labels - adjust column name to whatever you found in the metadata
celltypes_sub <- Kumar_seurat_sub$cell_type 

# replace column names with cell type labels
colnames(counts_sub) <- as.character(celltypes_sub)

# add gene symbols as first column
sig_matrix_df <- data.frame(GeneSymbol = rownames(counts_sub), counts_sub, check.names = FALSE)
```

### Clean up gene symbols 

```{r}
# remove duplicate gene symbols (keep first occurrence, or sum if you prefer)
sig_matrix_df <- sig_matrix_df[!duplicated(sig_matrix_df$GeneSymbol), ]

# drop genes with zero expression across all retained cells
gene_sums <- rowSums(sig_matrix_df[, -1])
sig_matrix_df <- sig_matrix_df[gene_sums > 0, ]

# Sanity check
table(colnames(sig_matrix_df)[-1]) # All celltypes only contain <=500 
```

### Construct signature matrix file 

```{r}
write.table(
  sig_matrix_df,
  file = "../data/CIBERSORTx/reference/Kumar_raw_reference.txt",
  sep = "\t",
  quote = FALSE,
  row.names = FALSE
)
```

## Construct input (mixture) data 

CIBERSORTx expects input counts data CPM or TPM, not raw counts, in the form of .tsv files. 

### Ep

```{r}
# Extract raw counts data from Seurat object 
Ep_raw_counts <- GetAssayData(BCRF_Ep, assay = "RNA", layer = "counts")

# Convert to regular matrix (GetAssayData() sometimes returns sparse)
Ep_raw_counts <- as.matrix(Ep_raw_counts)

# Convert to CPM (counts per million)
# Divide every column sample by total counts for each sample and convert to CPM 
Ep_cpm <- sweep(Ep_raw_counts, 2, colSums(Ep_raw_counts), FUN = "/") * 1e6 

# Construct dataframe w/ gene names as first column 
# CIBERSORTx requires first column name to be "GeneSymbol" 
Ep_cpm_df <- data.frame(
  GeneSymbol = rownames(Ep_cpm), 
  Ep_cpm,
  check.names = FALSE 
)

# Save as .tsv file 
write.table(Ep_cpm_df,
            "data/CIBERSORTx/Ep_cibersortx_input.tsv",
            sep       = "\t",
            row.names = FALSE,
            quote     = FALSE
            )
```

### Str

```{r}
# Extract raw counts data from Seurat object 
Str_raw_counts <- GetAssayData(BCRF_Str, assay = "RNA", layer = "counts")

# Convert to regular matrix (GetAssayData() sometimes returns sparse)
Str_raw_counts <- as.matrix(Str_raw_counts)

# Remove empty (zero counts) sample 
# Check totals per sample
sample_totals <- colSums(Str_raw_counts)
sample_totals[sample_totals == 0]  # confirm which sample(s) are zero

# Drop it
Str_raw_counts <- Str_raw_counts[, sample_totals > 0]

# Convert to CPM (counts per million)
# Divide every column sample by total counts for each sample and convert to CPM 
Str_cpm <- sweep(Str_raw_counts, 2, colSums(Str_raw_counts), FUN = "/") * 1e6 

# Construct dataframe w/ gene names as first column 
# CIBERSORTx requires first column name to be "GeneSymbol" 
Str_cpm_df <- data.frame(
  GeneSymbol = rownames(Str_cpm), 
  Str_cpm,
  check.names = FALSE 
)

# Save as .tsv file 
write.table(Str_cpm_df,
            "../data/CIBERSORTx/Str_cibersortx_input.tsv",
            sep       = "\t",
            row.names = FALSE,
            quote     = FALSE
            )
```

## Parse output (results) data

```{r}
library(ggplot2)
library(tidyr)
library(dplyr)
library(pals)
```

ALH will be excluded from plots because there is only one sample, so there is little value in comparative analysis. 

### Ep (Bassez)

```{r}
# Load CIBERSORTx output (downloaded from the website)
Ep_Bassez_results <- read.table(
  "data/CIBERSORTx/CIBERSORTx_Ep_Bassez_results.txt",
  header    = TRUE,
  sep       = "\t",
  row.names = 1,
  check.names = FALSE
)

# Keep only the cell fraction columns (remove p-value, correlation, RMSE)
Ep_Bassez_fraction_cols <- setdiff(
  colnames(Ep_Bassez_results),
  c("P-value", "Correlation", "RMSE")
)
Ep_Bassez_fractions <- Ep_Bassez_results[, Ep_Bassez_fraction_cols]

# Add condition metadata
Ep_Bassez_fractions$sample <- rownames(Ep_Bassez_fractions)
Ep_Bassez_fractions$dxs <- BCRF_Ep@meta.data[rownames(Ep_Bassez_fractions), "dxs"]

# Reshape for plotting
Ep_Bassez_fractions_long <- Ep_Bassez_fractions |>
  tidyr::pivot_longer(
    cols      = all_of(Ep_Bassez_fraction_cols),
    names_to  = "cell_type",
    values_to = "proportion"
  )

# Stacked bar chart of cell type proportions per sample
# All dxs 
Ep_Bassez_all_dxs <- Ep_Bassez_fractions_long |>
  filter(dxs %in% c("NL", "BBD", "BBD_RS", "ADH", "IN", "DCIS", "Invasive")) |>
  mutate(dxs = factor(
    dxs,levels = c("NL", "BBD", "BBD_RS", "ADH", "IN", "DCIS", "Invasive"))
    ) |>
  ggplot(aes(x = sample, y = proportion, fill = cell_type)) +
  geom_col() +
  scale_fill_brewer(palette = "Set3", name = "Cell type") +
  labs(
    x     = "Sample",
    y     = "Estimated proportion",
    title = "Cell type composition of epithelial samples (CIBERSORTx)"
  ) +
  theme_classic(base_size = 10) +
  theme(axis.text.x = element_blank()) +
  facet_grid(.~ dxs, scales = "free_x", space = "free_x")

print(Ep_Bassez_all_dxs)
ggsave("plots/CIBERSORTx/Ep_Bassez/Ep_Bassez_all_dxs.png", Ep_Bassez_all_dxs, 
       width = 8, height = 5, dpi = 300)

# NL, BBD, BBD_RS dxs 
Ep_Bassez_NL_BBD_BBDRS <- Ep_Bassez_fractions_long |>
  filter(dxs %in% c("NL", "BBD", "BBD_RS")) |>
  mutate(dxs = factor(dxs, levels = c("NL", "BBD", "BBD_RS"))) |>
  ggplot(aes(x = sample, y = proportion, fill = cell_type)) + 
    geom_col() +
    scale_fill_brewer(palette = "Set3", name = "Cell type") + 
    labs(
      x     = "Sample",
      y     = "Estimated proportion",
      title = "Cell type composition of epithelial samples (CIBERSORTx)"
    ) + 
    theme_classic() + 
    theme(axis.text.x = element_blank()) +
    facet_grid(.~ dxs, scales = "free_x", space = "free_x")
  
print(Ep_Bassez_NL_BBD_BBDRS)
ggsave("plots/CIBERSORTx/Ep_Bassez/Ep_Bassez_NL_BBD_BBDRS.png", Ep_Bassez_NL_BBD_BBDRS, 
       width = 7, height = 5, dpi = 300)

# ADH, IN, DCIS, Invasive dxs
Ep_Bassez_ADH_IN_DCIS_Inv <- Ep_Bassez_fractions_long |>
  filter(dxs %in% c("ADH", "IN", "DCIS", "Invasive")) |>
  mutate(dxs = factor(dxs, levels = c("ADH", "IN", "DCIS", "Invasive"))) |>
  ggplot(aes(x = sample, y = proportion, fill = cell_type)) + 
    geom_col() +
    scale_fill_brewer(palette = "Set3", name = "Cell type") + 
    labs(
      x     = "Sample",
      y     = "Estimated proportion",
      title = "Cell type composition of epithelial samples (CIBERSORTx)"
    ) + 
    theme_classic() + 
    theme(axis.text.x = element_blank()) +
    facet_grid(.~ dxs, scales = "free_x", space = "free_x")
  
print(Ep_Bassez_ADH_IN_DCIS_Inv)
ggsave("plots/CIBERSORTx/Ep_Bassez/Ep_Bassez_ADH_IN_DCIS_Inv.png", Ep_Bassez_ADH_IN_DCIS_Inv, 
       width = 7, height = 5, dpi = 300)
```

### Ep (Kumar)

```{r}
# Load CIBERSORTx output (downloaded from the website)
Ep_Kumar_results <- read.table(
  "data/CIBERSORTx/CIBERSORTx_Ep_Kumar_results.txt",
  header    = TRUE,
  sep       = "\t",
  row.names = 1,
  check.names = FALSE
)

# Keep only the cell fraction columns (remove p-value, correlation, RMSE)
Ep_Kumar_fraction_cols <- setdiff(
  colnames(Ep_Kumar_results),
  c("P-value", "Correlation", "RMSE")
)
Ep_Kumar_fractions <- Ep_Kumar_results[, Ep_Kumar_fraction_cols]

# Add condition metadata
Ep_Kumar_fractions$sample <- rownames(Ep_Kumar_fractions)
Ep_Kumar_fractions$dxs <- BCRF_Ep@meta.data[rownames(Ep_Kumar_fractions), "dxs"]

# Reshape for plotting
Ep_Kumar_fractions_long <- Ep_Kumar_fractions |>
  tidyr::pivot_longer(
    cols      = all_of(Ep_Kumar_fraction_cols),
    names_to  = "cell_type",
    values_to = "proportion"
  )

# Build color palette with 39 distinct colors for Kumar cell types
my_colors <-as.vector(colorRampPalette(glasbey())(39))

# Stacked bar chart of cell type proportions per sample
# All dxs 
Ep_Kumar_all_dxs_glasbey_colors <- Ep_Kumar_fractions_long |>
  filter(dxs %in% c("NL", "BBD", "BBD_RS", "ADH", "IN", "DCIS", "Invasive")) |>
  mutate(dxs = factor(
    dxs,levels = c("NL", "BBD", "BBD_RS", "ADH", "IN", "DCIS", "Invasive"))
    ) |>
  ggplot(aes(x = sample, y = proportion, fill = cell_type)) +
  geom_col() +
  labs(
    x     = "Sample",
    y     = "Estimated proportion",
    title = "Cell type composition of epithelial samples (CIBERSORTx)"
  ) +
  scale_fill_manual(values = my_colors) + 
  theme_classic(base_size = 10) +
  theme(axis.text.x = element_blank()) +
  facet_grid(.~ dxs, scales = "free_x", space = "free_x")

print(Ep_Kumar_all_dxs_glasbey_colors)
ggsave("plots/CIBERSORTx/Ep_Kumar/Ep_Kumar_all_dxs_glasbey_colors.png", Ep_Kumar_all_dxs_glasbey_colors, 
       width = 12 , height = 6, dpi = 300)

Ep_Kumar_all_dxs <- Ep_Kumar_fractions_long |>
  filter(dxs %in% c("NL", "BBD", "BBD_RS", "ADH", "IN", "DCIS", "Invasive")) |>
  mutate(dxs = factor(
    dxs,levels = c("NL", "BBD", "BBD_RS", "ADH", "IN", "DCIS", "Invasive"))
    ) |>
  ggplot(aes(x = sample, y = proportion, fill = cell_type)) +
  geom_col() +
  labs(
    x     = "Sample",
    y     = "Estimated proportion",
    title = "Cell type composition of epithelial samples (CIBERSORTx)"
  ) +
  theme_classic(base_size = 10) +
  theme(axis.text.x = element_blank()) +
  facet_grid(.~ dxs, scales = "free_x", space = "free_x")

print(Ep_Kumar_all_dxs)
ggsave("plots/CIBERSORTx/Ep_Kumar/Ep_Kumar_all_dxs.png", Ep_Kumar_all_dxs, 
       width = 12 , height = 6, dpi = 300)

# NL, BBD, BBD_RS dxs 
Ep_Kumar_NL_BBD_BBDRS <- Ep_Kumar_fractions_long |>
  filter(dxs %in% c("NL", "BBD", "BBD_RS")) |>
  mutate(dxs = factor(dxs, levels = c("NL", "BBD", "BBD_RS"))) |>
  ggplot(aes(x = sample, y = proportion, fill = cell_type)) + 
    geom_col() +
    labs(
      x     = "Sample",
      y     = "Estimated proportion",
      title = "Cell type composition of epithelial samples (CIBERSORTx)"
    ) + 
    theme_classic() + 
    theme(axis.text.x = element_blank()) +
    facet_grid(.~ dxs, scales = "free_x", space = "free_x")
  
print(Ep_Kumar_NL_BBD_BBDRS)
ggsave("plots/CIBERSORTx/Ep_Kumar/Ep_Kumar_NL_BBD_BBDRS.png", Ep_Kumar_NL_BBD_BBDRS, 
       width = 12, height = 6, dpi = 300)

# ADH, IN, DCIS, Invasive dxs
Ep_Kumar_ADH_IN_DCIS_Inv <- Ep_Kumar_fractions_long |>
  filter(dxs %in% c("ADH", "IN", "DCIS", "Invasive")) |>
  mutate(dxs = factor(dxs, levels = c("ADH", "IN", "DCIS", "Invasive"))) |>
  ggplot(aes(x = sample, y = proportion, fill = cell_type)) + 
    geom_col() +
    labs(
      x     = "Sample",
      y     = "Estimated proportion",
      title = "Cell type composition of epithelial samples (CIBERSORTx)"
    ) + 
    theme_classic() + 
    theme(axis.text.x = element_blank()) +
    facet_grid(.~ dxs, scales = "free_x", space = "free_x")
  
print(Ep_Kumar_ADH_IN_DCIS_Inv)
ggsave("plots/CIBERSORTx/Ep_Kumar/Ep_Kumar_ADH_IN_DCIS_Inv.png", Ep_Kumar_ADH_IN_DCIS_Inv, 
       width = 12, height = 6, dpi = 300)
```

### Str (Bassez)

```{r}
# Load CIBERSORTx output (downloaded from the website)
Str_Bassez_results <- read.table(
  "data/CIBERSORTx/CIBERSORTx_Str_Bassez_results.txt",
  header    = TRUE,
  sep       = "\t",
  row.names = 1,
  check.names = FALSE
)

# Keep only the cell fraction columns (remove p-value, correlation, RMSE)
Str_Bassez_fraction_cols <- setdiff(
  colnames(Str_Bassez_results),
  c("P-value", "Correlation", "RMSE")
)
Str_Bassez_fractions <- Str_Bassez_results[, Str_Bassez_fraction_cols]

# Add condition metadata
Str_Bassez_fractions$sample <- rownames(Str_Bassez_fractions)
Str_Bassez_fractions$dxs <- BCRF_Str@meta.data[rownames(Str_Bassez_fractions), "dxs"]

# Reshape for plotting
Str_Bassez_fractions_long <- Str_Bassez_fractions |>
  tidyr::pivot_longer(
    cols      = all_of(Str_Bassez_fraction_cols),
    names_to  = "cell_type",
    values_to = "proportion"
  )

# Stacked bar chart of cell type proportions per sample
# All dxs 
Str_Bassez_all_dxs <- Str_Bassez_fractions_long |>
  filter(dxs %in% c("NL", "BBD", "BBD_RS", "ADH", "IN", "DCIS", "Invasive")) |>
  mutate(dxs = factor(
    dxs,levels = c("NL", "BBD", "BBD_RS", "ADH", "IN", "DCIS", "Invasive"))
    ) |>
  ggplot(aes(x = sample, y = proportion, fill = cell_type)) +
  geom_col() +
  scale_fill_brewer(palette = "Set3", name = "Cell type") +
  labs(
    x     = "Sample",
    y     = "Estimated proportion",
    title = "Cell type composition of stromal samples (CIBERSORTx)"
  ) +
  theme_classic(base_size = 10) +
  theme(axis.text.x = element_blank()) +
  facet_grid(.~ dxs, scales = "free_x", space = "free_x")

print(Str_Bassez_all_dxs)
ggsave("plots/CIBERSORTx/Str_Bassez/Str_Bassez_all_dxs.png", Str_Bassez_all_dxs, 
       width = 8, height = 5, dpi = 300)

# NL, BBD, BBD_RS dxs 
Str_Bassez_NL_BBD_BBDRS <- Str_Bassez_fractions_long |>
  filter(dxs %in% c("NL", "BBD", "BBD_RS")) |>
  mutate(dxs = factor(dxs, levels = c("NL", "BBD", "BBD_RS"))) |>
  ggplot(aes(x = sample, y = proportion, fill = cell_type)) + 
    geom_col() +
    scale_fill_brewer(palette = "Set3", name = "Cell type") + 
    labs(
      x     = "Sample",
      y     = "Estimated proportion",
      title = "Cell type composition of stromal samples (CIBERSORTx)"
    ) + 
    theme_classic() + 
    theme(axis.text.x = element_blank()) +
    facet_grid(.~ dxs, scales = "free_x", space = "free_x")
  
print(Str_Bassez_NL_BBD_BBDRS)
ggsave("plots/CIBERSORTx/Str_Bassez/Str_Bassez_NL_BBD_BBDRS.png", Str_Bassez_NL_BBD_BBDRS, 
       width = 7, height = 5, dpi = 300)

# ADH, IN, DCIS, Invasive dxs
Str_Bassez_ADH_IN_DCIS_Inv <- Str_Bassez_fractions_long |>
  filter(dxs %in% c("ADH", "IN", "DCIS", "Invasive")) |>
  mutate(dxs = factor(dxs, levels = c("ADH", "IN", "DCIS", "Invasive"))) |>
  ggplot(aes(x = sample, y = proportion, fill = cell_type)) + 
    geom_col() +
    scale_fill_brewer(palette = "Set3", name = "Cell type") + 
    labs(
      x     = "Sample",
      y     = "Estimated proportion",
      title = "Cell type composition of stromal samples (CIBERSORTx)"
    ) + 
    theme_classic() + 
    theme(axis.text.x = element_blank()) +
    facet_grid(.~ dxs, scales = "free_x", space = "free_x")
  
print(Str_Bassez_ADH_IN_DCIS_Inv)
ggsave("plots/CIBERSORTx/Str_Bassez/Str_Bassez_ADH_IN_DCIS_Inv.png", Str_Bassez_ADH_IN_DCIS_Inv, 
       width = 7, height = 5, dpi = 300)
```

### Str (Kumar)

```{r}
# Load CIBERSORTx output (downloaded from the website)
Str_Kumar_results <- read.table(
  "data/CIBERSORTx/CIBERSORTx_Str_Kumar_results.txt",
  header    = TRUE,
  sep       = "\t",
  row.names = 1,
  check.names = FALSE
)

# Keep only the cell fraction columns (remove p-value, correlation, RMSE)
Str_Kumar_fraction_cols <- setdiff(
  colnames(Str_Kumar_results),
  c("P-value", "Correlation", "RMSE")
)
Str_Kumar_fractions <- Str_Kumar_results[, Str_Kumar_fraction_cols]

# Add condition metadata
Str_Kumar_fractions$sample <- rownames(Str_Kumar_fractions)
Str_Kumar_fractions$dxs <- BCRF_Str@meta.data[rownames(Str_Kumar_fractions), "dxs"]

# Reshape for plotting
Str_Kumar_fractions_long <- Str_Kumar_fractions |>
  tidyr::pivot_longer(
    cols      = all_of(Str_Kumar_fraction_cols),
    names_to  = "cell_type",
    values_to = "proportion"
  )

# Build color palette with 39 distinct colors for Kumar cell types
my_colors <-as.vector(colorRampPalette(glasbey())(39))

# Stacked bar chart of cell type proportions per sample
# All dxs 
Str_Kumar_all_dxs_glasbey_colors <- Str_Kumar_fractions_long |>
  filter(dxs %in% c("NL", "BBD", "BBD_RS", "ADH", "IN", "DCIS", "Invasive")) |>
  mutate(dxs = factor(
    dxs,levels = c("NL", "BBD", "BBD_RS", "ADH", "IN", "DCIS", "Invasive"))
    ) |>
  ggplot(aes(x = sample, y = proportion, fill = cell_type)) +
  geom_col() +
  labs(
    x     = "Sample",
    y     = "Estimated proportion",
    title = "Cell type composition of stromal samples (CIBERSORTx)"
  ) +
  scale_fill_manual(values = my_colors) + 
  theme_classic(base_size = 10) +
  theme(axis.text.x = element_blank()) +
  facet_grid(.~ dxs, scales = "free_x", space = "free_x")

print(Str_Kumar_all_dxs_glasbey_colors)
ggsave("plots/CIBERSORTx/Str_Kumar/Str_Kumar_all_dxs_glasbey_colors.png", Str_Kumar_all_dxs_glasbey_colors, 
       width = 12 , height = 6, dpi = 300)

Str_Kumar_all_dxs <- Str_Kumar_fractions_long |>
  filter(dxs %in% c("NL", "BBD", "BBD_RS", "ADH", "IN", "DCIS", "Invasive")) |>
  mutate(dxs = factor(
    dxs,levels = c("NL", "BBD", "BBD_RS", "ADH", "IN", "DCIS", "Invasive"))
    ) |>
  ggplot(aes(x = sample, y = proportion, fill = cell_type)) +
  geom_col() +
  labs(
    x     = "Sample",
    y     = "Estimated proportion",
    title = "Cell type composition of stromal samples (CIBERSORTx)"
  ) +
  theme_classic(base_size = 10) +
  theme(axis.text.x = element_blank()) +
  facet_grid(.~ dxs, scales = "free_x", space = "free_x")

print(Str_Kumar_all_dxs)
ggsave("plots/CIBERSORTx/Str_Kumar/Str_Kumar_all_dxs.png", Str_Kumar_all_dxs, 
       width = 12 , height = 6, dpi = 300)

# NL, BBD, BBD_RS dxs 
Str_Kumar_NL_BBD_BBDRS <- Str_Kumar_fractions_long |>
  filter(dxs %in% c("NL", "BBD", "BBD_RS")) |>
  mutate(dxs = factor(dxs, levels = c("NL", "BBD", "BBD_RS"))) |>
  ggplot(aes(x = sample, y = proportion, fill = cell_type)) + 
    geom_col() +
    labs(
      x     = "Sample",
      y     = "Estimated proportion",
      title = "Cell type composition of stromal samples (CIBERSORTx)"
    ) + 
    theme_classic() + 
    theme(axis.text.x = element_blank()) +
    facet_grid(.~ dxs, scales = "free_x", space = "free_x")
  
print(Str_Kumar_NL_BBD_BBDRS)
ggsave("plots/CIBERSORTx/Str_Kumar/Str_Kumar_NL_BBD_BBDRS.png", Str_Kumar_NL_BBD_BBDRS, 
       width = 12, height = 6, dpi = 300)

# ADH, IN, DCIS, Invasive dxs
Str_Kumar_ADH_IN_DCIS_Inv <- Str_Kumar_fractions_long |>
  filter(dxs %in% c("ADH", "IN", "DCIS", "Invasive")) |>
  mutate(dxs = factor(dxs, levels = c("ADH", "IN", "DCIS", "Invasive"))) |>
  ggplot(aes(x = sample, y = proportion, fill = cell_type)) + 
    geom_col() +
    labs(
      x     = "Sample",
      y     = "Estimated proportion",
      title = "Cell type composition of stromal samples (CIBERSORTx)"
    ) + 
    theme_classic() + 
    theme(axis.text.x = element_blank()) +
    facet_grid(.~ dxs, scales = "free_x", space = "free_x")
  
print(Str_Kumar_ADH_IN_DCIS_Inv)
ggsave("plots/CIBERSORTx/Str_Kumar/Str_Kumar_ADH_IN_DCIS_Inv.png", Str_Kumar_ADH_IN_DCIS_Inv, 
       width = 12, height = 6, dpi = 300)
```
