---
title: "07_CIBERSORTx"
format: gfm
---

# 07_CIBERSORTx

The goal of this script is to perform bulk RNA-Seq deconvolution via CIBERSORTx to analyze what fraction of each bulk sample is composed of what cell types. 

## Set-up

```{r}
# Load libraries 
library(Seurat)

# Set global options 
options(future.globals.maxSize = Inf)
print(getOption("future.globals.maxSize"))
options(Seurat.object.assay.version = "v5")

# Load in data
BCRF_Ep <- readRDS("../data/RE-DONE integrate BCRF v1_Ep 1.rds")
BCRF_Str <- readRDS("../data/RE-DONE integrated BCRF v1_Str 3.rds")
```

## Construct signature matrix 

```{r}
# Load in data 
counts <- Read10X("../data/CIBERSORTx/reference/Wu_data")
meta <- read.csv("../data/CIBERSORTx/reference/Whole_miniatlas_meta.csv", row.names = 1)

Wu_seurat <- CreateSeuratObject(counts = counts, meta.data = meta)
saveRDS(Wu_seurat, "../data/CIBERSORTx/reference/Wu_seurat.rds")

Wu_seurat <- readRDS("../data/CIBERSORTx/reference/Wu_seurat.rds")
```

The signature matrix should be constructed as a .tsv file where rows are genes and columns are individual cells, labeled by their cell type name (column header).

```{r}
# Pull raw counts and cell type labels 
# raw counts - NOT normalized, CIBERSORTx does its own internal normalization
counts <- GetAssayData(Wu_seurat, assay = "RNA", layer = "counts")
counts <- as.matrix(counts)

# pull cell type labels - adjust column name to whatever you found in the metadata
celltypes <- Wu_seurat$celltype_major  # or celltype_minor, etc.
```

### Downsample cells per type 

Account for unbalanced cell type group sizes by adjusting to a consistent number per cell type. 

```{r}
set.seed(67)
max_cells_per_type <- 500  # adjust based on your smallest group size

cells_to_keep <- unlist(lapply(unique(celltypes), function(ct) {
  ct_cells <- names(celltypes)[celltypes == ct]
  if (length(ct_cells) > max_cells_per_type) {
    sample(ct_cells, max_cells_per_type)
  } else {
    ct_cells
  }
}))

counts_sub <- counts[, cells_to_keep]
celltypes_sub <- celltypes[cells_to_keep]
```

### Build the CIBERSORTx-format matrix (cell-type labels as column headers)

```{r}
# replace column names with cell type labels
colnames(counts_sub) <- as.character(celltypes_sub)

# add gene symbols as first column
sig_matrix_df <- data.frame(GeneSymbol = rownames(counts_sub), counts_sub, check.names = FALSE)
```

### Clean up gene symbols 

```{r}
# remove duplicate gene symbols (keep first occurrence, or sum if you prefer)
sig_matrix_df <- sig_matrix_df[!duplicated(sig_matrix_df$GeneSymbol), ]

# drop genes with zero expression across all retained cells
gene_sums <- rowSums(sig_matrix_df[, -1])
sig_matrix_df <- sig_matrix_df[gene_sums > 0, ]

# Sanity check
table(colnames(sig_matrix_df)[-1]) # All celltypes only contain 750 
```

### Construct signature matrix file 

```{r}
write.table(
  sig_matrix_df,
  file = "../data/Wu2021_scRef_for_CIBERSORTx.txt",
  sep = "\t",
  quote = FALSE,
  row.names = FALSE
)
```

## Construct input data 

CIBERSORTx expects input counts data CPM or TPM, not raw counts, in the form of .tsv files. 

### Ep

```{r}
# Extract raw counts data from Seurat object 
Ep_raw_counts <- GetAssayData(BCRF_Ep, assay = "RNA", layer = "counts")

# Convert to regular matrix (GetAssayData() sometimes returns sparse)
Ep_raw_counts <- as.matrix(Ep_raw_counts)

# Convert to CPM (counts per million)
# Divide every column sample by total counts for each sample and convert to CPM 
Ep_cpm <- sweep(Ep_raw_counts, 2, colSums(Ep_raw_counts), FUN = "/") * 1e6 

# Construct dataframe w/ gene names as first column 
# CIBERSORTx requires first column name to be "GeneSymbol" 
Ep_cpm_df <- data.frame(
  GeneSymbol = rownames(Ep_cpm), 
  Ep_cpm,
  check.names = FALSE 
)

# Save as .tsv file 
write.table(Ep_cpm_df,
            "../data/CIBERSORTx/Ep_cibersortx_input.tsv",
            sep       = "\t",
            row.names = FALSE,
            quote     = FALSE
            )
```

### Str

```{r}
# Extract raw counts data from Seurat object 
Str_raw_counts <- GetAssayData(BCRF_Str, assay = "RNA", layer = "counts")

# Convert to regular matrix (GetAssayData() sometimes returns sparse)
Str_raw_counts <- as.matrix(Str_raw_counts)

# Remove empty (zero counts) sample 
# Check totals per sample
sample_totals <- colSums(Str_raw_counts)
sample_totals[sample_totals == 0]  # confirm which sample(s) are zero

# Drop it
Str_raw_counts <- Str_raw_counts[, sample_totals > 0]

# Convert to CPM (counts per million)
# Divide every column sample by total counts for each sample and convert to CPM 
Str_cpm <- sweep(Str_raw_counts, 2, colSums(Str_raw_counts), FUN = "/") * 1e6 

# Construct dataframe w/ gene names as first column 
# CIBERSORTx requires first column name to be "GeneSymbol" 
Str_cpm_df <- data.frame(
  GeneSymbol = rownames(Str_cpm), 
  Str_cpm,
  check.names = FALSE 
)

# Save as .tsv file 
write.table(Str_cpm_df,
            "../data/CIBERSORTx/Str_cibersortx_input.tsv",
            sep       = "\t",
            row.names = FALSE,
            quote     = FALSE
            )
```

## Parse output (results) data

```{r}
library(ggplot2)
library(tidyr)
library(dplyr)

# Load CIBERSORTx output (downloaded from the website)
deconv_results <- read.table(
  "CIBERSORTx_Results.txt",
  header    = TRUE,
  sep       = "\t",
  row.names = 1,
  check.names = FALSE
)

# Keep only the cell fraction columns (remove p-value, correlation, RMSE)
fraction_cols <- setdiff(
  colnames(deconv_results),
  c("P-value", "Correlation", "RMSE")
)
fractions <- deconv_results[, fraction_cols]

# Add condition metadata
fractions$sample    <- rownames(fractions)
fractions$condition <- metadata[rownames(fractions), "condition"]

# Reshape for plotting
fractions_long <- fractions |>
  tidyr::pivot_longer(
    cols      = all_of(fraction_cols),
    names_to  = "cell_type",
    values_to = "proportion"
  )

# Stacked bar chart of cell type proportions per sample
ggplot(fractions_long,
       aes(x = sample, y = proportion, fill = cell_type)) +
  geom_col(width = 0.8) +
  scale_fill_brewer(palette = "Set3", name = "Cell type") +
  labs(
    x     = NULL,
    y     = "Estimated proportion",
    title = "Cell type composition by sample (CIBERSORTx)"
  ) +
  theme_classic(base_size = 10) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    legend.text = element_text(size = 7)
  ) +
  facet_grid(. ~ condition, scales = "free_x", space = "free_x")
```

