---
title: "09_DESeq2_InferCNV_Subclusters"
format: gfm
---

# 09_DESeq2_InferCNV_Subclusters

The goal of this script will be to run DESeq2 for several comparisons in the BCRF dataset subclusters defined by inferCNV including:

- BBD_s1: normal-like v. BBD_s2: high-CNVs (1 and 16)
  - outliers BCRF_4-10_15659 and 1583 removed from BBD_s2 

## Set-up to run DESeq2

```{r}
# Load libraries
library(Seurat)
library(DESeq2)
library(BiocParallel)
register(SnowParam(7))
library(tidyverse)
library(EnhancedVolcano)

options(Seurat.object.assay.version = "v5")
```

```{r}
# Import data
BCRF_Ep <- readRDS("data/raw_data/RE-DONE BCRF v1_Ep_combatseq.rds")
```

## Create new metadata column for inferCNV BBD subclusters 

```{r}
# Create new metadata column of "NA" values to fill in
BCRF_Ep$inferCNV_subcluster <- NA 

# Import subcluster sample ID text files into named list 
subclusters_list <- list(
  "BBD_s1" = readLines("data/inferCNV_subcluster_sample_IDs/BBD_s1.txt"),
  "BBD_s2" = readLines("data/inferCNV_subcluster_sample_IDs/BBD_s2.txt"),
  "BBD_s3" = readLines("data/inferCNV_subcluster_sample_IDs/BBD_s3.txt"),
  "BBD_s4" = readLines("data/inferCNV_subcluster_sample_IDs/BBD_s4.txt"),
  "BBD_s5" = readLines("data/inferCNV_subcluster_sample_IDs/BBD_s5.txt"),
  "BBD_s6" = readLines("data/inferCNV_subcluster_sample_IDs/BBD_s6.txt")
)

# Clean: Remove quotes and trailing spaces from sample IDs in text files 
subclusters_list <- lapply(subclusters_list, function(x) {
  trimws(gsub('"', '', x), which = "right")
})

# Loop through and assign the metadata
for (subcluster_ID in names(subclusters_list)) {
  sample_IDs <- subclusters_list[[subcluster_ID]]
  
  # Only assign to samples that actually exist in your Seurat object
  valid_sample_IDs <- intersect(sample_IDs, colnames(BCRF_Ep))
  
  BCRF_Ep@meta.data[valid_sample_IDs, "inferCNV_subcluster"] <- subcluster_ID
}

# Check that sum of BBD samples in subcluster_IDs aligns w/ sum in origial Seurat object 
table(BCRF_Ep$inferCNV_subcluster)
sum(50, 10, 27, 25, 31, 2) # 145 
table(BCRF_Ep$dxs)         # 145 BBD samples 
```

## BBD_s1 v. BBD_s2

```{r}
# Remove outliers 
outliers <- c("BCRF4_10_15659_755_EN_Ep", "BCRF4_10_15837_771_EN_1_Ep")

# Subset BCRF data to BBD_s1 and BBD_s2 inferCNV subclusters, removing outliers
keep <- BCRF_Ep@meta.data$inferCNV_subcluster %in% c("BBD_s1", "BBD_s2") &
        !(colnames(BCRF_Ep) %in% outliers)

BBD_s1_s2_sub_BCRF <- subset(BCRF_Ep, cells = colnames(BCRF_Ep)[keep])

# Get counts data 
BBD_s1_s2_counts <- GetAssayData(BBD_s1_s2_sub_BCRF, assay = "RNA", layer = "counts")

# Build colData from meta.data
BBD_s1_s2_coldata <- data.frame(
  patient = factor(BBD_s1_s2_sub_BCRF@meta.data$patient), 
  inferCNV_subcluster = factor(BBD_s1_s2_sub_BCRF@meta.data$inferCNV_subcluster, levels = c("BBD_s1", "BBD_s2")),
  row.names = colnames(BBD_s1_s2_counts)
)
```

```{r}
# Create DESeqDataSet from subset counts and group information 
BBD_s1_s2_dds <- DESeqDataSetFromMatrix(
  countData = BBD_s1_s2_counts,
  colData = BBD_s1_s2_coldata, 
  design = ~ patient + inferCNV_subcluster 
)
```

### Filtering decisions

```{r}
library(tidyverse)
library(patchwork)

# Before filtering
counts_long <- as.data.frame(as.matrix(BBD_s1_s2_counts)) %>%
  rownames_to_column("gene") %>%
  pivot_longer(
    cols = -gene,
    names_to = "sample",
    values_to = "raw_count"
  ) %>%
  mutate(
    group = BBD_s1_s2_coldata$inferCNV_subcluster[match(sample, rownames(BBD_s1_s2_coldata))]
  )

# Stats per group
group_stats <- counts_long %>%
  group_by(group) %>%
  summarise(
    n_genes = length(unique(gene)),
    pct_zeros = round(100 * mean(raw_count == 0), 0)
  )
group_stats

before_filter_plot <- ggplot(counts_long, aes(x = raw_count, fill = group)) +
  geom_histogram(binwidth = 1, position = "identity", alpha = 0.6) +
  coord_cartesian(xlim = c(0, 100)) +
  labs(
    title = "Distribution of Raw Counts Before Filtering",
    x = "Raw Count", y = "Frequency", fill = "Subcluster",
    subtitle = paste0(format(n_genes, big.mark = ","), " genes\n", pct_zeros, "% zeros")
  ) +
  theme_minimal()
```

```{r}
# Filter genes: keep genes with ≥10 counts in at least 8 samples (smallest group size)
smallest_group_size <- min(table(BBD_s1_s2_coldata$inferCNV_subcluster))  # = 8

keep_genes <- rowSums(counts(BBD_s1_s2_dds) >= 10) >= smallest_group_size
BBD_s1_s2_dds <- BBD_s1_s2_dds[keep_genes, ]

# Confirm how many genes survived
nrow(BBD_s1_s2_dds) # 14448
```

```{r}
# After filtering plot
filtered_counts_long <- as.data.frame(as.matrix(counts(BBD_s1_s2_dds))) %>% # Get post-filter counts
  rownames_to_column("gene") %>% 
  pivot_longer(
    cols = -gene,
    names_to = "sample",
    values_to = "raw_count"
  ) %>%
  mutate(
    group = BBD_s1_s2_coldata$inferCNV_subcluster[match(sample, rownames(BBD_s1_s2_coldata))]
  )

# Stats per group
group_stats_filtered <- filtered_counts_long %>%
  group_by(group) %>%
  summarise(
    n_genes_filtered = length(unique(gene)),
    pct_zeros_filtered = round(100 * mean(raw_count == 0), 0)
  )
group_stats_filtered

after_filter_plot <- ggplot(filtered_counts_long, aes(x = raw_count, fill = group)) +
  geom_histogram(binwidth = 1, position = "identity", alpha = 0.6) +
  coord_cartesian(xlim = c(0, 100)) +
  labs(
    title = "Distribution of Raw Counts After Filtering",
    x = "Raw Count", y = "Frequency", fill = "Subcluster",
    subtitle = paste0(format(n_genes_filtered, big.mark = ","), " genes\n", pct_zeros_filtered, "% zeros")
  ) +
  theme_minimal()

ggsave("plots/DESeq2_InferCNV_Subclusters/BBD_s1_s2_filtering.png", before_filter_plot + after_filter_plot, width = 12, height = 6, dpi = 300)
saveRDS(BBD_s1_s2_dds, "data/DESeq2_InferCNV_Subclusters/BBD_s1_s2_dds_torun.rds")
```

### Run DESeq

```{r}
BBD_s1_s2_dds <- readRDS("data/DESeq2_InferCNV_Subclusters/BBD_s1_s2_dds_torun.rds")

# Run DEseq
BBD_s1_s2_dds <- DESeq(BBD_s1_s2_dds, parallel = TRUE)
```

```{r}
# Save DESeq results 
BBD_s1_s2_dds_res <- results(BBD_s1_s2_dds, contrast = c("inferCNV_subcluster", "BBD_s2", "BBD_s1"))
summary(BBD_s1_s2_dds_res)
saveRDS(BBD_s1_s2_dds_res, "data/DESeq2_InferCNV_Subclusters/BBD_s1_s2_res.rds")

# Convert results to dataframe
BBD_s1_s2_dds_res_df <- BBD_s1_s2_dds_res |>
  as.data.frame() |>
  rownames_to_column(var = "Gene") # Make column for gene ids 

# Save results 
write.csv(BBD_s1_s2_dds_df, "data/DESeq2_InferCNV_Subclusters/BBD_s1_s2_res.csv")
```

### Volcano plot 

```{r}
# Load in results data 
StrNL_StrInv_res <- read.csv("data/DESeq2/StrNL_StrInv_res.csv")
```

```{r}
# Generate volcano plot 
keyGenes <- c("ESR1", "ERBB2", "MYC", "TP53", "PIK3CA", "CCND1")

# Any expression may be dissection artifacts? 

StrNL_StrInv_VP <- EnhancedVolcano(StrNL_StrInv_res,
  x = "log2FoldChange",                 
  y = "padj",                
  lab = StrNL_StrInv_res$Gene, 
  pCutoff = 1e-05, 
  FCcutoff = 1,
  title = "Genes differentially expressed in Str Invasive relative to Str NL",
  selectLab = keyGenes,         # Label only key genes
  labSize = 3.0, 
  labCol = "black",
  labFace = "bold",
  boxedLabels = TRUE,           # boxes make labels stand out from the cloud
  drawConnectors = TRUE,        # connector lines from point to label
  widthConnectors = 0.5,
  colConnectors = 'black',
  arrowheads = FALSE 
) +
  # change the overall theme
  theme_classic() +
  # move the legend to the bottom
  theme(legend.position = "bottom")

# Save plots as .png
ggsave("StrNL_StrInv_VP.png", StrNL_StrInv_VP, path = "plots/DESeq2",
       width = 7, height = 7, dpi = 300)
```