---
title: "09_DESeq2_InferCNV_Subclusters"
format: gfm
---

# 09_DESeq2_InferCNV_Subclusters

The goal of this script will be to run DESeq2 for several comparisons in the BCRF dataset subclusters defined by inferCNV including:

- BBD_s1: normal-like v. BBD_s2: high-CNVs (1 and 16)
  - outliers BCRF_4-10_15659 and 1583 removed from BBD_s2 
  - To analyze CNV differences 
- BBD_s1: normal-like v. BBD_s3 & BBD_s5: low-CNVs, more normal-like  
  - outliers BCRF4_10_14894 and 15687 removed from BBD_s3
  - To analyze general disease progression from normal-like to more disease-state BBD 

## Set-up to run DESeq2

```{r}
# Load libraries
library(Seurat)
library(DESeq2)
library(BiocParallel)
register(SnowParam(7))
library(tidyverse)
library(EnhancedVolcano)

options(Seurat.object.assay.version = "v5")
```

```{r}
# Import data
BCRF_Ep <- readRDS("data/raw_data/BCRF_Ep_combatseq_w_InferCNV.rds")
```

## BBD_s1 v. BBD_s2

```{r}
# Remove outliers 
outliers <- c("BCRF4_10_15659_755_EN_Ep", "BCRF4_10_15837_771_EN_1_Ep")

# Subset BCRF data to BBD_s1 and BBD_s2 inferCNV subclusters, removing outliers
keep <- BCRF_Ep@meta.data$inferCNV_subcluster %in% c("BBD_s1", "BBD_s2") &
        !(colnames(BCRF_Ep) %in% outliers)

BBD_s1_s2_sub_BCRF <- subset(BCRF_Ep, cells = colnames(BCRF_Ep)[keep])

# Get counts data 
BBD_s1_s2_counts <- GetAssayData(BBD_s1_s2_sub_BCRF, assay = "RNA", layer = "counts")

# Build colData from meta.data
BBD_s1_s2_coldata <- data.frame(
  patient = factor(BBD_s1_s2_sub_BCRF@meta.data$patient), 
  inferCNV_subcluster = factor(BBD_s1_s2_sub_BCRF@meta.data$inferCNV_subcluster, levels = c("BBD_s1", "BBD_s2")),
  row.names = colnames(BBD_s1_s2_counts)
)
```

```{r}
# Create DESeqDataSet from subset counts and group information 
BBD_s1_s2_dds <- DESeqDataSetFromMatrix(
  countData = BBD_s1_s2_counts,
  colData = BBD_s1_s2_coldata, 
  design = ~ patient + inferCNV_subcluster 
)
```

### Filtering decisions

```{r}
library(tidyverse)
library(patchwork)

# Before filtering
counts_long <- as.data.frame(as.matrix(BBD_s1_s2_counts)) %>%
  rownames_to_column("gene") %>%
  pivot_longer(
    cols = -gene,
    names_to = "sample",
    values_to = "raw_count"
  ) %>%
  mutate(
    group = BBD_s1_s2_coldata$inferCNV_subcluster[match(sample, rownames(BBD_s1_s2_coldata))]
  )

# Stats per group
group_stats <- counts_long %>%
  group_by(group) %>%
  summarise(
    n_genes = length(unique(gene)),
    pct_zeros = round(100 * mean(raw_count == 0), 0)
  )
group_stats

before_filter_plot <- ggplot(counts_long, aes(x = raw_count, fill = group)) +
  geom_histogram(binwidth = 1, position = "identity", alpha = 0.6) +
  coord_cartesian(xlim = c(0, 100)) +
  labs(
    title = "Distribution of Raw Counts Before Filtering",
    x = "Raw Count", y = "Frequency", fill = "Subcluster",
    subtitle = paste0(format(n_genes, big.mark = ","), " genes\n", pct_zeros, "% zeros")
  ) +
  theme_minimal()
```

```{r}
# Filter genes: keep genes with ≥10 counts in at least 8 samples (smallest group size)
smallest_group_size <- min(table(BBD_s1_s2_coldata$inferCNV_subcluster))  # = 8

keep_genes <- rowSums(counts(BBD_s1_s2_dds) >= 10) >= smallest_group_size
BBD_s1_s2_dds <- BBD_s1_s2_dds[keep_genes, ]

# Confirm how many genes survived
nrow(BBD_s1_s2_dds) # 14448
```

```{r}
# After filtering plot
filtered_counts_long <- as.data.frame(as.matrix(counts(BBD_s1_s2_dds))) %>% # Get post-filter counts
  rownames_to_column("gene") %>% 
  pivot_longer(
    cols = -gene,
    names_to = "sample",
    values_to = "raw_count"
  ) %>%
  mutate(
    group = BBD_s1_s2_coldata$inferCNV_subcluster[match(sample, rownames(BBD_s1_s2_coldata))]
  )

# Stats per group
group_stats_filtered <- filtered_counts_long %>%
  group_by(group) %>%
  summarise(
    n_genes_filtered = length(unique(gene)),
    pct_zeros_filtered = round(100 * mean(raw_count == 0), 0)
  )
group_stats_filtered

after_filter_plot <- ggplot(filtered_counts_long, aes(x = raw_count, fill = group)) +
  geom_histogram(binwidth = 1, position = "identity", alpha = 0.6) +
  coord_cartesian(xlim = c(0, 100)) +
  labs(
    title = "Distribution of Raw Counts After Filtering",
    x = "Raw Count", y = "Frequency", fill = "Subcluster",
    subtitle = paste0(format(n_genes_filtered, big.mark = ","), " genes\n", pct_zeros_filtered, "% zeros")
  ) +
  theme_minimal()

ggsave("plots/DESeq2_InferCNV_Subclusters/BBD_s1_s2_filtering.png", before_filter_plot + after_filter_plot, width = 12, height = 6, dpi = 300)
saveRDS(BBD_s1_s2_dds, "data/DESeq2_InferCNV_Subclusters/BBD_s1_s2_dds_torun.rds")
```

We want to validate whether filtering out so many genes causes the inferCNV subclusters to lose their identities i.e. will the data still cluster into two distinct groups of BBD_s1 and BBD_s2?  

```{r}
# Get the filtered gene list from your DESeq2 dds object
filtered_genes <- rownames(BBD_s1_s2_dds)  # genes that passed your filter
length(filtered_genes)

# Subset the Seurat object to only those genes (keeping all cells/samples)
BBD_s1_s2_sub_BCRF_filtered <- subset(
  BBD_s1_s2_sub_BCRF,
  features = filtered_genes
)

# Standard pipeline, restricted to the filtered gene set as the feature space
BBD_s1_s2_sub_BCRF_filtered <- NormalizeData(BBD_s1_s2_sub_BCRF_filtered)
BBD_s1_s2_sub_BCRF_filtered <- ScaleData(BBD_s1_s2_sub_BCRF_filtered, features = filtered_genes)
BBD_s1_s2_sub_BCRF_filtered <- RunPCA(
  BBD_s1_s2_sub_BCRF_filtered, 
  features = filtered_genes,
  npcs = min(30, ncol(BBD_s1_s2_sub_BCRF_filtered) - 1)  # cap PCs at n_samples - 1 since you only have 58 samples
)
BBD_s1_s2_sub_BCRF_filtered <- RunUMAP(BBD_s1_s2_sub_BCRF_filtered, dims = 1:min(20, ncol(BBD_s1_s2_sub_BCRF_filtered) - 1))

# Plot colored by inferCNV subcluster
BBD_s1_s2_filtered_umap <- DimPlot(
  BBD_s1_s2_sub_BCRF_filtered, 
  group.by = "inferCNV_subcluster", 
  pt.size = 2
) + 
  labs(title = "BBD_s1 vs BBD_s2 (Filtered Genes Only)")

ggsave("plots/DESeq2_InferCNV_Subclusters/BBD_s1_s2_filtered_umap.png", BBD_s1_s2_filtered_umap, width = 6, height = 5, dpi = 300)
```

### Run DESeq

```{r}
BBD_s1_s2_dds <- readRDS("data/DESeq2_InferCNV_Subclusters/BBD_s1_s2_dds_torun.rds")

# Run DEseq
BBD_s1_s2_dds <- DESeq(BBD_s1_s2_dds, parallel = TRUE)

# Check summary
summary(BBD_s1_s2_dds_res)

# Check plot dispersion estimates
png("plots/DESeq2_InferCNV_Subclusters/BBD_s1_s2_dispersion_plot.png", width = 600, height = 500)
plotDispEsts(BBD_s1_s2_dds)
dev.off()
```

```{r}
# Save DESeq results 
BBD_s1_s2_dds_res <- results(BBD_s1_s2_dds, contrast = c("inferCNV_subcluster", "BBD_s2", "BBD_s1"))
saveRDS(BBD_s1_s2_dds_res, "data/DESeq2_InferCNV_Subclusters/BBD_s1_s2_res.rds")

# Convert results to dataframe
BBD_s1_s2_dds_res_df <- BBD_s1_s2_dds_res |>
  as.data.frame() |>
  rownames_to_column(var = "Gene") # Make column for gene ids 

# Save results 
write.csv(BBD_s1_s2_dds_res_df, "data/DESeq2_InferCNV_Subclusters/BBD_s1_s2_res.csv")
```

### Volcano plot 

```{r}
# Load in results data 
BBD_s1_s2_res <- read.csv("data/DESeq2_InferCNV_Subclusters/BBD_s1_s2_res.csv")
```

```{r}
# Generate volcano plot 
keyGenes <- c("ESR1", "ERBB2", "MYC", "TP53", "PIK3CA", "CCND1")

# Any expression may be dissection artifacts? 

BBD_s1_s2_VP <- EnhancedVolcano(BBD_s1_s2_res,
  x = "log2FoldChange",                 
  y = "padj",                
  lab = BBD_s1_s2_res$Gene, 
  pCutoff = 1e-05, 
  FCcutoff = 1,
  title = "Genes differentially expressed in BBD_s2 relative to BBD_s1",
  selectLab = keyGenes,         # Label only key genes
  labSize = 3.0, 
  labCol = "black",
  labFace = "bold",
  boxedLabels = TRUE,           # boxes make labels stand out from the cloud
  drawConnectors = TRUE,        # connector lines from point to label
  widthConnectors = 0.5,
  colConnectors = 'black',
  arrowheads = FALSE 
) +
  # change the overall theme
  theme_classic() +
  # move the legend to the bottom
  theme(legend.position = "bottom")

# Save plots as .png
ggsave("BBD_s1_s2_VP.png", BBD_s1_s2_VP, path = "plots/DESeq2_InferCNV_Subclusters",
       width = 7, height = 7, dpi = 300)
```

## BBD_s1 v. BBD_s3&5

```{r}
# Remove outliers 
outliers <- c("BCRF4_10_14894_674_BBD_Ep_S31_R1_001", "BCRF4_10_15687_757_UDH_Ep")

# Subset BCRF data to BBD_s1 and BBD_s2 inferCNV subclusters, removing outliers
keep <- BCRF_Ep@meta.data$inferCNV_subcluster %in% c("BBD_s1", "BBD_s3", "BBD_s5") &
        !(colnames(BCRF_Ep) %in% outliers)

BBD_s1_s3_s5_sub_BCRF <- subset(BCRF_Ep, cells = colnames(BCRF_Ep)[keep])

# Get counts data 
BBD_s1_s3_s5_counts <- GetAssayData(BBD_s1_s3_s5_sub_BCRF, assay = "RNA", layer = "counts")

# Build colData from meta.data, collapsing BBD_s3 & s5 into one group 
BBD_s1_s3_s5_coldata <- data.frame(
  patient = factor(BBD_s1_s3_s5_sub_BCRF@meta.data$patient),
  inferCNV_subcluster = factor(
    ifelse(BBD_s1_s3_s5_sub_BCRF@meta.data$inferCNV_subcluster %in% c("BBD_s3", "BBD_s5"),
           "BBD_s3_s5",
           as.character(BBD_s1_s3_s5_sub_BCRF@meta.data$inferCNV_subcluster)),
    levels = c("BBD_s1", "BBD_s3_s5")
  ),
  row.names = colnames(BBD_s1_s3_s5_sub_BCRF)
)
```

```{r}
# Create DESeqDataSet from subset counts and group information 
BBD_s1_s3_s5_dds <- DESeqDataSetFromMatrix(
  countData = BBD_s1_s3_s5_counts,
  colData = BBD_s1_s3_s5_coldata, 
  design = ~ patient + inferCNV_subcluster 
)
```

### Filtering decisions

```{r}
library(tidyverse)
library(patchwork)

# Before filtering
counts_long <- as.data.frame(as.matrix(BBD_s1_s3_s5_counts)) %>%
  rownames_to_column("gene") %>%
  pivot_longer(
    cols = -gene,
    names_to = "sample",
    values_to = "raw_count"
  ) %>%
  mutate(
    group = BBD_s1_s3_s5_coldata$inferCNV_subcluster[match(sample, rownames(BBD_s1_s3_s5_coldata))]
  )

# Stats per group
group_stats <- counts_long %>%
  group_by(group) %>%
  summarise(
    n_genes = length(unique(gene)),
    pct_zeros = round(100 * mean(raw_count == 0), 0)
  )
group_stats

before_filter_plot <- ggplot(counts_long, aes(x = raw_count, fill = group)) +
  geom_histogram(binwidth = 1, position = "identity", alpha = 0.6) +
  coord_cartesian(xlim = c(0, 100)) +
  labs(
    title = "Distribution of Raw Counts Before Filtering",
    x = "Raw Count", y = "Frequency", fill = "Subcluster",
    subtitle = paste0(format(group_stats$n_genes, big.mark = ","), " genes\n", group_stats$pct_zeros, "% zeros")
  ) +
  theme_minimal()
```

```{r}
# Filter genes: keep genes with ≥10 counts
keep_genes <- rowSums(counts(BBD_s1_s3_s5_dds)) >= 10
BBD_s1_s3_s5_dds <- BBD_s1_s3_s5_dds[keep_genes, ]

# Confirm how many genes survived
nrow(BBD_s1_s3_s5_dds) # 34138
```

```{r}
# After filtering plot
filtered_counts_long <- as.data.frame(as.matrix(counts(BBD_s1_s3_s5_dds))) %>% # Get post-filter counts
  rownames_to_column("gene") %>% 
  pivot_longer(
    cols = -gene,
    names_to = "sample",
    values_to = "raw_count"
  ) %>%
  mutate(
    group = BBD_s1_s3_s5_coldata$inferCNV_subcluster[match(sample, rownames(BBD_s1_s3_s5_coldata))]
  )

# Stats per group
group_stats_filtered <- filtered_counts_long %>%
  group_by(group) %>%
  summarise(
    n_genes_filtered = length(unique(gene)),
    pct_zeros_filtered = round(100 * mean(raw_count == 0), 0)
  )
group_stats_filtered

after_filter_plot <- ggplot(filtered_counts_long, aes(x = raw_count, fill = group)) +
  geom_histogram(binwidth = 1, position = "identity", alpha = 0.6) +
  coord_cartesian(xlim = c(0, 100)) +
  labs(
    title = "Distribution of Raw Counts After Filtering",
    x = "Raw Count", y = "Frequency", fill = "Subcluster",
    subtitle = paste0(format(group_stats_filtered$n_genes_filtered, big.mark = ","), " genes\n", group_stats_filtered$pct_zeros_filtered, "% zeros")
  ) +
  theme_minimal()

ggsave("plots/DESeq2_InferCNV_Subclusters/BBD_s1_s3_s5_filtering.png", before_filter_plot + after_filter_plot, width = 12, height = 6, dpi = 300)
saveRDS(BBD_s1_s3_s5_dds, "data/DESeq2_InferCNV_Subclusters/BBD_s1_s3_s5_dds_torun.rds")
```

We want to validate whether filtering out so many genes causes the inferCNV subclusters to lose their identities i.e. will the data still cluster into distinct groups of BBD_s1, BBD_s3, and BBD_s5?  

```{r}
# Get the filtered gene list from your DESeq2 dds object
filtered_genes <- rownames(BBD_s1_s3_s5_dds)  # genes that passed your filter
length(filtered_genes)

# Subset the Seurat object to only those genes (keeping all cells/samples)
BBD_s1_s3_s5_sub_BCRF_filtered <- subset(
  BBD_s1_s3_s5_sub_BCRF,
  features = filtered_genes
)

# Standard pipeline, restricted to the filtered gene set as the feature space
BBD_s1_s3_s5_sub_BCRF_filtered <- NormalizeData(BBD_s1_s3_s5_sub_BCRF_filtered)
BBD_s1_s3_s5_sub_BCRF_filtered <- ScaleData(BBD_s1_s3_s5_sub_BCRF_filtered, features = filtered_genes)
BBD_s1_s3_s5_sub_BCRF_filtered <- RunPCA(
  BBD_s1_s3_s5_sub_BCRF_filtered, 
  features = filtered_genes,
  npcs = min(30, ncol(BBD_s1_s3_s5_sub_BCRF_filtered) - 1)
)
BBD_s1_s3_s5_sub_BCRF_filtered <- RunUMAP(BBD_s1_s3_s5_sub_BCRF_filtered, dims = 1:min(20, ncol(BBD_s1_s3_s5_sub_BCRF_filtered) - 1))

# Plot colored by inferCNV subcluster
BBD_s1_s3_s5_filtered_umap <- DimPlot(
  BBD_s1_s3_s5_sub_BCRF_filtered, 
  group.by = "inferCNV_subcluster", 
  pt.size = 2
) + 
  labs(title = "BBD_s1 vs BBD_s3&5 (Filtered Genes Only)")

ggsave("plots/DESeq2_InferCNV_Subclusters/BBD_s1_s3_s5_filtered_umap.png", BBD_s1_s3_s5_filtered_umap, width = 6, height = 5, dpi = 300)
```

The UMAP of the filtered genes distorts the distinct clustering of s1 from s3&s5, so DESeq will be run without any filtering. 

### Run DESeq

```{r}
BBD_s1_s3_s5_dds <- readRDS("data/DESeq2_InferCNV_Subclusters/BBD_s1_s3_s5_dds_torun.rds")

# Run DEseq
BBD_s1_s3_s5_dds <- DESeq(BBD_s1_s3_s5_dds, parallel = TRUE)

# Save DESeq results 
BBD_s1_s3_s5_dds_res <- results(BBD_s1_s3_s5_dds, contrast = c("inferCNV_subcluster", "BBD_s3_s5", "BBD_s1"))
saveRDS(BBD_s1_s3_s5_dds_res, "data/DESeq2_InferCNV_Subclusters/BBD_s1_s3_s5_res.rds")

# Convert results to dataframe
BBD_s1_s3_s5_dds_res_df <- BBD_s1_s3_s5_dds_res |>
  as.data.frame() |>
  rownames_to_column(var = "Gene") # Make column for gene ids 

# Save results 
write.csv(BBD_s1_s3_s5_dds_res_df, "data/DESeq2_InferCNV_Subclusters/BBD_s1_s3_s5_res.csv")

# Check summary
summary(BBD_s1_s3_s5_dds_res)
```
