---
title: "08_CIBERSORTx_Analysis"
format: gfm
---

# 08_CIBERSORTx_Analysis

The goal of this script will be to conduct statistical tests on CIBERSORTx results to compare whether each cell type's fraction differs across diagnostic groups, focusing on differences between "NL", "BBD", and "DCIS" epithelial samples.

Comparisons: 
- BBD_s1 (normal-like) v. BBD_s2 (high CNVs)
- BBD_s1 (normal-like) v. BBD_s3_s5 (low CNVs, more normal-like) 

The general workflow will be to perform omnibus testing with Kruskal-Wallis screen (if 2+ groups) -\> FDR -\> pairwise Wilcoxon -\> FDR.

# Set-up

```{r}
# Load libraries 
library(Seurat)  
library(tidyr)
library(dplyr)
library(purrr)
library(rstatix)
library(ggplot2)

# Set global options 
options(future.globals.maxSize = Inf)
print(getOption("future.globals.maxSize"))
options(Seurat.object.assay.version = "v5")
```

# Read in metadata & CIBERSORTx results

```{r}
# Load-in metadata
BCRF_Ep <- readRDS("data/raw_data/BCRF_Ep_combatseq_w_InferCNV.rds")

# Define outliers 
outliers <- c("BCRF4_10_15659_755_EN_Ep", "BCRF4_10_15837_771_EN_1_Ep")

# Filter relevant metadata 
meta <- BCRF_Ep@meta.data |>
  rownames_to_column("sample_id") |>
  filter(inferCNV_subcluster %in% c("BBD_s1", "BBD_s2")) |>
  filter(!sample_id %in% outliers)

# Load-in raw CIBERSORTx results table
ciber_raw <- read.delim(
  "data/CIBERSORTx/CIBERSORTx_Ep_Kumar_results.txt", 
  sep = "\t",
  header = TRUE,
  check.names = FALSE          # keep cell type names exactly as CIBERSORTx wrote them
)

# Drop non-cell-type summary columns
cols_to_drop <- c("P-value", "Correlation", "RMSE")

# Identify which columns are actual cell type fraction columns:
# everything except the sample ID column ("Mixture") and the QC columns above.
cell_type_cols <- setdiff(colnames(ciber_raw), c("Mixture", cols_to_drop))
```

# Reshape to long format and attach group labels

```{r}
df_long <- ciber_raw |>
  select(Mixture, all_of(cell_type_cols)) |>       # keep sample ID + fraction columns only
  pivot_longer(
    cols = all_of(cell_type_cols),                 # columns to stack into long format
    names_to = "cell_type",                        # new column holding cell type names
    values_to = "fraction"                         # new column holding the fraction values
  ) |>
  left_join(meta, by = c("Mixture" = "sample_id")) # attach diagnosis_group per sample

# How many total samples are in df_long vs how many actually matched?
n_distinct(df_long$Mixture)                     # total unique samples in CIBERSORTx output: 552
n_distinct(df_long$Mixture[!is.na(df_long$inferCNV_subcluster)])  # how many matched to BBD_s1/s2: 58
nrow(meta)  # 58 = 58 

# Make inferCNV_subcluster a factor with an explicit level order for plot facet labeling and how pairwise comparisons are labeled
df_long <- df_long %>%
  mutate(inferCNV_subcluster = factor(inferCNV_subcluster, levels = c("BBD_s1", "BBD_s2")))

```

# Kruskal-Wallis screen across cell types

Kruskal-Wallis is a omnibus/global test that will answer: for EACH cell type, does fraction differ across ANY of the groups? This is not the same as defining which specific pairs of groups differ.

```{r}
# Get the list of unique cell types to loop over.
cell_types <- unique(df_long$cell_type)

# map_dfr runs the function over each cell type and row-binds the results
# into a single data frame (dfr = "data frame, row-bind").
kw_results <- map_dfr(cell_types, function(ct) {

  # Subset to just this cell type's rows across all samples.
  sub_df <- df_long %>% filter(cell_type == ct)

  # Run Kruskal-Wallis: fraction ~ diagnosis_group.
  # This tests whether the rank distributions differ across the 3 dxs groups.
  kw_test <- kruskal.test(fraction ~ dxs, data = sub_df)

  # Return a one-row summary for this cell type.
  data.frame(
    cell_type = ct,
    kw_statistic = kw_test$statistic,   # the H statistic
    kw_pvalue = kw_test$p.value,        # raw (unadjusted) p-value
    n_samples = nrow(sub_df)            # sanity-check sample count
  )
})
```

## FDR (Benjamini-Hochberg) correction across all cell types tested

```{r}
# Pull all cell types tested in this KW screening step
kw_results <- kw_results %>%
  mutate(kw_padj = p.adjust(kw_pvalue, method = "BH")) %>%
  arrange(kw_padj)  

print(kw_results)

# Define which cell types move on to pairwise follow-up testing (p<0.05)
sig_cell_types <- kw_results %>%
  filter(kw_padj < 0.05) %>%
  pull(cell_type)

message(paste("Cell types passing KW FDR screen:", paste(sig_cell_types, collapse = ", ")))
```

# Pairwise Wilcoxon rank-sum, only for KW-significant types

For cell types where KW found SOME difference, Wilcoxon figures out WHICH specific group pairs differ i.e. NL vs Invasive but not NL vs DCIS.

- rstatix::wilcox_effsize() provides the rank-biserial correlation, which informs the size of the difference between two dxs expression of the same cell type
  - ranges from -1 to 1
  - \|r\| \~0.1 small, \~0.3 medium, \~0.5 large differences (Cohen-style rough guide)
  - r \> 0: group1 has larger fractions than group2
  - r \< 0: group2 has larger fractions than group1
- pairwise.wilcox.test() provides significance of these differences
  - FDR automatically applied

```{r}
# Computes effect sizes for a Wilcoxon test:
wilcox_effect_size <- function(x, y) {
  
  # Create tidy frame (req by rstatix) of fraction values 
  df <- data.frame(
    fraction = c(x, y),
    group = factor(c(rep("group1", length(x)),
                     rep("group2", length(y))))
  )
  
  rstatix::wilcox_effsize(
    df,
    value ~ group
  )$effsize
}

# Loop over only the significant cell types from Round 1.
pairwise_results <- map_dfr(sig_cell_types, function(ct) {

  sub_df <- df_long %>% filter(cell_type == ct)

  # pairwise.wilcox.test runs Wilcoxon rank-sum for every pair of groups
  # and automatically applies FDR correction WITHIN this cell type's
  # family of pairwise comparisons (i.e., not pooled with other cell types).
  pw_test <- pairwise.wilcox.test(
    x = sub_df$fraction,
    g = sub_df$dxs,
    p.adjust.method = "BH",   # FDR correction within this cell type's pairwise family
    exact = FALSE              # avoid warnings from ties in the fraction data
  )

  # reshape results matrix into tidy long data frame 
  pmat <- pw_test$p.value
  pw_long <- as.data.frame(as.table(pmat)) %>%
    rename(group1 = Var1, group2 = Var2, p_adj = Freq) %>%
    filter(!is.na(p_adj))   # matrix has NAs in the unused upper triangle

  # Add cell type label and effect size for each pair.
  pw_long <- pw_long %>%
    mutate(cell_type = ct) %>%
    rowwise() %>%
    mutate(
      effect_size_r = wilcox_effect_size(
        sub_df$fraction[sub_df$dxs == group1],
        sub_df$fraction[sub_df$dxs == group2]
      )
    ) %>%
    ungroup()

  pw_long
})

# Reorder columns for readability.
pairwise_results <- pairwise_results %>%
  select(cell_type, group1, group2, p_adj, effect_size_r)

print(pairwise_results)
```

# Save results for downstream

```{r}
write.csv(kw_results, "data/CIBERSORTx_Analysis/kw_screen_results.csv", row.names = FALSE)
write.csv(pairwise_results, "data/CIBERSORTx_Analysis/pairwise_wilcoxon_results.csv", row.names = FALSE)
```
