---
title: "10_GSEA_InferCNV_Subclusters"
format: gfm
---

# 10_GSEA_InferCNV_Subclusters

The goal of this script will be to run gene set enrichment analysis on DESeq2 results for several comparisons in the BCRF dataset subclusters defined by inferCNV including:

- BBD_s1: normal-like v. BBD_s2: high-CNVs (1 and 16)
  - outliers BCRF_4-10_15659 and 1583 removed from BBD_s2
- BBD_s1: normal-like v. BBD_s3 & BBD_s5: low-CNVs, more normal-like  
  - outliers BCRF4_10_14894 and 15687 removed from BBD_s3

# Load libraries 

```{r}
# Load libraries
library(readr)
library(clusterProfiler)
library(msigdbr)
library(org.Hs.eg.db)
library(enrichplot)
library(ggplot2)
library(tidyverse)
library(ggplotify)
```

# BBD_s1 v. BBD_s2

## Set up
```{r}
# Load in data 
BBD_s1_s2 <- read.csv("data/DESeq2_InferCNV_Subclusters/BBD_s1_s2_res.csv", row.names = 1)
```

```{r}
# Prepare ranked list of genes
# Use signed p value as ranking: sign(df$log2fc)*(-log10(df$pval))) to account for genes w/ large FC but are non-significant
BBD_s1_s2_rankings <- sign(BBD_s1_s2$log2FoldChange)*(-log10(BBD_s1_s2$pvalue))
names(BBD_s1_s2_rankings) <- BBD_s1_s2$Gene # genes as names

# Sort by decreasing signed p values 
BBD_s1_s2_rankings <- sort(BBD_s1_s2_rankings, decreasing = TRUE)

# Convert from SYMBOL to Entrez nomenclature 
BBD_s1_s2_genes_mapped <- bitr(
  names(BBD_s1_s2_rankings),
  fromType = "SYMBOL", 
  toType = c("ENTREZID"),
  OrgDb = org.Hs.eg.db)

nrow(BBD_s1_s2_genes_mapped) # Check how many genes were mapped successfully
length(BBD_s1_s2_rankings)   # 13145 v. 14448
```

```{r}
# Create named vector w/ Entrez IDs as names and corresponding signed p values 
BBD_s1_s2_rankings_mapped <- BBD_s1_s2_rankings[BBD_s1_s2_genes_mapped$SYMBOL]
names(BBD_s1_s2_rankings_mapped) <- BBD_s1_s2_genes_mapped$ENTREZID

# Check all genes mapped correctly
length(BBD_s1_s2_rankings_mapped) == nrow(BBD_s1_s2_genes_mapped)
head(BBD_s1_s2_rankings_mapped)
```

## Running gseGO (BP) 

```{r}
set.seed(67) 

# Run gseGO
BBD_s1_s2_gsea_res_BP <- clusterProfiler::gseGO(
  geneList = BBD_s1_s2_rankings_mapped, 
  OrgDb = org.Hs.eg.db, 
  ont = "BP", 
  minGSSize = 5, 
  maxGSSize = 500, 
  pvalueCutoff = 0.05,
  pAdjustMethod = "BH",
  nPermSimple = 50000, # Increased from 10000
  scoreType = "std",
  seed = TRUE 
)

# Convert Entrez IDs to gene symbols for readability
BBD_s1_s2_gsea_BP_readable <- clusterProfiler::setReadable(
  BBD_s1_s2_gsea_res_BP,
  OrgDb = org.Hs.eg.db,
  keyType = "ENTREZID"
  )

saveRDS(BBD_s1_s2_gsea_BP_readable, "data/GSEA_InferCNV_Subclusters/BBD_s1_s2_gsea_res_BP.rds")

# Convert gseaResult object into dataframe
BBD_s1_s2_gsea_res_BP_df <- as.data.frame(BBD_s1_s2_gsea_BP_readable)
write_csv(BBD_s1_s2_gsea_res_BP_df, "data/GSEA_InferCNV_Subclusters/BBD_s1_s2_gsea_res_BP.csv")
```

## Visualization (BP)

```{r}
# Load results 
BBD_s1_s2_gsea_res_BP <- readRDS("data/GSEA_InferCNV_Subclusters/BBD_s1_s2_gsea_res_BP.rds")
BBD_s1_s2_gsea_res_BP_df <- na.omit(read.csv("data/GSEA_InferCNV_Subclusters/BBD_s1_s2_gsea_res_BP.csv"))
```

```{r}
# Significant # of pathways 
nrow(BBD_s1_s2_gsea_res_BP_df |> filter(p.adjust < 0.05)) # 0
```

## Running GSEA (Hallmark)

We'll use the Hallmark gene set from Molecular Signatures Database (MSigDB) for our PEA.

```{r}
# Hallmark gene set from msigdbr_collections()
hallmark_sets <- msigdbr(species = 'Homo sapiens', collection = 'H')

# Convert to named list format required by clusterprofiler: the first column should have pathway names
# the second column should have genes in Entrez ID format 
hallmark_sets <- hallmark_sets |>
  dplyr::select(gs_name, ncbi_gene)
```

```{r}
set.seed(67)

# Run GSEA
BBD_s1_s2_gsea_res_HM <- clusterProfiler::GSEA(
  geneList = BBD_s1_s2_rankings_mapped, 
  TERM2GENE = hallmark_sets, 
  minGSSize = 15, 
  maxGSSize = 500, 
  pvalueCutoff = 0.05,
  pAdjustMethod = "BH",
  nPermSimple = 50000, # Increase this value as need
  scoreType = "std",
  seed = TRUE
)

# Convert Entrez IDs to gene symbols for readability
BBD_s1_s2_gsea_HM_readable <- clusterProfiler::setReadable(
  BBD_s1_s2_gsea_res_HM,
  OrgDb = org.Hs.eg.db,
  keyType = "ENTREZID"
  )

# Clean up pathway titles for readability 
BBD_s1_s2_gsea_HM_readable@result <- BBD_s1_s2_gsea_HM_readable@result |>
  mutate(
    Description = Description |>
      str_replace_all("HALLMARK_", " ") |> # Delete HALLMARK_
      str_replace_all("_", " ") |>         # Delete underscores in pathway titles
      str_to_title(),                      # Make pathways title case instead of caps
      .before = 1                          # Put new Description column at front
  )

saveRDS(BBD_s1_s2_gsea_HM_readable, "data/GSEA_InferCNV_Subclusters/BBD_s1_s2_gsea_res_HM.rds")

# Convert gseaResult object into dataframe
BBD_s1_s2_gsea_res_HM_df <- as.data.frame(BBD_s1_s2_gsea_HM_readable)
write_csv(BBD_s1_s2_gsea_res_HM_df, "data/GSEA_InferCNV_Subclusters/BBD_s1_s2_gsea_res_HM.csv")
```

## Visualization (HM) 

```{r}
# Load results 
BBD_s1_s2_gsea_res_HM <- readRDS("data/GSEA_InferCNV_Subclusters/BBD_s1_s2_gsea_res_HM.rds")
BBD_s1_s2_gsea_res_HM_df <- read.csv("data/GSEA_InferCNV_Subclusters/BBD_s1_s2_gsea_res_HM.csv")
```

```{r}
# Significant # of pathways 
nrow(BBD_s1_s2_gsea_res_HM_df |> filter(p.adjust < 0.1)) # 8
```

```{r}
# Generate barplot
BBD_s1_s2_barplot_HM <- BBD_s1_s2_gsea_res_HM_df |>
  filter(p.adjust < 0.1) |>
  arrange(desc(NES)) |>
  ggplot(aes(x = NES, y = reorder(Description, NES), fill = p.adjust)) + 
  geom_col() + 
  scale_fill_gradient(low = "red", high = "blue") +
  labs(x = "Normalized Enrichment Score (NES)", y = "Pathway", title = "Hallmark Pathways Enriched\nin BBD_s2 Over BBD_s1") +
  theme_bw() 
  
ggsave("BBD_s1_s2_barplot_HM.png", BBD_s1_s2_barplot_HM, path = "plots/GSEA_InferCNV_Subclusters",
       width = 9, height = 6, dpi = 300)
```

```{r}
# Cnetplot
BBD_s1_s2_cnetplot_HM <- enrichplot::cnetplot(
  BBD_s1_s2_gsea_res_HM,
  categorySizeBy = ~NES, # or "pvalue", "p.adjust"
  showCategory = 5
  ) 

ggsave("BBD_s1_s2_cnetplot_HM.png", BBD_s1_s2_cnetplot_HM, path = "plots/GSEA_InferCNV_Subclusters",
       width = 12, height = 8, dpi = 300)
```

```{r}
# Dotplot
BBD_s1_s2_dotplot_HM <- dotplot(BBD_s1_s2_gsea_res_HM, showCategory = 10) +
  ggtitle("Top 10 Hallmark Pathways from GSEA\nin BBD_s2 Over BBD_s1") +
  theme(axis.text.y = element_text(size = 9, family = "Arial"))

ggsave("BBD_s1_s2_dotplot_HM.png", BBD_s1_s2_dotplot_HM, path = "plots/GSEA_InferCNV_Subclusters",
       width = 6, height = 7, dpi = 300)
```

# BBD_s1 v. BBD_s3_s5

## Set up
```{r}
# Load in data 
BBD_s1_s3_s5 <- read.csv("data/DESeq2_InferCNV_Subclusters/BBD_s1_s3_s5_res.csv", row.names = 1)

# Check for NAs before ranking (DESeq2 often flags low-count genes with NA pvalue)
sum(is.na(BBD_s1_s3_s5$pvalue))          # 7176
sum(is.na(BBD_s1_s3_s5$log2FoldChange))  # 7176

# Drop rows with NA pvalue or log2FoldChange - can't rank these
BBD_s1_s3_s5 <- BBD_s1_s3_s5[!is.na(BBD_s1_s3_s5$pvalue) & !is.na(BBD_s1_s3_s5$log2FoldChange), ]
```

```{r}
# Prepare ranked list of genes
# Use signed p value as ranking: sign(df$log2fc)*(-log10(df$pval))) to account for genes w/ large FC but are non-significant
BBD_s1_s3_s5_rankings <- sign(BBD_s1_s3_s5$log2FoldChange)*(-log10(BBD_s1_s3_s5$pvalue))
names(BBD_s1_s3_s5_rankings) <- BBD_s1_s3_s5$Gene # genes as names

# Sort by decreasing signed p values 
BBD_s1_s3_s5_rankings <- sort(BBD_s1_s3_s5_rankings, decreasing = TRUE)

# Convert from SYMBOL to Entrez nomenclature 
BBD_s1_s3_s5_genes_mapped <- bitr(
  names(BBD_s1_s3_s5_rankings),
  fromType = "SYMBOL", 
  toType = c("ENTREZID"),
  OrgDb = org.Hs.eg.db)

nrow(BBD_s1_s3_s5_genes_mapped) # Check how many genes were mapped successfully
length(BBD_s1_s3_s5_rankings)   # 43668 v. 28479
```

Converting produces warning: 'select()' returned 1:many mapping between keys and columns because some single gene symbols map to multiple EntrezIDs (or vice versa). 

```{r}
# Dedupe: some SYMBOLs map to multiple ENTREZIDs (1:many warning) - keep first hit
BBD_s1_s3_s5_genes_mapped <- BBD_s1_s3_s5_genes_mapped[!duplicated(BBD_s1_s3_s5_genes_mapped$SYMBOL), ]

# Merge Entrez IDs back onto ranked vector
BBD_s1_s3_s5_rankings_df <- data.frame(
  SYMBOL = names(BBD_s1_s3_s5_rankings),
  rank = BBD_s1_s3_s5_rankings
)
BBD_s1_s3_s5_rankings_df <- merge(BBD_s1_s3_s5_rankings_df, BBD_s1_s3_s5_genes_mapped, by = "SYMBOL")

# Dedupe on ENTREZID too (in case multiple SYMBOLs map to same ID) - keep the one with larger |rank|
BBD_s1_s3_s5_rankings_df <- BBD_s1_s3_s5_rankings_df[order(-abs(BBD_s1_s3_s5_rankings_df$rank)), ]
BBD_s1_s3_s5_rankings_df <- BBD_s1_s3_s5_rankings_df[!duplicated(BBD_s1_s3_s5_rankings_df$ENTREZID), ]

# Rebuild named, sorted vector using ENTREZID
BBD_s1_s3_s5_rankings_entrez <- BBD_s1_s3_s5_rankings_df$rank
names(BBD_s1_s3_s5_rankings_entrez) <- BBD_s1_s3_s5_rankings_df$ENTREZID
BBD_s1_s3_s5_rankings_entrez <- sort(BBD_s1_s3_s5_rankings_entrez, decreasing = TRUE)

# Sanity check: no NAs, no duplicate names, strictly decreasing
sum(is.na(BBD_s1_s3_s5_rankings_entrez))             # 0
sum(duplicated(names(BBD_s1_s3_s5_rankings_entrez))) # 0
```

## Running gseGO (BP) 

```{r}
set.seed(67) 

# Run gseGO
BBD_s1_s3_s5_gsea_res_BP <- clusterProfiler::gseGO(
  geneList = BBD_s1_s3_s5_rankings_entrez, 
  OrgDb = org.Hs.eg.db, 
  ont = "BP", 
  minGSSize = 5, 
  maxGSSize = 500, 
  pvalueCutoff = 0.05,
  pAdjustMethod = "BH",
  nPermSimple = 50000, # Increased from 10000
  scoreType = "std",
  seed = TRUE 
)

# Convert Entrez IDs to gene symbols for readability
BBD_s1_s3_s5_gsea_BP_readable <- clusterProfiler::setReadable(
  BBD_s1_s3_s5_gsea_res_BP,
  OrgDb = org.Hs.eg.db,
  keyType = "ENTREZID"
  )

saveRDS(BBD_s1_s3_s5_gsea_BP_readable, "data/GSEA_InferCNV_Subclusters/BBD_s1_s3_s5_gsea_res_BP.rds")

# Convert gseaResult object into dataframe
BBD_s1_s3_s5_gsea_res_BP_df <- as.data.frame(BBD_s1_s3_s5_gsea_BP_readable)
write_csv(BBD_s1_s3_s5_gsea_res_BP_df, "data/GSEA_InferCNV_Subclusters/BBD_s1_s3_s5_gsea_res_BP.csv")
```

## Visualization (BP)

```{r}
# Load results 
BBD_s1_s3_s5_gsea_res_BP <- readRDS("data/GSEA_InferCNV_Subclusters/BBD_s1_s3_s5_gsea_res_BP.rds")
BBD_s1_s3_s5_gsea_res_BP_df <- na.omit(read.csv("data/GSEA_InferCNV_Subclusters/BBD_s1_s3_s5_gsea_res_BP.csv"))
```

```{r}
# Significant # of pathways 
nrow(BBD_s1_s3_s5_gsea_res_BP_df |> filter(p.adjust < 0.05)) # 320
```

```{r}
# Generate barplot
# Select the top 10 upregulated pathways according to NES 
BBD_s1_s3_s5_top10_up_BP <- BBD_s1_s3_s5_gsea_res_BP_df |>
  filter(p.adjust < 0.05) |>
  arrange(desc(NES)) |>
  filter(NES > 0) |>
  slice_head(n = 10)

# Select the top 10 downregulated pathways according to NES 
BBD_s1_s3_s5_top10_down_BP <- BBD_s1_s3_s5_gsea_res_BP_df |>
  filter(p.adjust < 0.05) |>
  arrange(NES) |>
  filter(NES < 0) |>
  slice_head(n = 10)

# Combine dfs 
BBD_s1_s3_s5_top10_updown_BP <- bind_rows(BBD_s1_s3_s5_top10_up_BP, BBD_s1_s3_s5_top10_down_BP)

BBD_s1_s3_s5_barplot_BP <- BBD_s1_s3_s5_top10_updown_BP |>
  ggplot(aes(x = NES, y = reorder(Description, NES), fill = p.adjust)) + 
  geom_col() + 
  scale_fill_gradient(low = "red", high = "blue") +
  labs(x = "Normalized Enrichment Score (NES)", y = "Pathway", title = "Top-Ranked GO:BP Terms by NES\nin BBD_s3_s5 Over BBD_s1") +
  theme_bw() 
  
ggsave("BBD_s1_s3_s5_barplot_BP.png", BBD_s1_s3_s5_barplot_BP, path = "plots/GSEA_InferCNV_Subclusters",
       width = 9, height = 6, dpi = 300)
```

```{r}
# Cnetplot
BBD_s1_s3_s5_cnetplot_BP <- enrichplot::cnetplot(
  BBD_s1_s3_s5_gsea_res_BP,
  categorySizeBy = ~NES, # or "pvalue", "p.adjust"
  showCategory = 5
  ) 

ggsave("BBD_s1_s3_s5_cnetplot_BP.png", BBD_s1_s3_s5_cnetplot_BP, path = "plots/GSEA_InferCNV_Subclusters",
       width = 12, height = 8, dpi = 300)
```

```{r}
# Dotplot
BBD_s1_s3_s5_dotplot_BP <- dotplot(BBD_s1_s3_s5_gsea_res_BP, showCategory = 15) +
  ggtitle("Top-Ranked GO:BP Terms from GSEA\nin BBD_s3_s5 Over BBD_s1") +
  theme(axis.text.y = element_text(size = 9, family = "Arial"))

ggsave("BBD_s1_s3_s5_dotplot_BP.png", BBD_s1_s3_s5_dotplot_BP, path = "plots/GSEA_InferCNV_Subclusters",
       width = 6, height = 8, dpi = 300)
```

## Running GSEA (Hallmark)

We'll use the Hallmark gene set from Molecular Signatures Database (MSigDB) for our PEA.

```{r}
# Hallmark gene set from msigdbr_collections()
hallmark_sets <- msigdbr(species = 'Homo sapiens', collection = 'H')

# Convert to named list format required by clusterprofiler: the first column should have pathway names
# the second column should have genes in Entrez ID format 
hallmark_sets <- hallmark_sets |>
  dplyr::select(gs_name, ncbi_gene)
```

```{r}
set.seed(67)

# Run GSEA
BBD_s1_s3_s5_gsea_res_HM <- clusterProfiler::GSEA(
  geneList = BBD_s1_s3_s5_rankings_entrez, 
  TERM2GENE = hallmark_sets, 
  minGSSize = 15, 
  maxGSSize = 500, 
  pvalueCutoff = 0.05,
  pAdjustMethod = "BH",
  nPermSimple = 50000, # Increase this value as need
  scoreType = "std",
  seed = TRUE
)

# Convert Entrez IDs to gene symbols for readability
BBD_s1_s3_s5_gsea_HM_readable <- clusterProfiler::setReadable(
  BBD_s1_s3_s5_gsea_res_HM,
  OrgDb = org.Hs.eg.db,
  keyType = "ENTREZID"
  )

# Clean up pathway titles for readability 
BBD_s1_s3_s5_gsea_HM_readable@result <- BBD_s1_s3_s5_gsea_HM_readable@result |>
  mutate(
    Description = Description |>
      str_replace_all("HALLMARK_", " ") |> # Delete HALLMARK_
      str_replace_all("_", " ") |>         # Delete underscores in pathway titles
      str_to_title(),                      # Make pathways title case instead of caps
      .before = 1                          # Put new Description column at front
  )

saveRDS(BBD_s1_s3_s5_gsea_HM_readable, "data/GSEA_InferCNV_Subclusters/BBD_s1_s3_s5_gsea_res_HM.rds")

# Convert gseaResult object into dataframe
BBD_s1_s3_s5_gsea_res_HM_df <- as.data.frame(BBD_s1_s3_s5_gsea_HM_readable)
write_csv(BBD_s1_s3_s5_gsea_res_HM_df, "data/GSEA_InferCNV_Subclusters/BBD_s1_s3_s5_gsea_res_HM.csv")
```

## Visualization (HM) 

```{r}
# Load results 
BBD_s1_s3_s5_gsea_res_HM <- readRDS("data/GSEA_InferCNV_Subclusters/BBD_s1_s3_s5_gsea_res_HM.rds")
BBD_s1_s3_s5_gsea_res_HM_df <- read.csv("data/GSEA_InferCNV_Subclusters/BBD_s1_s3_s5_gsea_res_HM.csv")
```

```{r}
# Significant # of pathways 
nrow(BBD_s1_s3_s5_gsea_res_HM_df |> filter(p.adjust < 0.1)) #21
```

```{r}
# Generate barplot
BBD_s1_s3_s5_barplot_HM <- BBD_s1_s3_s5_gsea_res_HM_df |>
  filter(p.adjust < 0.1) |>
  arrange(desc(NES)) |>
  ggplot(aes(x = NES, y = reorder(Description, NES), fill = p.adjust)) + 
  geom_col() + 
  scale_fill_gradient(low = "red", high = "blue") +
  labs(x = "Normalized Enrichment Score (NES)", y = "Pathway", title = "Hallmark Pathways Enriched\nin BBD_s3_s5 Over BBD_s1") +
  theme_bw() 
  
ggsave("BBD_s1_s3_s5_barplot_HM.png", BBD_s1_s3_s5_barplot_HM, path = "plots/GSEA_InferCNV_Subclusters",
       width = 9, height = 6, dpi = 300)
```

```{r}
# Cnetplot
BBD_s1_s3_s5_cnetplot_HM <- enrichplot::cnetplot(
  BBD_s1_s3_s5_gsea_res_HM,
  categorySizeBy = ~NES, # or "pvalue", "p.adjust"
  showCategory = 5
  ) 

ggsave("BBD_s1_s3_s5_cnetplot_HM.png", BBD_s1_s3_s5_cnetplot_HM, path = "plots/GSEA_InferCNV_Subclusters",
       width = 12, height = 8, dpi = 300)
```

```{r}
# Dotplot
BBD_s1_s3_s5_dotplot_HM <- dotplot(BBD_s1_s3_s5_gsea_res_HM, showCategory = 10) +
  ggtitle("Top 10 Hallmark Pathways from GSEA\nin BBD_s3_s5 Over BBD_s1") +
  theme(axis.text.y = element_text(size = 9, family = "Arial"))

ggsave("BBD_s1_s3_s5_dotplot_HM.png", BBD_s1_s3_s5_dotplot_HM, path = "plots/GSEA_InferCNV_Subclusters",
       width = 6, height = 7, dpi = 300)
```

